# Text Mining

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(tidytext)
library(textstem)     # lemmatize words
library(topicmodels)  # LDA
library(stm)          # structure topic modeling
```



Text mining is the process of distilling actionable insights from text, typically by identifying patterns with statistical pattern learning. Commong text mining tasks include topic modeling, sentiment analysis, classification modeling.

Be aware that there are two main approaches to text mining.  *Semantic parsing* identifies words by type and order (sentences, phrases, nouns/verbs, proper nouns, etc.).  *Bag of words* simply treats words as attributes of the document. Bag of words is the simpler way to go, the only one I've encountered, and the one used throughout these notes.

There are three packages with text mining frameworks: **tidytext**, **stm**, and **tm**. These notes mostly use **tidytext**, but I made an appendix with **tm** basics because **tidytext** uses its class structure (see `cast_dtm()`).

As much as possible, I will build examples using the text from <u>The Adventures of Tom Sawyer</u>, available on https://gutenberg.org and accessible using the **gutenbergr** package. I added the chapter identifier and dropped the book id column. Each row is a line of text <=80 characters long.

```{r warning=FALSE, message=FALSE}
library(gutenbergr)
sawyer_raw <- gutenberg_works(title == "The Adventures of Tom Sawyer") %>%
  gutenberg_download()

sawyer <- sawyer_raw %>%
  tail(-455) %>%  # chapter 1 starts on line 456
  mutate(
    is_chap = str_detect(text, regex("^chapter [\\divxlc]", ignore_case = TRUE)),
    chapter = cumsum(is_chap)
  ) %>%
  select(chapter, text)

skimr::skim(sawyer)
```

## Topic Modeling Chapter

Topic models are generative probabilistic models to quantitatively investigate a large text corpora. In generative probability models, the researcher postulates a model of a data-generating process, then uses the corpus data to find the most likely values for the parameters within the model.

Topic models such as Latent Dirichlet Allocation (LDA) and Structural Topic Modeling (STM) treat documents within a corpora as a "bags of words" and identifies groups of words that tend to co-occur. These groups of words are the topics, formally conceptualized as probability distributions over vocabulary. While the topics are a product of model estimation, the topic name and interpretation are assigned by the researcher by examining the most important words for each topic and the particular texts featuring a topic most prominently. They are not produced to match words and documents to concrete issues specified in advance by the researcher as they would be in a supervised model.

LDA and STM are generative models, meaning you define a data generating process for each document then use the document text to find the most likely values for the model parameters. The generative model defines document-topic and word-topic distributions generating documents. The sum of topic proportions (topical prevalence) in each document is one, and the sum of word probabilities in each topic (topical content) is one. STM differs from LDA in that it uses document metadata as covariates to explain prevalence and topical content. 

Whether you use LDA or STM, you'll start by creating a bag-of-words representation of the data. In the chunk below, I [lemmatize](https://en.wikipedia.org/wiki/Lemmatisation) the tokens and remove my own list of stop words (e.g., "tom" is so common that I don't want to include it in anything.).

```{r}
my_stop_words <- c("tom")

sawyer_tokens <- sawyer %>%
  unnest_tokens(output = "word", input = text, token = "words") %>%
  mutate(word_stem = lemmatize_words(word)) %>%
  anti_join(stop_words, by = "word") %>%
  filter(!word %in% my_stop_words)
```

It's a good idea to explore the data before modeling. One good way to do that is with the TF-IDF statistic. A term's frequency (TF) is its proportion of the words in the document. The inverse document frequency (IDF) is the log of the inverse ratio of documents in which the term appears. The product of TF and IDF (TF-IDF) indicates how important a word is to a document in a collection or corpus. In this case, the TF-IDF indicates how important a word is to the chapter. A term's TF-IDF increases with its frequency in the document (chapter) and decreases with the number of documents (chapters) in the corpus that contain it. Use `tidytext::bind_tf_idf()` to append the `tf`, `idf`, and `tf_idf` columns to your data frame. Figure \@ref(fig:nice-fig).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
sawywer_tf_idf <- sawyer_tokens %>%
  count(chapter, word_stem, sort = TRUE) %>%
  bind_tf_idf(word_stem, chapter, n) %>%
  group_by(chapter) %>%
  slice_max(order_by = tf_idf, n = 10) %>%
  ungroup()

sawywer_tf_idf %>%
  filter(chapter <= 6) %>%
  mutate(word_stem = reorder_within(word_stem, by = tf_idf, within = chapter)) %>%
  ggplot(aes(word_stem, tf_idf, fill = as.factor(chapter))) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(n = 6, name = "Set2"), name = "Topic") +
  facet_wrap(~ chapter, scales = "free", ncol = 3) +
  scale_x_reordered() +
  coord_flip() +
  theme(strip.text=element_text(size=11)) +
  labs(x = NULL, y = "tf-idf",
       title = "Highest tf-idf words in The Adventures of Tom Sawyer (ch 1-6)",
       subtitle = "Individual chapters focus on different characters and narrative elements")
```

### LDA 

LDA treats each document as a mixture of topics (X% topic *A*, Y% topic *B*, etc.), and each topic as a mixture of word probabilities (x% word *a*, y% word *b*, etc.) for all words in the corpus. LDA is implemented in the **topicmodels** package.

Prepare the data by creating a document-term matrix. You will improve performance by removing infrequently appearing tokens with `tm::removeSparseTemrs()`. 

```{r}
sawyer_dtm <- sawyer_tokens %>%
  count(chapter, word_stem) %>%
  cast_dtm(document = chapter, term = word_stem, value = n) %>%
  tm::removeSparseTerms(sparse = 0.9)

sawyer_dtm
```
 
LDA is an optimization algorithm that performs a random search through the parameter space to find the model with the largest log-likelihood. There are multiple search algorithms, but the preferred one appears to be Gibbs sampling, a type of Monte Carlo Markov Chain (MCMC) algorithm.

Create a topic model with `topicmodels::LDA()`. Parameter `k` specifies the number of topics. In general, you only want as many topics as are clearly distinct and that you can easily communicate to others. You can use the perplexity statistic to help identify *k*. Perplexity is a measure of how well a probability model fits a new set of data. Look for the elbow in a scree plot. Here are models with 5 - 50 topics.

```{r message=FALSE}
set.seed(1012)
train_ind <- sample(nrow(sawyer_dtm), floor(0.75*nrow(sawyer_dtm)))
sawyer_train <- sawyer_dtm[train_ind, ]
sawyer_test <- sawyer_dtm[-train_ind, ]

k = c(seq(from = 5, to = 50, by = 5))
perp <- k %>% 
  map(~ LDA(sawyer_train, k = .x, method = "Gibbs", control = list(seed = 1234))) %>%
  map(~ perplexity(.x, newdata = sawyer_test)) %>%
  as.numeric()

data.frame(k = k, perplexity = perp) %>%
  ggplot(aes(x = k, y = perplexity)) + 
  geom_point() + 
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Perplexity Plot for LDM model",
       subtitle = "Elbow at k = 30?")
```

There may be an elbow at *k* = 15 topics. Using STM, the optimal number of clusters was 30. That looks like an elbow to me too. Frankly, if my arm looked like that I would consult an orthopedic surgeon.

```{r}
sawyer_lda <- topicmodels::LDA(sawyer_dtm, k = 30, method = "Gibbs", control = list(seed = 1234))
```

`LDA()` returns two outputs: a "beta" matrix of probabilities of terms belonging to topics; a "gamma" matrix of probabilities of topics contributing to documents. The tidytext package provides a `tidy()` method for extracting these matrices.

```{r}
sawyer_lda_beta <- tidy(sawyer_lda, matrix = "beta") 
sawyer_lda_gamma <- tidy(sawyer_lda, matrix = "gamma", document_names = rownames(sawyer_dfm)) 

# All sums equal 1
# sawyer_lda_beta %>% group_by(topic) %>% summarize(.groups = "drop", sum_beta = sum(beta))
# sawyer_lda_gamma %>% group_by(document) %>% summarize(.groups = "drop", sum_beta = sum(gamma))
```

I have 30 topics here, so it would be hard to show the top words per topic, but here are the first six topics. 

```{r}
sawyer_lda_beta %>%
  filter(topic <= 6) %>%
  group_by(topic) %>%
  slice_max(order_by = beta, n = 10, with_ties = FALSE) %>%
  mutate(term = reorder_within(term, beta, topic)) %>%
  ggplot(aes(x = term, y = beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(n = 6, name = "Set2"), name = "Topic") +
  scale_x_reordered() +
  coord_flip() +
  facet_wrap(~ topic, scales = "free") +
  labs(title = "LDA Beta Matrix for first six topics",
       subtitle = "Showing top 10 word probabilities")
```

You can use a Word cloud as an alternative to the bar chart. The advantage of the word cloud is that it creates an instantaneous impression rather than providing you the analytic means to construct the same conclusion. Here are word clouds for the first two topics.

```{r results='hide'}
colors6 <- RColorBrewer::brewer.pal(n = 6, name = "Set2")
x <- map(c(1:2), ~ with(sawyer_lda_beta %>% filter(topic == .x), 
                        wordcloud::wordcloud(term, beta, max.words = 20,
                                             colors = colors6[.x])))
```

And here are the most prevalent topics across chapters. The figure shows the average probability that the topic appears in the chapter for all chapters.

```{r}
top_terms <- sawyer_lda_beta %>%
  group_by(topic) %>%
  slice_max(order_by = beta, n = 7) %>%
  summarise(.groups = "drop", terms = list(term)) %>%
  mutate(terms = map(terms, paste, collapse = ", ")) %>%
  unnest(terms)

sawyer_lda_gamma %>%
  group_by(topic) %>%
  summarize(.groups = "drop", gamma = mean(gamma)) %>%
  left_join(top_terms, by = "topic") %>%
  mutate(topic = paste("topic", topic),
         topic = fct_reorder(topic, gamma)) %>%
  slice_max(order_by = gamma, n = 10) %>%
  ggplot(aes(x = topic, y = gamma, label = terms)) +
  geom_col(fill = "#D8A7B1", show.legend = FALSE) +
  geom_text(hjust = 0, nudge_y = 0.0005, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0, 0.09),
                     labels = scales::percent_format()) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(x = NULL, y = expression(gamma),
       title = "Top 10 LDA topics by prevalence in The Adentures of Tom Sawyer",
       subtitle = "With top words in each topic")
```

Another way to look at the betas is to identify terms that had the greatest difference in beta between the first and second most probable topic. A good way to do this is with their log ratio, $log_2(\beta_2 / \beta_1)$. Filter for relatively common words having a beta greater than 1/100 in at least one topic.
  
```{r}
sawyer_lda_beta %>% 
  mutate(topic = paste0("topic", topic)) %>%
  group_by(term) %>%
  slice_max(order_by = beta, n = 2) %>%
  summarize(.groups = "drop", min_beta = min(beta), max_beta = max(beta)) %>%
  filter(max_beta > 0.01) %>%
  mutate(log_ratio = log2(max_beta / min_beta)) %>%
  top_n(n = 20, w = abs(log_ratio)) %>%
  arrange(-log_ratio) %>%
  ggplot(aes(x = fct_rev(fct_inorder(term)), y = log_ratio)) +
  geom_col(fill = "#D8A7B1") +
  theme_minimal() +
  coord_flip() + 
  labs(title = "LDA beta matrix log ratios",
       subtitle = "showing greatest differences in beta values",
       x = "", y = "log(beta ratio)")
```

### STM



Prepare the data by creating a document-term matrix, just as you would with LDA, except this time use the `cast_dfm()` to create a document *feature* matrix for the **quanteda** package (the **stm** package uses **quanteda**). As LDA, you will improve performance by removing infrequently appearing tokens, this time with `tm::removeSparseTemrs()`. 

```{r}
sawyer_dfm <- sawyer_tokens %>%
  count(chapter, word_stem) %>%
  cast_dfm(document = chapter, term = word_stem, value = n) %>%
  quanteda::dfm_trim(min_docfreq = 0.1, docfreq_type = "prop")

sawyer_dfm
```

Create a topic model with `stm::stm()`. Parameter `K` specifies the number of topics. In LDA I used perplexity to determine the optimal number of clusters, in part because that is the only optimizing metric it `LDA()` offers. `stm` offers other metrics, including the held-out likelihood and coherence.  Here are models with 5 - 50 topics. This process can take a while, but the **furrr** package and `future_map()` function leverage parallel processing to make it quicker.

```{r}
library(furrr)
plan(multiprocess)

sawyer_stm_mdls <- data_frame(K = seq(from = 5, to = 50, by = 5)) %>%
  mutate(mdl = future_map(K, ~stm(sawyer_dfm, K = ., verbose = FALSE),
                          .options = furrr_options(seed = 123)))

heldout <- make.heldout(sawyer_dfm)

k_result <- sawyer_stm_mdls %>%
  mutate(exclusivity        = map(mdl, exclusivity),
         semantic_coherence = map(mdl, semanticCoherence, sawyer_dfm),
         eval_heldout       = map(mdl, eval.heldout, heldout$missing),
         residual           = map(mdl, checkResiduals, sawyer_dfm),
         bound              = map_dbl(mdl, ~max(.$convergence$bound)),
         lfact              = map_dbl(mdl, ~lfactorial(.$settings$dim$K)),
         lbound             = bound + lfact,
         iterations         = map_dbl(mdl, ~length(.$convergence$bound)))

k_result %>%
  transmute(K,
            `Lower bound` = lbound,
            Residuals = map_dbl(residual, "dispersion"),
            `Semantic coherence` = map_dbl(semantic_coherence, mean),
            `Held-out likelihood` = map_dbl(eval_heldout, "expected.heldout")) %>%
  gather(Metric, Value, -K) %>%
  ggplot(aes(K, Value, color = Metric)) +
  geom_line(size = 1.5, alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(x = "K (number of topics)",
       y = NULL,
       title = "Model diagnostics by number of topics",
       subtitle = "These diagnostics indicate that a good number of topics would be around 60")
```
The held-out likelihood is highest between 30 and 50, and the residuals are lowest at 20, so 30 might be the right number. Semantic coherence is maximized when the most probable words in a given topic frequently co-occur together. Coherence tends to fall as exclusivity increases. You'll want the topic size that balances the trade-off.

```{r}
k_result %>%
  select(K, exclusivity, semantic_coherence) %>%
  filter(K %in% c(20, 25, 30)) %>%
  unnest(cols = c(exclusivity, semantic_coherence)) %>%
  mutate(K = as.factor(K)) %>%
  ggplot(aes(semantic_coherence, exclusivity, color = K)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(x = "Semantic coherence",
       y = "Exclusivity",
       title = "Comparing exclusivity and semantic coherence",
       subtitle = "Models with fewer topics have higher semantic coherence for more topics, but lower exclusivity")
```

It looks like *k* = 30 may be optimal.

```{r}
sawyer_stm <- sawyer_stm_mdls %>%
  filter(K == 30) %>%
  pull(mdl) %>% 
  pluck(1)
```

Like `LDA()`, `stm()` returns two outputs: a "beta" matrix of probabilities of terms belonging to topics; a "gamma" matrix of probabilities of topics contributing to documents. The tidytext package provides a `tidy()` method for extracting these matrices.

```{r}
sawyer_stm_beta <- tidy(sawyer_stm, matrix = "beta") 
sawyer_stm_gamma <- tidy(sawyer_stm, matrix = "gamma", document_names = rownames(sawyer_dfm)) 
```

I have 30 topics here, so it would be hard to show the top words per topic, but here are the first six topics. 

```{r}
sawyer_stm_beta %>%
  filter(topic <= 6) %>%
  group_by(topic) %>%
  slice_max(order_by = beta, n = 10, with_ties = FALSE) %>%
  mutate(term = reorder_within(term, beta, topic)) %>%
  ggplot(aes(x = term, y = beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(n = 6, name = "Set2"), name = "Topic") +
  scale_x_reordered() +
  coord_flip() +
  facet_wrap(~ topic, scales = "free") +
  labs(title = "STM Beta Matrix for first six topics",
       subtitle = "Showing top 10 word probabilities")
```

Here is a Word cloud representation.

```{r results='hide'}
colors6 <- RColorBrewer::brewer.pal(n = 6, name = "Set2")
x <- map(c(1:2), ~ with(sawyer_stm_beta %>% filter(topic == .x), 
                        wordcloud::wordcloud(term, beta, max.words = 20,
                                             colors = colors6[.x])))
```

And here are the most prevalent topics across chapters.

```{r}
top_terms <- sawyer_stm_beta %>%
  group_by(topic) %>%
  slice_max(order_by = beta, n = 7) %>%
  summarise(.groups = "drop", terms = list(term)) %>%
  mutate(terms = map(terms, paste, collapse = ", ")) %>%
  unnest(terms)

sawyer_stm_gamma %>%
  group_by(topic) %>%
  summarize(.groups = "drop", gamma = mean(gamma)) %>%
  left_join(top_terms, by = "topic") %>%
  mutate(topic = paste("topic", topic),
         topic = fct_reorder(topic, gamma)) %>%
  slice_max(order_by = gamma, n = 10) %>%
  ggplot(aes(x = topic, y = gamma, label = terms)) +
  geom_col(fill = "#D8A7B1", show.legend = FALSE) +
  geom_text(hjust = 0, nudge_y = 0.0005, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0, 0.09),
                     labels = scales::percent_format()) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(x = NULL, y = expression(gamma),
       title = "Top 10 STM topics by prevalence in The Adentures of Tom Sawyer",
       subtitle = "With top words in each topic")
```

Another way to look at the betas is to identify terms that had the greatest difference in beta between the first and second most probable topic. A good way to do this is with their log ratio, $log_2(\beta_2 / \beta_1)$. Filter for relatively common words having a beta greater than 1/100 in at least one topic.
  
```{r}
sawyer_stm_beta %>% 
  mutate(topic = paste0("topic", topic)) %>%
  group_by(term) %>%
  slice_max(order_by = beta, n = 2) %>%
  summarize(.groups = "drop", min_beta = min(beta)+.001, max_beta = max(beta)+.001) %>%
  filter(max_beta > 0.01) %>%
  mutate(log_ratio = log2(max_beta / min_beta)) %>%
  top_n(n = 20, w = abs(log_ratio)) %>%
  arrange(-log_ratio) %>%
  ggplot(aes(x = fct_rev(fct_inorder(term)), y = log_ratio)) +
  geom_col(fill = "#D8A7B1") +
  theme_minimal() +
  coord_flip() + 
  labs(title = "STM beta matrix log ratios",
       subtitle = "showing greatest differences in beta values",
       x = "", y = "log(beta ratio)")
```

## Sentiment Analysis

A typical sentiment analysis involves unnesting tokens with `unnest_tokens()`, assigning sentiments with `inner_join(sentiments)`, counting tokens with `count()`, and summarizing and visualizing.

The tidytext package contains four sentiment lexicons, all based on unigrams. 

* **nrc**. binary “yes”/“no” for categories positive, negative, anger, anticipation, disgust, fear, joy, sadness, surprise, and trust. 
* **bing**. “positive”/“negative” classification. 
* **AFINN**. score between -5 (most negative) and 5 (most positive).
* **loughran**. "positive"/"negative"/"litigious"/"uncertainty"/"constraining"/"superflous" classification.

You can view the sentiment assignments with `get_sentiments(lexicon = c("afinn", "bing", nrc", "laughlin"))`

```{r}
x1 <- get_sentiments(lexicon = "nrc") %>%
  count(sentiment) %>%
  mutate(lexicon = "nrc")
x2 <- get_sentiments(lexicon = "bing") %>%
  count(sentiment) %>%
  mutate(lexicon = "bing")
x3 <- get_sentiments(lexicon = "afinn") %>%
  count(value) %>%
  mutate(lexicon = "afinn") %>%
  mutate(sentiment = as.character(value)) %>%
  select(-value)
x4 <- get_sentiments(lexicon = "loughran") %>%
  count(sentiment) %>%
  mutate(lexicon = "loughran")
x <- bind_rows(x1, x2, x3, x4)

ggplot(x, aes(x = fct_reorder(sentiment, n), y = n, fill = lexicon)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Sentiment Counts", x = "", y = "") +
  facet_wrap(~ lexicon, scales = "free")
```

Here is a sentiment analysis of sections of 80 lines of Jane Austin's books.  *(Small sections may not have enough words to get a good estimate of sentiment, and large sections can wash out the narrative structure.  80 lines seems about right.)*

```{r}
# austin_tidy %>%
#   inner_join(get_sentiments("bing")) %>%
#   count(book, index = linenumber %/% 80, sentiment) %>%
#   pivot_wider(names_from = sentiment, values_from = n, values_fill = list(n = 0)) %>%
#   mutate(sentiment = positive - negative) %>%
#   ggplot(aes(x = index, y = sentiment, fill = book)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(~book, ncol = 2, scales = "free_x")
```

Fair to say Jane Austin novels tend to have a happy ending?  The three sentiment lexicons provide different views of THE data.  Here is a comparison of the lexicons using one of Jane Austin's novels, "Pride and Prejudice".

```{r}
# # AFINN lexicon measures sentiment with a numeric score between -5 and 5.
# afinn <- austin_tidy %>% 
#   filter(book == "Pride & Prejudice") %>%
#   inner_join(get_sentiments("afinn"), by = "word") %>% 
#   group_by(index = linenumber %/% 80) %>% 
#   summarise(sentiment = sum(value)) %>% 
#   mutate(method = "AFINN")
# 
# # Bing and nrc categorize words in a binary fashion, either positive or negative.
# bing <- austin_tidy %>%
#   filter(book == "Pride & Prejudice") %>%
#   inner_join(get_sentiments("bing"), by = "word") %>% 
#   count(index = linenumber %/% 80, sentiment) %>%
#   pivot_wider(names_from = sentiment, values_from =  n, values_fill = list(n = 0)) %>%
#   mutate(sentiment = positive - negative) %>% 
#   mutate(method = "Bing") %>%
#   select(index, sentiment, method)
# 
# nrc <- austin_tidy %>%
#   filter(book == "Pride & Prejudice") %>%
#   inner_join(get_sentiments("nrc") %>% filter(sentiment %in% c("positive", "negative")), by = "word") %>% 
#   count(index = linenumber %/% 80, sentiment) %>%
#   pivot_wider(names_from = sentiment, values_from = n, values_fill = list(n = 0)) %>%
#   mutate(sentiment = positive - negative) %>% 
#   mutate(method = "NRC") %>%
#   select(index, sentiment, method)
# 
# bind_rows(afinn, bing, nrc) %>%
#   ggplot(aes(index, sentiment, fill = method)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(~method, ncol = 1, scales = "free_y")
```

In this example, and in general, **NRC** sentiment tends to be high, **AFINN** sentiment has more variance, and **Bing** sentiment finds longer stretches of similar text.  However, all three agree roughly on the overall trends in the sentiment through a narrative arc.

What are the top-10 positive and negative words?  Using the Bing lexicon, get the counts, then `group_by(sentiment)` and `top_n()` to the top 10 in each category.  

```{r}
# austin_tidy %>%
#   filter(book == "Pride & Prejudice") %>%
#   inner_join(get_sentiments("bing"), by = "word") %>% 
#   count(word, sentiment, sort = TRUE) %>%
#   group_by(sentiment) %>%
#   top_n(n = 10, wt = n) %>%
#   ggplot(aes(x = fct_reorder(word, n), y = n, fill = sentiment)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(~sentiment, scales = "free_y") +
#   coord_flip() +
#   labs(y = "Contribution to Sentiment",
#        x = "")
```

Uh oh, "miss" is a red-herring - in Jane Austin novels it often refers to an unmarried woman.  Drop it from the analysis by appending it to the stop-words list.

```{r}
# austin_tidy %>%
#   anti_join(bind_rows(stop_words,
#                       tibble(word = c("miss"), lexicon = c("custom")))) %>%
#   filter(book == "Pride & Prejudice") %>%
#   inner_join(get_sentiments("bing")) %>% 
#   count(word, sentiment, sort = TRUE) %>%
#   group_by(sentiment) %>%
#   top_n(n = 10, wt = n) %>%
#   ggplot(aes(x = fct_reorder(word, n), y = n, fill = sentiment)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(~sentiment, scales = "free_y") +
#   coord_flip() +
#   labs(y = "Contribution to Sentiment",
#        x = "")
```

Better!

A common way to visualize sentiments is with a word cloud.  

```{r}
# austin_tidy %>%
#   anti_join(bind_rows(stop_words,
#                       tibble(word = c("miss"), lexicon = c("custom")))) %>%
#   filter(book == "Pride & Prejudice") %>%
#   count(word) %>%
#   with(wordcloud(word, n, max.words = 100))
```

`comparison.cloud` is another implementation of a word cloud.  It takes a matrix input.

```{r}
# x <- austin_tidy %>%
#   anti_join(bind_rows(stop_words,
#                       tibble(word = c("miss"), lexicon = c("custom")))) %>%
#   inner_join(get_sentiments("bing")) %>% 
#   filter(book == "Pride & Prejudice") %>%
#   count(word, sentiment, sort = TRUE) %>%
#   pivot_wider(names_from = sentiment, values_from = n, values_fill = list(n = 0)) %>%
#   as.data.frame()
# rownames(x) <- x[,1]
# comparison.cloud(x[, 2:3])
# rm(x)
```

Sometimes it makes more sense to analyze entire sentences.  Specify `unnest_tokens(..., token = "sentences")` to override the default `token = "word"`.

```{r}
# austen_books() %>%
#   group_by(book) %>%
#   mutate(linenumber = row_number(),
#          chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]",
#                                                  ignore_case = TRUE)))) %>%
#   ungroup() %>%
#   unnest_tokens(output = word, input = text, token = "sentences")

```

### N-Grams

Create n-grams by specifying `unnest_tokens(..., token = "ngrams", n)` where `n = 2` is a bigram, etc.  To remove the stop words, `separate` the n-grams, then filter on the `stop_words` data set.

```{r}
# austin.2gram <- austen_books() %>%
#   group_by(book) %>%
#   mutate(linenumber = row_number(),
#          chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]",
#                                                  ignore_case = TRUE)))) %>%
#   ungroup() %>%
#   unnest_tokens(output = bigram, input = text, token = "ngrams", n = 2)
# 
# austin.2gram <- austin.2gram %>%
#   separate(bigram, c("word1", "word2"), sep = " ") %>%
#   filter(!word1 %in% stop_words$word &
#            !word2 %in% stop_words$word &
#            !is.na(word1) & !is.na(word2)) %>%
#   unite(bigram, word1, word2, sep = " ")
# 
# austin.2gram %>%
#   count(book, bigram, sort = TRUE)
```

Here are the most commonly mentioned streets in Austin's novels.

```{r}
# austin.2gram %>%
#   separate(bigram, c("word1", "word2"), sep = " ") %>%
#   filter(word2 == "street") %>%
#   count(book, word1, sort = TRUE)
```

Use the TF-IDF statistic to compare words among documents. Calculate the cosine similarity, the angle in multidimensional space between two vectors ($cos(\theta) = (A \cdot B) / ||A||||B||)$), to determine how similar two items are. Use `widyr::pairwise_similarity()` to calculate the cosine similarity of all pairs of items in a tidy table.

```{r}
# austen %>%
#   unnest_tokens(output = "word", input = "text", token = "words") %>%
#   anti_join(stop_words, by = "word") %>%
#   count(book, word) %>%
#   bind_tf_idf(term = word, document = book, n = n) %>%
#   pairwise_similarity(item = book, feature = word, value = tf_idf) %>%
#   arrange(desc(similarity))
```

```{r}
# austen_books() %>%
#   group_by(book) %>%
#   mutate(linenumber = row_number(),
#          chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]",
#                                                  ignore_case = TRUE)))) %>%
#   ungroup() %>%
#   unnest_tokens(output = bigram, input = text, token = "ngrams", n = 2) %>%
#   separate(bigram, c("word1", "word2"), sep = " ") %>%
#   filter(!word1 %in% stop_words$word &
#            !word2 %in% stop_words$word &
#            !is.na(word1) & !is.na(word2)) %>%
#   unite(bigram, word1, word2, sep = " ")
# austin.2gram %>%
#   count(book, bigram) %>%
#   bind_tf_idf(bigram, book, n) %>%
#   group_by(book) %>%
#   top_n(n = 10, wt = tf_idf) %>%
#   ggplot(aes(x = fct_reorder(bigram, n), y = tf_idf, fill = book)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(~book, scales = "free_y", ncol = 2) +
#   labs(y = "tf-idf of bigram to novel") +
#   coord_flip() 
```


A good way to visualize bigrams is with a network graph.  Packages `igraph` and `ggraph` provides tools for this purpose.

```{r}
# set.seed(2016)
# 
# austen_books() %>%
#   group_by(book) %>%
#   mutate(linenumber = row_number(),
#          chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]",
#                                                  ignore_case = TRUE)))) %>%
#   ungroup() %>%
#   unnest_tokens(output = bigram, input = text, token = "ngrams", n = 2) %>%
#   separate(bigram, c("word1", "word2"), sep = " ") %>%
#   filter(!word1 %in% stop_words$word &
#            !word2 %in% stop_words$word &
#            !is.na(word1) & !is.na(word2)) %>%
#   count(word1, word2) %>%
#   filter(n > 20) %>%
#   graph_from_data_frame() %>%  # creates unformatted "graph"
#   ggraph(layout = "fr") +
#   geom_edge_link(aes(edge_alpha = n), 
#                  show.legend = FALSE,
#                  arrow = grid::arrow(type = "closed", 
#                                      length = unit(.15, "inches")), 
#                  end_cap = circle(.07, 'inches')) +
#   geom_node_point(color = "lightblue", 
#                   size = 5) +
#   geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
#   theme_void()
```


If you want to count the number of times that two words appear within the same document, or to see how correlated they are, widen the data with the `widyr` package.

```{r}
# austen_books() %>%
#   filter(book == "Pride & Prejudice") %>%
#   mutate(section = row_number() %/% 10) %>%
#   filter(section > 0) %>%
#   unnest_tokens(word, text) %>%
#   filter(!word %in% stop_words$word) %>%
#   pairwise_count(word, section, sort = TRUE)
```

The correlation among words is how often they appear together relative to how often they appear separately.  The phi coefficient is defined

$$\phi = \frac{n_{11}n_{00} - n_{10}n_{01}}{\sqrt{n_{1.}n_{0.}n_{.1}n_{.0}}}$$

where $n_{10}$ means number of times section has word x, but not word y, and $n_{1.}$ means total times section has word x.  This lets us pick particular interesting words and find the other words most associated with them.

```{r}
# austen_books() %>%
#   filter(book == "Pride & Prejudice") %>%
#   mutate(section = row_number() %/% 10) %>%
#   filter(section > 0) %>%
#   unnest_tokens(word, text) %>%
#   filter(!word %in% stop_words$word) %>%
#   pairwise_cor(word, section, sort = TRUE) %>%
#   filter(item1 %in% c("elizabeth", "pounds", "married", "pride")) %>%
#   group_by(item1) %>%
#   top_n(n = 4) %>%
#   ungroup() %>%
#   mutate(item2 = reorder(item2, correlation)) %>%
#   ggplot(aes(x = item2, y = correlation)) +
#   geom_bar(stat = "identity") +
#   facet_wrap(~ item1, scales = "free") +
#   coord_flip()
```

You can use the correlation to set a threshold for a graph.

```{r eval=FALSE, include=FALSE}
set.seed(2016)

austen_books() %>%
  filter(book == "Pride & Prejudice") %>%
  mutate(section = row_number() %/% 10) %>%
  filter(section > 0) %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word) %>%
  # filter to relatively common words
  group_by(word) %>%
  filter(n() > 20) %>%
  pairwise_cor(word, section, sort = TRUE) %>%
  filter(correlation > .15) %>%  # relatively correlated words
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(edge_alpha = correlation), show.legend = FALSE) +
  geom_node_point(color = "lightblue", size = 5) +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_void()
```


### Converting to and from non-tidy formats

One of the most common objects in text mining packages is the document term matrix (DTM) where each row is a document, each column a term, and each value an appearance count.  The `broom` package contains functions to convert between DTM and tidy formats.

Convert a DTM object into a tidy data frame with `tidy()`.  Convert a tidy object into a sparse matrix with `cast_sparse()`, into a DTM with `cast_dtm()`, and into a "dfm" for quanteda with `cast_dfm()`.  

```{r eval=FALSE, include=FALSE}
data("AssociatedPress", package = "topicmodels")
AssociatedPress
Terms(AssociatedPress) %>% head()
```

Create a tidy version of `AssociatedPress` with `tidy()`.

```{r eval=FALSE, include=FALSE}
ap_td <- tidy(AssociatedPress)
ap_td %>% 
  inner_join(get_sentiments("bing"), by = c(term = "word")) %>%
  count(sentiment, term, wt = count) %>%
  ungroup() %>%
  filter(n >= 200) %>%
  mutate(n = ifelse(sentiment == "negative", -n, n)) %>%
  arrange(n) %>%
  ggplot(aes(x = fct_inorder(term), y = n, fill = sentiment)) +
  geom_bar(stat = "identity") +
  labs(title = "Most Common AP Sentiment Words",
       x = "",
       y = "Contribution to Sentiment") +
  theme(legend.position = "top", 
        legend.title = element_blank()) +
  coord_flip()
```

The document-feature matrix `dfm` class from the `quanteda` text-mining package is another implementation of a document-term matrix.  Here are the terms most specific (highest tf-idf) from each of four selected inaugural addresses.  

```{r eval=FALSE, include=FALSE}
data("data_corpus_inaugural", package = "quanteda")
inaug_dfm <- quanteda::dfm(data_corpus_inaugural, verbose = FALSE)
head(inaug_dfm)

inaug_td <- tidy(inaug_dfm)
head(inaug_td)

inaug_td %>% 
  bind_tf_idf(term = term, document = document, n = count) %>%
  group_by(document) %>%
  top_n(n = 10, wt = tf_idf) %>%
  ungroup() %>%
  filter(document %in% c("1861-Lincoln", "1933-Roosevelt", "1961-Kennedy", "2009-Obama")) %>%
  arrange(document, desc(tf_idf)) %>%
  ggplot(aes(x = fct_rev(fct_inorder(term)), y = tf_idf, fill = document)) +
  geom_col() +
  labs(x = "") +
  theme(legend.position = "none") +
  coord_flip() +
  facet_wrap(~document, ncol = 2, scales = "free")
```

And here is word frequency trend ocer time for six selected terms. (problem with `extract()` below).

```{r eval=FALSE, include=FALSE}
# inaug_td %>%
#   extract(document, "year", "(\\d+)", convert = TRUE) %>%
#   complete(year, term, fill = list(count = 0)) %>%
#   group_by(year) %>%
#   mutate(year_total = sum(count)) %>%
#   filter(term %in% c("god", "america", "foreign", "union", "constitution", "freedom")) %>%
#   ggplot(aes(x = year, y = count / year_total)) + 
#   geom_point() +
#   geom_smooth() +
#   facet_wrap(~ term, scales = "free_y") +
#   scale_y_continuous(labels = scales::percent_format()) +
#   labs(y = "",
#        title = "% frequency of word in inaugural address")
```

Cast tidy data into document-term matrix with `cast_dtm()`, quanteda's dfm with `cast_dfm()`, and sparese matrix with `cast_sparse()`.

```{r eval=FALSE, include=FALSE}
inaug_dtm <- cast_dtm(data = inaug_td, document = document, term = term, value = count)
inaug_dfm <- cast_dfm(data = inaug_td, document = document, term = term, value = count)
inaug_sparse <- cast_sparse(data = inaug_td, row = document, column = term, value = count)
```

An untokenized document collection is called a *corpus*. The corpuse may include metadata, such as ID, date/time, title, language, etc.  Corpus metadata is usually stored as lists.  Use `tidy()` to construct a table, one row per document.

```{r eval=FALSE, include=FALSE}
data("acq")
print(acq)
acq[[1]]  # first document

acq_td <- tidy(acq)
acq_td
```

### Example

Library `tm.plugin.webmining` connects to online feeds to retrieve news articles based on a keyword.

```{r eval=FALSE, include=FALSE}
library(tm.plugin.webmining)
library(purrr)

company <- c("Progressive", "Microsoft", "Apple")
symbol <- c("PRG", "MSFT", "AAPL")
download_articles <- function(symbol) {
  WebCorpus(GoogleFinanceSource(paste0("NASDAQ:", symbol)))
}
#stock_articles <- tibble(company = company,
#                         symbol = symbol) %>%
#  mutate(corpus = map(symbol, download_articles))
#download_articles("MSFT")
#?GoogleFinanceSource()
#corpus <- Corpus(GoogleFinanceSource("NASDAQ:MSFT"))
```

## Text Classification Modeling

The `tweets` data set contains politically charged tweets. Can you train a model to predict the `account_type` (Left or Right) from the tweet text? 

Prepare the data by filtering to just the relevant tweets, tokenizing, and creatiing a document-term matrix with TFIDF weighting. If resources are an issue, experiment with values of parameter `sparse` in `dt::removeSparseTerms()` to get less terms. Below, `sparse = 0.9999` finally gets a respectable term count 14% the size of the orginal count. 

```{r eval=FALSE, include=FALSE}
tweets_words <- tweets[1:500, ] %>%
  unnest_tokens(output = "word", input = content, token = "words") %>%
  anti_join(stop_words, by = "word") %>%
  mutate(word = SnowballC::wordStem(word))

tweets_dtm <- tweets_words %>%
  count(tweet_id, word) %>%
  cast_dtm(document = tweet_id, term = word, value = n, weighting = tm::weightTfIdf) %>%
  tm::removeSparseTerms(sparse = 0.999)
print(tweets_dtm)
```

Split the data into an 80:20 train:test split and fit a random forest model.

```{r eval=FALSE, include=FALSE}
set.seed(1012)
train_ind <- sample(nrow(tweets_dtm), floor(0.75*nrow(tweets_dtm)))
tweets_train <- tweets_dtm[train_ind, ]
tweets_test <- tweets_dtm[-train_ind, ]

rfc <- randomForest::randomForest(x = as.data.frame(as.matrix(tweets_train)),
                                  y = as.numeric(tweets_train$dimnames$Docs),
                                  nTree = 50)

rfc
plot(caret::varImp(rfc), main="Variable Importance with Random Forest")
```


## Named Entity Recognition

The **qdap** package provides parsing tools for preparing transcript data. 
```{r warning=FALSE, message=FALSE}
library(qdap)
```

For example, `freq_terms()` parses text and counts the terms.

```{r eval=FALSE, include=FALSE}
freq_terms(DATA$state, top = 5)
```

You can also plot the terms.

```{r eval=FALSE, include=FALSE}
plot(freq_terms(DATA$state, top = 5))
```

There are two kinds of the corpus data types, the permanent corpus, **PCorpus**, and the volatile corpus, **VCorpus**. The volatile corpus is held in RAM rather than saved to disk.  Create a volatile corpus with `tm::vCorpous()`. vCorpous() takes either a text source created with `tm::VectorSource()` or a dataframe source created with `Dataframe Source()` where the input dataframe has cols `doc_id`, `text_id` and zero or more metadata columns.

```{r eval=FALSE, include=FALSE}
tweets <- read_csv(file = "https://assets.datacamp.com/production/repositories/19/datasets/27a2a8587eff17add54f4ba288e770e235ea3325/coffee.csv")
coffee_tweets <- tweets$text

coffee_source <- VectorSource(coffee_tweets)
coffee_corpus <- VCorpus(coffee_source)

# Text of first tweet
coffee_corpus[[1]][1]
```

In bag of words text mining, cleaning helps aggregate terms, especially words with common stems like "miner" and "mining".  There are several functions useful for preprocessing: `tolower()`, `tm::removePunctuation()`, `tm::removeNumbers()`, `tm::stripWhiteSpace()`, and `removeWords()`.  Apply these functions to the documents in a `VCorpus` object with `tm_map()`.  If the function is not one of the pre-defined functions, wrap it in `content_transformer()`.  Another preprocessing function is `stemDocument()`.

```{r eval=FALSE, include=FALSE}
# Create the object: text
text <- "<b>She</b> woke up at       6 A.M. It\'s so early!  She was only 10% awake and began drinking coffee in front of her computer."
tolower(text)
removePunctuation(text)
removeNumbers(text)
stripWhitespace(text)
```

The **qdap** package offers other preprocessing functions.

```{r eval=FALSE, include=FALSE}
text <- "<b>She</b> woke up at       6 A.M. It\'s so early!  She was only 10% awake and began drinking coffee in front of her computer."
bracketX(text)
replace_number(text)
replace_abbreviation(text)
replace_contraction(text)
replace_symbol(text)
```

`tm::stopwords("en")` returns a vector of stop words.  You can add to the list with concatenation.

```{r eval=FALSE, include=FALSE}
new_stops <- c("coffee", "bean", stopwords("en"))
removeWords(text, new_stops)
```

`tm::stemDocument()` and `tm::stemCompletion()` reduce the variation in terms.

```{r eval=FALSE, include=FALSE}
complicate <- c("complicated", "complication", "complicatedly")

stem_doc <- stemDocument(complicate)

comp_dict <- c("complicate")

complete_text <- stemCompletion(stem_doc, comp_dict)
complete_text
```


```{r eval=FALSE, include=FALSE}
clean_corpus <- function(corpus){
  corpus <- tm_map(corpus, removePunctuation)
  corpus <- tm_map(corpus, content_transformer(tolower))
  corpus <- tm_map(corpus, removeWords, words = c(stopwords("en"), "coffee", "mug"))
  corpus <- tm_map(corpus, stripWhitespace)
  return(corpus)
}
clean_corp <- clean_corpus(coffee_corpus)
content(clean_corp[[1]])

# compare to original
content(coffee_corpus[[1]])
```

To perform the analysis of the tweets, convert the corpus into either a document term matrix (DTM, documents as rows, terms as cols), or a term document matrix (TDM, terms as rows, documents as cols). 

```{r eval=FALSE, include=FALSE}
coffee_tdm <- TermDocumentMatrix(clean_corp)

# Print coffee_tdm data
coffee_tdm

# Convert coffee_tdm to a matrix
coffee_m <- as.matrix(coffee_tdm)

# Print the dimensions of the matrix
dim(coffee_m)

# Review a portion of the matrix
coffee_m[c("star", "starbucks"), 25:35]
```


## Tidy Text

Here are four Jane Austin books from the `janeaustenr`. "Sense & Sensibility" acts as the baseline count, and the other books are faceted for comparison.

```{r eval=FALSE, include=FALSE}
custom_stop_words <- tribble(
  ~word, ~lexicon,
  "edward", "CUSTOM",
  "frank", "CUSTOM",
  "thomas", "CUSTOM",
  "fanny", "CUSTOM",
  "anne", "CUSTOM"
)

austin_tidy <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]", ignore_case = TRUE)))
    ) %>%
  ungroup() %>%
  unnest_tokens(output = word, input = text) %>%
  anti_join(stop_words) %>%
  anti_join(custom_stop_words)

austin_tidy %>%
  count(book, word) %>%
  group_by(book) %>%
  mutate(proportion = n / sum(n)) %>% 
  select(-n) %>% 
  pivot_wider(names_from = book, values_from = proportion) %>%
  pivot_longer(cols = `Pride & Prejudice`:`Persuasion`, names_to = "book", values_to = "proportion") %>%
  ggplot(aes(x = proportion, 
             y = `Sense & Sensibility`,
             color = abs(`Sense & Sensibility` - proportion))) +
  geom_abline(color = "gray40", lty = 2) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.3, height = 0.3) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = percent_format()) +
  scale_y_log10(labels = percent_format()) +
  scale_color_gradient(limits = c(0, 0.001), low = "darkslategray4", high = "gray75") +
  facet_wrap(~book, ncol = 2) +
  theme(legend.position = "none") +
  labs(y = "Sense & Sensibility", x = NULL)
```


## Appendix: Regular Expressions

There are two choices for string manipulation. Base R has `grep()` and `gsub()`. Or you can use the **stringr** package. Both will use the same regular expression patterns. Here are some important ones.

### Base R

`grep()` returns the indices of pattern matches in a vector of strings. Specify parameter `value = TRUE` to get the strings that match the pattern. 

```{r collapse=TRUE}
str <- c("abc123", "xyz   789.")
grep("\\w+123", str)   # \w any alphanumeric, + for any length
grep("abc\\d", str)  # \d any digit
grep("\\s", str) # s spaces, \S non-spaces
grep("\\.", str) # . anything, so \. for period
```



### stringr

The **stringr** package is a simple wrapper around the more complete **stringi** package.

Example of finding rows in data frame of book text with "chapter [#]". In `regex()`, "[]" means "any of", "\\d" is any number, and "v", "x", "l", and "c" are just roman numerals. Here is the word-count per chapter.

```{r}
austen <- janeaustenr::austen_books()
chapter_regex <- stringr::regex("^chapter [\\divxlc]", ignore_case = TRUE)
austen %>%
  filter(book == "Emma") %>%
  mutate(linenumber = row_number(),
         chapter = cumsum(str_detect(text, chapter_regex))) %>%
  ungroup() %>%
  unnest_tokens(output = "word", input = text, token = "words") %>%
  count(chapter)
```


* `str_c()` concatenates strings, similar to with `paste()` and `paste0()`.  
```{r}
str_c("hello", "world", sep = " ")
```

* `str_replace(string, pattern, replacment)` replaces `pattern` with `replacement`.  
```{r}
str_replace("If the future's looking bleek", pattern = fixed("bleek"), replacement = "dark")
```

* `str_replace_na(string, replacement)` replaces NAs.
```{r}
str_replace_na(c("We're the ones ", NA, "have to shine"), replacement = "who ")
```

* `str_split(string, pattern, simplify = FALSE)` splits `string` by `pattern` into a list of vectors, or matrix if `simplify = TRUE`. 
```{r}
str_split("If there's no one in control", pattern = " ", simplify = TRUE)
```

* `str_c(..., sep)` concatenates a vector of strings, separated by `sep`.
```{r}
str_c("we're", "the", "ones", "who", "draw", "the", "line", sep = " ")
```

`str_sub(string, start, end)` returns substring of `string` from `start` to `end`.  Use negatives to start from the end of the string.
```{r}
my_str <- "Although we live in trying times"
str_sub(my_str, start = 1, end = 5)
str_sub(my_str, start = -4, end = -1)
```

* `str_length(string)` returns the number of characters in a string.
```{r}
str_length("We're the ones who have to try")
```

* `str_detect(string, pattern)` returns booleans where `string` matches `pattern`.
```{r}
str_detect(c("Although we know", "that time",  "has wings"), pattern = fixed("wings"))
```

* `str_match(string, pattern)` returns matching strings where `string` matches `pattern`.
```{r}
str_match(c("Although we know", "that time",  "has wings"), pattern = "wings")
```

* `str_subset(string, pattern)` returns string matches where `string` matches `pattern`.
```{r}
str_subset(c("Although we know", "that time",  "has wings"), pattern = fixed("wings"))
```

* `str_count(string, pattern)` returns a count of matches where `string` matches `pattern`.
```{r}
str_count(c("Although we know", "that time",  "has wings"), pattern = fixed("wings"))
```

* `str_extract(string, pattern)` returns the part of the `string` matching `pattern`.
```{r}
str_extract(c("We're the ones", "who have to fly"), pattern = " t..")
```

### Regular Expressions

The **stringr** package functions use [ICU regular expressions](https://unicode-org.github.io/icu/). Here are common regular expressions.

```{r eval=FALSE}
str_subset(sawyer$text, "^tom")  # starts with "tom"
str_subset(sawyer$text, "[0-9]")  # contains a digit
str_subset(sawyer$text, "^[0-9]")  # begins with a digit
str_subset(sawyer$text, "[^0-9]")  # does not contain a digit
str_subset(sawyer$text, ".handsome.")  # contains "handsome"
sawyer %>% filter(str_detect(sawyer$text, "handsome"))  # rows with "handsome"
sawyer %>% filter(str_detect(sawyer$text, ".handsome."))  # same thing

# Using perl-style regexp, 
# "(.)" is a subexpression (defined with parens) consisting of any character 
# defined with a dot. "\\U" makes the next character uppercase. \\1 is a 
# reference to subexpression 1
sub("(.)", ("\\U\\1"), sawyer_tokens$word, perl=TRUE)
str_to_title(sawyer_tokens$word)  # nearly the same - converts first letter, not char
str_to_title(sawyer$text)

# [] defines a set of chars
str_subset(sawyer$text, "^[qQ]")  # starts with "q" or "Q"
str_subset(sawyer$text, "(^[\\_])(.*-)")  # starts with "_" and contains "-"
str_subset(sawyer$text, "dog.*tricks|tricks.*dog")  # contains "dog" & "tricks"
```

## Appendix: tidytext

The tidy text format is a table with one token (meaningful unit of text, such as a word) per row.  The **tidytext** package assists with the major tasks in text analysis. 

A typical text analysis using tidy data principles unnests tokens with `unnest_tokens()`, anti-joins unimportant "stop words" from `tidytext::stop_words`, and summarizes token counts `count()`. In between you might also reduce words to their stems with `SnowballC::wordStem()`.

```{r eval=FALSE, include=FALSE}
custom_stop_words <- tribble(
  ~word, ~lexicon,
  "t.co", "custom",
  "https", "custom",
  "http", "custom",
  "di", "custom",
  "il", "custom",
  "la", "custom"
)

tweets_tidy <- tweets %>%
  unnest_tokens(output = "word", input = content, token = "words") %>%
  filter(xfun::is_ascii(word) == TRUE) %>%  # garbage chars
  anti_join(stop_words, by = "word") %>%
  anti_join(custom_stop_words, by = "word") %>%
  mutate(word = SnowballC::wordStem(word))
```

Summarize the tidy text into a "bag of words" (table of word frequencies). Once in a bag of words format, you can feed into a word cloud or other ggplot geom.

```{r fig.width=6.5, fig.height=5, eval=FALSE, include=FALSE}
tweets_bow <- tweets_tidy %>%
  count(word)

wordcloud(
  words = tweets_bow$word,
  freq = tweets_bow$n,
  max.words = 30,
  colors = brewer.pal(9,"BuGn")[-(1:4)]
)
```

To perform topic modeling, cast the bag of words into a document term matrix with `cast_dtm()`.

```{r eval=FALSE, include=FALSE}
tweets_dtm <- tweets_bow %>%
  cast_dtm(document = tweet_id, term = word, value = n, weighting = tm::weightTfIdf)
```

After performing a topic model, cast the output (either a beta matrix or gamma matrix) into tidy format with `tidy()`.

```{r eval=FALSE, include=FALSE}
tweets_lda <- topicmodels::LDA(tweets_dtm, k = 2, method = "Gibbs", control = list(seed = 1234))
tweets_gamma <- tidy(tweets_lda, matrix = "gamma")
tweets_beta <- tidy(tweets_lda, matrix = "beta")

tweets_beta_1 <- tweets_beta %>%
  filter(topic == 1) %>%
  mutate(n = trunc(beta * 10000))

# wordcloud requires integer frequencies, so convert betat
wordcloud::wordcloud(
  words = tweets_beta_1$term,
  freq = trunc(tweets_beta_1$beta * 1000),
  max.words = 30,
  colors = RColorBrewer::brewer.pal(9,"BuGn")[-(1:4)]
)
```



## Appendix: tm

The Text Mining package **tm** is a text mining framework for importing data, handling a corpus, and creating term-document matrices.

The main structure document structure is a corpus, implemented either as an in-memory *volatile corpus* `VCorpus()` or on-disk *permanent corpus* `Pcorpus()`. Create a corpus from a directory of text documents

```{r eval=FALSE}
my_rmd_corpus <- VCorpus(DirSource(pattern = "*.Rmd"))
```

or from a data frame using helper function `VectorSource()`.

```{r eval=FALSE, include=FALSE}
docs <- c("This is a text.", "This another one.")
my_vctr_corpus <- VCorpus(VectorSource(docs))
```

A corpus is a list of document objects, each containing `meta` data and `content`. Here is the first document of the one I just created.

```{r eval=FALSE, include=FALSE}
my_vctr_corpus[[1]]$meta
my_vctr_corpus[[1]]$content
```

You can perform transformations on your `tm` object, including removing extra whitespace `tm_map(x, stripWhitespace)`, converting to lowercase `tm_map(x, content_transformer(tolower))`, removing stopwords `tm_map(x, removeWords, stopwords("english"))`, and stemming `tm_map(x stemDocument)`.

Some packages (e.g., **topicmodels**) operate on a "bag of words" representation called a *document term matrix*. A bag of words is a frequency count of words. Convert the *tm* object into a matrix with `TermDocumentMatrix()` or `DocumentTermMatrix()` (the first presents terms as rows and documents as columns).

```{r eval=FALSE, include=FALSE}
dtm <- DocumentTermMatrix(my_vctr_corpus)
inspect(dtm)
```

Here is a quick analysis of the Russian Tweets data loaded at the beginning of the chapter.

```{r  eval=FALSE, include=FALSE}
tweets_dtm <- VCorpus(VectorSource(tweets$content))
tweets_dtm[[1]]$meta
inspect(tweets_dtm[[1]])  # first "document"

# Add the following and followers cols as metadata
meta(tweets_dtm, 'following') <- tweets$following
meta(tweets_dtm, 'followers') <- tweets$followers
head(meta(tweets_dtm))
```



# DROP
These notes use the [Russian Troll Tweets](https://www.kaggle.com/vikasg/russian-troll-tweets) data set for examples of analyzing a single document of text with many entries, and the `janeaustenr::austen_books()` data set to analyze a corpus of documents.

```{r}
# library(janeaustenr)
# library(gutenbergr)
# library(scales)
# library(wordcloud)
# library(igraph)
# library(ggraph)
# library(widyr)
# library(tm)
# library(gridExtra)
```


```{r cache=TRUE, message=FALSE, warning=FALSE, eval=FALSE, include=FALSE}
tweets_url <- "https://github.com/fivethirtyeight/russian-troll-tweets/raw/master/IRAhandle_tweets_1.csv"
tweets <- read_csv(url(tweets_url)) %>%
  head(n = 5000)
glimpse(tweets)

austen <- janeaustenr::austen_books()
```

