# Survival Analysis

These notes rely on the [Survival Analysis in R](https://campus.datacamp.com/courses/survival-analysis-in-r) DataCamp course, [STHDA](http://www.sthda.com/english/wiki/survival-analysis), [BJC](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2394262/), and Applied Survival Analysis Using R [@Moore2016].

Survival analysis models time to event. It differs from linear regression in two respects. Event times are typically skewed right with many early events and few late ones. This violates linear regression's normality assumption. Time-to-event outcomes also incur censoring: an unknown starting event (left censoring) and/or ending event (right censoring). Censuring occurs if the event has not yet happened or another event prevents further follow-up.

A typical survival analysis uses Kaplan-Meier (KM) plots to visualize survival curves, log-rank tests to compare survival curves among groups, and Cox proportional hazards regression to describe the effect of explanatory variables on survival. In R, use the **survival** package for modeling and the **survminer** package for visualization.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(survival)
library(survminer)
```

## Basic Concepts

Let $T^*$ be a random variable representing the time to failure, and $U$ be a random variable representing the time to a sensoring event (right-sensoring). The study observes whichever comes first, $T = \mathrm{min}(T^*, U)$. The data set will look something like this one, where `time` is $T$ and `status` is $\delta = I[T^* < U]$ (1 = not-failed, 2 = failed).

```{r}
survival::lung %>% head()
```


Sensoring sometimes occurs when subjects are monitored for a fixed period of time (Type I), a pre-specified level of failures is reached, halting the study (Type II), or the subject drops out for a reason other than failure (random censoring).

You can specify the survival distribution function either as a *survival function* or as a *hazard function*. Let $F(t) = Pr(T \le t)$ be the cumulative risk function, the probability of failing on or before time $t$. The survival function, $S(t)$, is its complement. 

$$S(t) = 1 - F(t) = P(T > t), \hspace{3mm} 0 < t < \infty.$$

The hazard function is the instantaneous failure rate at $t$ given survival up to $t$,

$$h(t) = \lim_{\delta \rightarrow 0}{\frac{P(t < T < t + \delta|T > t)}{\delta}}.$$

The functions are related by the probability multiplication rule, $P(AB) = P(A|B)P(B)$. The probability of failure at $t$, $f(t) = F'(t)$, is the probability of failure at $t$ given survival up to $t$ multiplied by the probability of surviving up to $t$.

$$f(t) = h(t) S(t)$$

$S(t)$ is also the negative exponent of the cumulative hazard function,

$$S(t) = e^{-H(t)}.$$

Use the survival function to estimate the mean survival time, $E(T) = \int S(t)dt$, and median survival time, $S(t) = 0.5$.

Take the exponential distribution as a quick example. It has a constant hazard, $h(t) = \lambda$. The cumulative hazard is $H(t) = \int_0^t \lambda du = \lambda t$. The survival function is $S(t) = e^{-\lambda t}$. The probability of failure at time $t$ is $f(t) = \lambda e^{-\lambda t}$. The expected time to failure is $E(t) = \int_0^\infty e^{-\lambda t} dt = 1 / \lambda$, and the median time to failure is $S(t) = e^{-\lambda t} = .5$, or $t_{med} = \log(2) / \lambda$.

## Survival Curve Estimation

There are parametric and non-parametric methods to estimate a survival curve.  The usual non-parametric method is the *Kaplan-Meier* estimator.  The usual parametric method is the *Weibull* distribution. In between is the most common way to estimate a survivor curve, the *Cox proportional hazards* model.

### Kaplan-Meier

The Kaplan-Meier (KM) estimator for the survival function is the product over failure times of the conditional probabilities of surviving to the next failure time.

$$\hat{S} = \prod_{i: t_i < t}{\frac{n_i - d_i}{n_i}}$$

where $n_i$ is the number of subjects at risk at time $i$ and $d_i$ is the number failing. The KM curve falls only when a subject fails, not when a subject is censored. Confidence limits are calculated using the "delta" method to obtain the variance of $\log \left(\hat{S}(t) \right)$.

Calculate $\hat{S}$ with `survival::survfit()`. The `lung` data set contains the `status` (1 = censored, 2 = dead) of 228 advanced lung cancer patients. `survfit()` operates on a `Surv` object, created by `Surv()`. The event indicator should normally be a numeric 0/1 coding, but for historical reasons 1/2 works too as is used here.^[See neat discussion in Note section of `Surv()` help file.] `?Surv()` explains that it treats factor event variables as multiple endpoint data, so leave it as numeric. It's okay to define explanatory variables as factors though.

```{r}
lung_1 <- survival::lung %>%
  mutate(sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")))

lung_1 %>% head()
```

`survfit()` creates survival curves from a formula or from a previously fitted Cox model. If the CI crosses zero, you can cap it, or use the complimentary log-log transformation parameter `conf.type = "log-log"`. This one doesn't need it.

```{r}
# km_fit <- survfit(Surv(time, status) ~ sex, data = lung, conf.type = "log-log")
km_fit <- survfit(Surv(time, status) ~ sex, data = lung_1)
km_fit
```

112 of 138 males and 53 of 90 females had a failure event (died). The median survival is `r km_fit %>% summary %>% pluck("table") %>% .["sex=Male", "median"]` days for males and `r km_fit %>% summary %>% pluck("table") %>% .["sex=Female", "median"]` days for females. The 95% CI is around the median time to event. Retrieve these values from the `summary()$table` object.

`broom::tidy()` returns the number at risk and number of failures and censors at each event time. At time 11, 3 of the 138 males at risk died, so survival at time 11 is *S*(11) = 1 - 3/138 = .978. At time 12, 1 of the 135 that remained died, so *S*(12) = *S*(11) $\cdot$ (1/135) = .971, and so on.

```{r}
km_fit %>% broom::tidy()
```

**survminer** does a good job plotting KM models. Vertical drops indicate failure events and vertical ticks indicate censoring events.

```{r warning=FALSE}
ggsurvplot(
  km_fit,
  fun = "pct",
  linetype = "strata", # Change line type by groups
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  fontsize = 3, # used in risk table
  surv.median.line = "hv", # median horizontal and vertical ref lines
  ggtheme = theme_light(),
  palette = c("#E7B800", "#2E9FDF"),
  title = "Kaplan-Meier Survival Function Estimate"
)
```

`ggsurvplot()` can also plot the cumulative risk function, $F(t) = 1 - S(t)$, with parameter `fun = "event"`, and the cumulative hazard function, $H(t) = -\log(S(t))$, with parameter `fun = "cumhaz"`.

<!-- The colons (::::, :::) create a style html within document -->
:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r echo=FALSE}
ggsurvplot(
  km_fit,
  fun = "event",
  linetype = "strata", # Change line type by groups
  pval = TRUE,
  conf.int = TRUE,
  ggtheme = theme_bw(),
  palette = c("#E7B800", "#2E9FDF"),
  title = "Cumulative Risk"
)
```
:::

::: {}
```{r echo=FALSE}
ggsurvplot(
  km_fit,
  fun = "cumhaz",
  linetype = "strata", # Change line type by groups
  pval = TRUE,
  conf.int = TRUE,
  ggtheme = theme_bw(),
  palette = c("#E7B800", "#2E9FDF"),
  title = "Cumulative Hazard"
)
```
:::

::::

### Log-Rank Test

It's not immediately obvious how to compare two survival distributions, since they can cross, diverge, etc. When observations do not follow a defined parametric distribution, use a non-parametric test. In this case, the log-rank test is the solution.

For each event time, you could construct a 2x2 contingency table between failure/non-failure and curves 0/1. Holding the margins as fixed, then the probability of observing $d_{0i}$ failures in curve 0 at time $i$ follows a [hypergeometric](https://bookdown.org/mpfoley1973/probability/hypergeometric.html) distribution.

|              |Control           |Treatment         |Total         |
|--------------|------------------|------------------|--------------|
|Failures      |$d_{0i}$          |$d_{1i}$          |$d_i$         |
|non-Failures  |$n_{0i} - d_{0i}$ |$n_{1i} - d_{1i}$ |$n_i - d_i$   |
|Total         |$n_{0i}$          |$n_{1i}$          |$n_i$         |

$$f(d_{0i} | n_{0i}, n_{1i}, d_i) = \frac{{{n_{0i}}\choose{d_{0i}}}{{n_{1i}}\choose{d_{1i}}}}{{n_i}\choose{d_i}}$$

The expected value is $e_{0i} = E(d_{0i}) = \frac{d_i}{n_i} \cdot n_{0i}$ with variance $v_{0i} = Var(d_{0i}) = d_{i} \cdot \frac{n_{0i}}{n_i}  \cdot \frac{n_{1i}}{n_i} \cdot \frac{n_i - d_i}{n_i - 1}$. 

The log-rank test statistic is the sum of the differences between the observed expected failures, $U_0 = \sum (d_{0i} - e_{0i})$, normalized by dividing by the square-root of its variance, $V_0 = Var({U_0}) = \sum v_{0i}$.

$$U = \frac{U_0}{\sqrt{V_0}} \sim N(0, 1)$$

or equivalently, its square, a chi-square random variable with one degree of freedom.

$$U^2 = \frac{U_0^2}{V_0} \sim \chi_1^2$$

```{r}
survdiff(Surv(time, status) ~ sex, data = lung_1)
```

The chi-sq test statistic is 10.3 on one d.f., for a p-value of 0.001, so yes, males and females had different survival patterns.

### Parametric Models

Kaplan-Meier curves and logrank tests are examples of univariate analysis. They describe survival as a function of a single categorical factor variable. Weibull and other parametric models describe the function of *multiple* covariates.

Parametric models are less common than KM and Cox because they are less flexible. But the process follows the parametric distribution, these models are preferable because they are easier to work with. They have fixed parameters and you use standard likelihood theory for parameter estimation.

#### Exponential

The exponential distribution is the easiest to use because it has a time-independent hazard function.

$$\begin{eqnarray}
\log h(t) &=& \alpha + \beta X \\
h(t) &=& e^{\left(\alpha + \beta X \right)} \\
&=& \lambda
\end{eqnarray}$$

Interpret $\alpha$ as the baseline log-hazard because $\log h(t) = \alpha$, or $h(t) = e^\alpha$, when $X$ is zero. The cumulative hazard is $H(t) = \int_0^t \lambda dt = \lambda t$ and the corresponding survival function is 

$$S(t) = e^{-H(t)} = e^{-\lambda t}.$$

The expected survival time is $E(T) = \int_0^\infty S(t)dt = \int_0^\infty e^{-\lambda t} dt = 1 / \lambda.$. The median survival time is $S(t) = e^{-\lambda t} = 0.5$, or $t_{med} = \log(2) / \lambda$.

Given a data set for a process that follows an exponential distribution, the survival curve is fit using *maximimum likelihood estimation* (MLE). My notes [here](https://bookdown.org/mpfoley1973/statistics/maximum-likelihood-estimation.html) explain MLE for the exponential distribution. The survival version is a little more complicated because of censoring. The likelihood that $\lambda$ produces the observed outcomes is the product of the probability densities for each observation because they are a sequence of independent variables. Let $\delta_i = 1$ for unsensored observations and $0$ for censored. 

$$
L(\lambda; t_1, t_2, \dots, t_n) = \Pi_{i=1}^n f(t_i; \lambda)^{\delta_i} S(t_i; \lambda)^{1-\delta_i}
$$

Substituting $f(t) = h(t) S(t)$, and then substituting $h(t) = \lambda$ and $S(t) = e^{-\lambda t}$ and simplifying,

$$\begin{eqnarray}
L(\lambda; t_1, t_2, \dots, t_n) &=& \Pi_{i=1}^n h(t_i; \lambda)^{\delta_i} S(t_i; \lambda) \\
&=& \Pi_{i=1}^n \lambda^{\delta_1} e^{-\lambda t_i} \\
&=& \lambda^{\sum \delta_i} \exp \left(-\lambda \sum_{i=1}^n t_i \right)
\end{eqnarray}$$

Simplify the notation by letting $d = \sum \delta_i$, the total number of failures, and $V = \sum t_i$, the number of person-years (or days or whatever).

$$L(\lambda; t_1, t_2, \dots, t_n) = \lambda^d e^{-\lambda V}$$

That is difficult to optimize, but the log of it is simple.

$$l(\lambda; t_1, t_2, \dots, t_n) = d \ln(\lambda) - \lambda V$$

Maximize the log-likelihood equation by setting its derivative to zero and solving for $\lambda$.

$$\begin{eqnarray}
\frac{d}{d \lambda} l(\lambda; t_1, t_2, \dots, t_n) &=& \frac{d}{d \lambda} \left(d \ln(\lambda) - \lambda V \right) \\
0 &=& \frac{d}{\lambda} - V \\
\lambda &=& \frac{d}{V}
\end{eqnarray}$$

$\lambda$ is the reciprocal of the sample mean, person-years divided by failures.

The second derivative, $-\frac{d}{\lambda^2}$, is approximately the negative of the variance of $\lambda$.

$$V(\lambda) = d / V^2$$

#### Weibull

Although the exponential function is convenient, the [Weibull distribution](https://bookdown.org/mpfoley1973/probability/weibull.html) is more appropriate for modeling lifetimes. Its hazard function is 

$$\begin{eqnarray}
h(t) &=& \alpha \lambda (\lambda t)^{\alpha - 1} \\
&=& \alpha \lambda^\alpha t^{\alpha-1}
\end{eqnarray}$$  

The cumulative hazard function is $H(t) = (\lambda t)^\alpha$ and the corresponding survival function is

$$S(t) = e^{-(\lambda t)^\alpha}.$$

The exponential distribution is a special case of the Weibull where $\alpha = 1$. The expected survival time is $E(t) = \frac{\Gamma (1 + 1 / \alpha)}{\lambda}$. The median survival time is $t_{med} = \frac{[\log(2)]^{1 / \alpha}}{\lambda}$.

Whereas the Kaplan-Meier estimate produces a step function, the Weibull model produces a smooth survival curve. Fit Weibull with `survreg()` (KM is fit with `survfit()`).

```{r}
dat <- lung %>% mutate(sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")))
wb_fit <- survreg(Surv(time, status) ~ sex, data = dat)
summary(wb_fit) 
```

```{r}
dat <- tibble(
  sex = c(rep("Male", 99), rep("Female", 99)),
  s = rep(1 - seq(.01, .99, by = .01), 2),
  pred = map2(sex, s, ~predict(wb_fit, type = "quantile", p = 1 - .y, se = TRUE, 
                               newdata = data.frame(sex = .x))),
  t = map_dbl(pred, ~pluck(.x, "fit")),
  se = map_dbl(pred, ~pluck(.x, "se.fit")),
  ucl = t + 1.96*se,
  lcl = t - 1.96*se
  ) 
  
ggplot() +
  geom_line(data = dat, aes(x = t, color = sex, y = s)) +
  geom_line(data = dat, aes(x = ucl, color = sex, y = s), linetype = 2) +
  geom_line(data = dat, aes(x = lcl, color = sex, y = s), linetype = 2)
```


Use `predict()` to get survival expectations. 


```{r collapse=TRUE}
# 90% of subjects fail by time 
wb_fit %>% predict(type = "quantile", p = .9, newdata = data.frame(sex = factor(c(1, 2), levels = "Male", "Female")))
```


```{r}

data(GBSG2, package = "TH.data")
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# 90% of patients survive beyond time point 385
# Alternatively, 10% of patients die at time 385
predict(wb, type = "quantile", p = 1 - 0.9, newdata = data.frame(1))
# The median survival time is 1694
predict(wb, type = "quantile", p = 1 - 0.5, newdata = data.frame(1))

surv <- seq(.99, .01, by = -.01)
t <- predict(wb, type = "quantile", p = 1 - surv, newdata = data.frame(1))
head(data.frame(time = t, surv = surv))

surv_wb <- data.frame(time = t, surv = surv, 
                      upper = NA, lower = NA, std.err = NA)
ggsurvplot_df(fit = surv_wb, surv.geom = geom_line)
```

Fit a Weibull model controlling for hormonal therapy `horTh` and tumor size `tsize`.

```{r}
wbmod <- survreg(Surv(time, cens) ~ horTh + tsize, data = GBSG2)

coef(wbmod)
summary(wbmod)

surv <- seq(.99, .01, by = -.01)
newdata <- expand.grid(
    horTh = levels(GBSG2$horTh),
    tsize = quantile(GBSG2$tsize, probs = c(0.25, 0.50, 0.75))
)
t <- predict(wbmod, type = "quantile", p = 1 - surv, newdata = newdata)

surv_wbmod <- surv_wbmod_wide <- cbind(newdata, t) %>%
    pivot_longer(names_to = "surv_id", values_to = "time", cols = -c(1:2)) %>%
    mutate(tsize = as.numeric(tsize),
           surv_id = as.factor(as.numeric(surv_id))) %>%
    data.frame()
surv_wbmod$surv = surv[as.numeric(surv_wbmod$surv_id)]
surv_wbmod$upper = NA
surv_wbmod$lower = NA
surv_wbmod$std.err = NA
surv_wbmod$strata = NA
surv_wbmod[, c("upper", "lower", "std.err", "strata")] <- NA
ggsurvplot_df(surv_wbmod, surv.geom = geom_line,
  linetype = "horTh", color = "tsize", legend.title = NULL)
```

Interpret the coefficient as the probability of surviving falls by 0.012 per unit increase in the tumor size and increases by 0.312 if taking hormonal therapy.

You can fit other models with the `dist = c("lognormal", "exponential")` parameter.


### Cox

Whereas parametric models specify the baseline hazard function, $\alpha(t) = \log h_0(t)$, the Cox proportional hazards model is semi-parametric in that it leaves it unspecified.  

$$\log h_i(t) = \alpha(t) + \beta X_i$$

or

$$h_i(t) = h_0(t) \cdot e^{X_i \beta} = h_0(t) \psi_i$$

The $e^{\beta_j}$ values are called the hazard ratios.  A positive $e^{\beta_j}$ means the hazard increases with the covariate. The proportionality of the model comes from the lack of time dependence in the $X$ variables.  Two ratio of the hazard functions of two individuals is 

$$\frac{h_i(t)}{h_{i'}(t)} = \frac{h_0(t) \cdot e^{X_i \beta}}{h_0(t) \cdot e^{X_{i'} \beta}}$$

Remarkably, even though the baseline hazard is unspecified, the Cox model can still be estimated by the method of partial likelihood. Consider the first failure time, $t_1$.  The probability that patient $i$ is the one to fail is the proportion of patient $i$'s hazard divided by the sum of the hazards of all $R_1$ patients at risk,

$$p_1 = \frac{h_i(t_1)}{\sum_{k \in R_1} h_k(t_1)} = \frac{h_0(t_1) \psi_i}{\sum_{k \in R_1} h_0(t_1) \psi_k} = \frac{\psi_i}{\sum_{k \in R_1} \psi_k}$$

The next failure event has a reduced $R_2$ patients at risk.  The partial likelihood for $D$ failure times is the product $L_i = p_1 p_2 \cdots p_D$. The Cox model uses maximum partial likelihood estimation to find the value of $\phi$ that maximizes the likelihood function.

Fit a Cox proportional hazards model with `coxph()`.  Here is a simple Cox proportional hazards model for a single covariate, `sex`.

```{r}
lung_cox <- coxph(Surv(time, status) ~ sex, data = lung)
summary(lung_cox)
```

The Wald statistic (z) is the ratio of each regression coefficient to its standard error (`z = coef/se(coef)`), just as with linear regression.

The negative coefficient estimator sign means that the hazard (risk of death) decreases with increasing values of the variable. `sex` is encoded as a 1 = male, 2 = female, so sex = -0.5310 means females have a lower risk of death than males.

The exponentiated coefficients (`exp(coef)`) are the hazard ratios, the effect-size of the covariates. Being female (sex=2) reduces the hazard by a factor of 0.5880 (41%).

Below the coefficients estimators table is a table of hazard ratio Confidence intervals. 

The last section of the summary object is three tests for the overall significance of the model: the likelihood-ratio test, Wald test, and score logrank statistics. These three methods are asymptotically equivalent. The Likelihood ratio test has better behavior for small sample sizes, so it is generally preferred.

A multivariate analysis works the same way.  Here is the Cox model with two additional covariates: `age` and `ph.ecog`.

```{r}
lung_cox_2 <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung)
summary(lung_cox_2)
```

The p-values for all three overall tests (likelihood, Wald, and score) are significant, indicating that the model is significant (not all $\beta$ values are 0). 

Having fit a Cox model to the data, it’s possible to visualize the predicted survival proportion at any given time for a particular risk group. Function `survfit()` estimates the survival proportion, by default at the mean values of covariates.

```{r}
#ggsurvplot(survfit(lung_cox_2), data = lung, palette = "#2E9FDF", ggtheme = theme_minimal())
```

To display the effects of one or more particular covariates, construct a data frame with test cases and pass to `survfit()` with the `newdata` argument.

```{r}
newdata <- expand.grid(
    sex = unique(lung$sex),
#    age = quantile(lung$age, probs = c(0.25, 0.50, 0.75)),
    age = median(lung$age),
    ph.ecog = 1
)
rownames(newdata) <- c("male", "female")

# Create survival curves.  The rownames show up in the model
lung_pred <- survfit(lung_cox_2, newdata = newdata, data = lung)

# surv_summary() creates the data.frame with a summary of the survfit() results, including columns like time (survival time) and surv (survival probability).
lung_pred0 <- surv_summary(lung_pred)
# get the corresponding new_data cols
lung_pred1 <- cbind(lung_pred0, newdata[as.character(lung_pred0$strata), ])
ggsurvplot_df(
  lung_pred1, 
  color = "sex", 
  legend.labs = c("M", "F"),
  legend.title = "Sex", 
  conf.int = TRUE, 
  ggtheme = theme_minimal())
```

