# Case Study {#study}

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(survival)
library(survminer)
library(gtsummary)
library(scales)
```

The guidelines for reporting the Kaplan-Meier test are from Laerd's Kaplan-Meier using SPSS Statistics [@Laerd2015]. The data is from `survival::lung`. 

```{r}
d_lung <- survival::lung %>%
  filter(!is.na(ph.ecog)) %>%
  mutate(
    ph.ecog = factor(
      ph.ecog,
      levels = c(0, 1, 2, 3, 4),
      labels = c("Asymptomatic", "Ambulatory", "In bed <50%", "In bed >50%", "Bedridden")),
    ph.ecog = fct_drop(ph.ecog),
    ph.ecog = fct_lump_n(ph.ecog, n = 2, other_level = "Bedridden"),
    patient_id = row_number()
  )
```

A study investigated differences in all-cause mortality between men and women diagnosed with advanced lung cancer. `r nrow(d_lung) %>% comma(1)` participants aged `r min(d_lung$age)` to `r max(d_lung$age)` were monitored up to three years until time of death. The participants were segmented into three groups according to their ECOG performance score: *asymptomatic*, symptomatic but completely *ambulatory*, and *bedridden* at least part of the day. Participant age was captured as a controlling covariate.

```{r}
(t1 <- d_lung %>%
  mutate(status = factor(status, levels = c(1, 2), labels = c("censored", "died"))) %>%
  tbl_summary(by = "ph.ecog", include = c(time, status, ph.ecog, age)) %>%
  add_overall())
```

## Cox

The Cox proportional hazards model describes the effect of explanatory variables on survival and can include controlling variables. Add `age` as a controlling variable.

```{r}
cox_fit <- coxph(Surv(time, status) ~ ph.ecog + age, data = d_lung)
(cox_tbl <- cox_fit %>% gtsummary::tbl_regression(exponentiate = TRUE))
```

Using a Cox proportional hazards regression analysis, we find the association between ECOG and survival time statistically significant with a Bedridden estimate of `r gtsummary::inline_text(cox_tbl, variable = ph.ecog, level = "Bedridden")` relative to Asymptomatic. The hazard ratio for Ambulatory was not statistically significant, `r gtsummary::inline_text(cox_tbl, variable = ph.ecog, level = "Ambulatory")`.
 
```{r}
survfit(cox_fit, newdata = list(ph.ecog = levels(d_lung$ph.ecog),
                                             age = rep(median(d_lung$age), 3)), data = d_lung) %>%
  surv_summary() %>%
  ggsurvplot_df(
    fun = "pct",
    # linetype = "strata", # Change line type by groups
    # conf.int = TRUE,
    surv.median.line = "hv", # median horizontal and vertical ref lines
    ggtheme = theme_light(),
    palette = c("goldenrod", "sienna", "tomato"),
    title = "Fitted Cox model at Median Participant Age.",
    legend.title = "",
    legend.labs = levels(d_lung$ph.ecog)
  )
```

## Discrete Time

To fit a discrete time model, first create a discrete time data set, on record per relevant time period. You could do this by hand with `expand.grid()`

```{r eval=FALSE}
cs_glm_dat <- expand.grid(
  time = seq(min(d_lung$time), max(d_lung$time), 1),
  patient_id = unique(d_lung$patient_id)
) %>%
  inner_join(d_lung, by = "patient_id") %>%
  filter(time.x <= time.y) %>%
  mutate(status = if_else(time.x < time.y, 0, status - 1)) %>%
  rename(time = time.x)
```

But `discSurv::dataLong` does this for you.

```{r}
cs_glm_dat <- d_lung %>% 
  mutate(status = status - 1) %>%
  discSurv::dataLong("time", "status", timeAsFactor = FALSE)
```

For example, patient 1 died at time 306, so they have 306 rows in the new data set, with `y = 0` for all but the last row.

```{r}
cs_glm_dat %>% filter(obj == 1) %>% tail()
```

`glm()` fits either the `log(-log(1-hazard))` or `log(hazard / (1-hazard))` as a function of `time`, controlling for `ph.ecog` and `age`. The two links are similar here^[Full discussion on [Rens van de Schoot](https://www.rensvandeschoot.com/tutorials/discrete-time-survival/).].

```{r}
cs_glm_dat %>%
  group_by(ph.ecog, timeInt) %>%
  summarise(.groups = "drop", died = sum(y), at_risk = n()) %>%
  mutate(
    hazard = died / at_risk,
    logit_hazard = log(hazard / (1 - hazard)),
    cloglog_hazard = log(-log(1 - hazard))
  ) %>%
  pivot_longer(cols = c(logit_hazard, cloglog_hazard)) %>%
  filter(!hazard %in% c(0, 1)) %>%
  ggplot(aes(x = timeInt, y = value, color = ph.ecog)) +
  geom_point() +
  geom_smooth(method = "lm", formula = "y ~ x") +
  facet_wrap(facets = vars(name)) +
  theme_light()
```

```{r}
logit_fit <- glm(
  y ~ timeInt + ph.ecog + age, 
  data = cs_glm_dat, 
  family = binomial(link = "logit")
)
(logit_tbl <- logit_fit %>% gtsummary::tbl_regression(exponentiate = TRUE))
```

Using a discrete time logistic regression analysis with logit link function, we find the association between ECOG and survival time statistically significant with a Bedridden estimate of `r gtsummary::inline_text(logit_tbl, variable = ph.ecog, level = "Bedridden")` relative to Asymptomatic. The hazard odds ratio for Ambulatory was not statistically significant, `r gtsummary::inline_text(logit_tbl, variable = ph.ecog, level = "Ambulatory")`.

```{r}
cloglog_fit <- glm(
  y ~ timeInt + ph.ecog + age, 
  data = cs_glm_dat, 
  family = binomial(link = "cloglog")
)
(cloglog_tbl <- cloglog_fit %>% gtsummary::tbl_regression(exponentiate = TRUE))
```

Using a discrete time logistic regression analysis with complementary log-log link function, we find the association between ECOG and survival time statistically significant with a Bedridden estimate of `r gtsummary::inline_text(cloglog_tbl, variable = ph.ecog, level = "Bedridden")` relative to Asymptomatic. The hazard ratio for Ambulatory was not statistically significant, `r gtsummary::inline_text(cloglog_tbl, variable = ph.ecog, level = "Ambulatory")`.

## Reporting

The reporting below uses the Cox model rather than the discrete time model.

> Participants were randomly assigned to three different interventions in order to quit smoking: a hypnotherapy programme (n = 50), wearing nicotine patches (n = 50) and using e-cigarettes (n = 50). Kaplan-Meier survival analysis (Kaplan & Meier, 1958) was conducted to compare the three different interventions for their effectiveness in preventing smoking resumption. A similar percentage of censored cases was present in the hypnotherapy (12.0%), nicotine patch (14.0%) and e-cigarette (14.0%) intervention groups and the pattern of censoring was similar. Participants that underwent the hypnotherapy programme had a median time to smoking resumption of 69.0 (95% CI, 45.2 to 92.8) days. This was longer than the groups receiving nicotine patches or e-cigarettes, which had identical median times to smoking resumption of 9.0 (95% CI, 6.6 to 11.4) days and 9.0 (95% CI, 7.1 to 10.9) days, respectively. A log rank test was conducted to determine if there were differences in the survival distributions for the different types of intervention. The survival distributions for the three interventions were statistically significantly different, χ2(2) = 25.818, p < .0005. Pairwise log rank comparisons were conducted to determine which intervention groups had different survival distributions. A Bonferroni correction was made with statistical significance accepted at the p < .0167 level. There was a statistically significant difference in survival distributions for the hypnotherapy vs nicotine patch intervention, χ2(1) = 11.035, p = .001, and hypnotherapy vs e-cigarette intervention, χ2(1) = 29.003, p < .0005. However, the survival distributions for the e-cigarette and nicotine patch interventions were not statistically significantly different, χ2(1) = 1.541, p = .214.
