```{r}
library(tidyverse)
```

# Coding Practices

## Efficient Coding

This section explains how to make code run faster.^[These notes are from the DataCamp course [Writing Efficient R Code](https://campus.datacamp.com/courses/writing-efficient-r-code/).]

The most basic practice in efficient coding is to keep your R, RStudio, and package versions up to date. Check you version from the `version` global list object.

```{r}
version
```
Update R from inside the R GUI.

```{r eval=FALSE}
installr::updateR()
```

Update RStudio from Help > Check for Updates. RStudio closes for the update. Once updated, RStudio should default to using the new version of R too.

Update your packages from the RStudio's Packages panel.

### Benchmarking

Benchmarking is the capture of the performance time for comparison to alternative solutions. Benchmark a section of code by wrapping it within a function and calling the function with `system.time()`. Ignore the user and system times - they are components of the overall elapsed time.

```{r}
my_f <- function(n) {
  for(i in 1:n) { x <- runif(1) }
}
system.time(my_f(1e4))
```

The `benchmark()` function in the **microbenchmark** package does this, but also compares functions, runs them multiple times, and calculates summary statistics.

```{r}
library(microbenchmark)
microbenchmark(my_f(1e3), my_f(1e4), my_f(1e5), times = 10)
```

The `benchmark_io()` function in the **benchmarkme** package reads and writes a file and compares your performance to other users.

```{r collapse=TRUE}
library(benchmarkme)

# read/write a 5MB file
my_io <- benchmark_io(runs = 1, size = 5)
plot(my_io)
```

You can also use the package to retrieve hardware data.

```{r collapse=TRUE}
get_ram()
get_cpu()
```

### Profiling

Code profiling is taking time snapshots at intervals throughout the code in order to find the location of bottlenecks. Base R function `Rprof()` does this, but it is not user friendly. Instead, use `profvis()` from the **profvis** package.

```{r}
library(profvis)
profvis({
  for(i in 2:3) { 
    my_f(10^i)
  }
  my_f(1e4)
  my_f(1e5)
})
```

### Parallel Programming

Use the **parallel** package to parallelize your code. Parallelization adds communication overhead among the cpus, so it's not always helpful. 

```{r collapse=TRUE}
library("parallel")
mat <- as.matrix(mtcars)

# make a cluster using all cores, or maybe all but one
n_cores <- benchmarkme::get_cpu() %>% pluck("no_of_cores") - 1
cl <- makeCluster(n_cores)

# make copies of data and functions for each cluster
clusterExport(cl, "my_f")
system.time(my_f(1e5))

# use a parallel version of a function, like parApply instead of apply.
# In this case, the serial version is faster!
microbenchmark(apply(mat, 1, median), 
               parApply(cl, mat, 1, median), 
               times = 100)
# stop the cluster
stopCluster(cl)
```

### Other Efficiency Tips

Tip #1: Don't allocate memory on the fly.

```{r collapse=TRUE}
# bad
fun_bad <- function(n) {
  x <- NULL 
  for(i in 1:n) { x <- c(x, rnorm(1)) }
}

# good
fun_good <- function(n) {
  x <- numeric(n)
  for(i in 1:length(x)) { x[i] <- rnorm(1) }
}

microbenchmark(fun_bad(1000), fun_good(1000), times = 10)
```

Tip #2: Use a vectorized solution whenever possible.

```{r eval=FALSE}
# makes 100 calls to rnorm() and makes 100 assignments to x
x <- numeric(100)
for(i in 1:length(x)) { x[i] <- rnorm(1) }

# makes 1 call to rnorm() and 1 assignment to x
x <- rnorm(100)
```

Tip #3: Use a matrix instead of a dataframe if possible. Matrix operations are fast because with predefined dimensions, accessing any row, col, or cell is a multiple of a dimension length.

```{r collapse=TRUE}
# matrix is faster for column selection...
mat <- mtcars %>% as.matrix()
df <- mtcars
microbenchmark(mat[, 1], df[, 1])

# and even faster for row selection (because of variable data types.
microbenchmark(mat[1, ], df[1, ])
```

Tip #4: Use `&&` for smarter logical testing - if condition 1 is FALSE, then R will not evaluate condition 2. `&&` only works for single logical values - not vectors.

```{r collapse=TRUE}
slwr <- function() { 
  for(i in 1:10000) {
    x <- rnorm(1); 
    if(x > .4 & x < .6) {y <- x}
  }
}
fstr <- function() { 
  for(i in 1:10000) {
    x <- rnorm(1); 
    if(x > .4 && x < .6) {y <- x}
  }
}
microbenchmark(slwr, fstr, times = 10)
```

## Defensive Coding