--- 
title: "Coding in R"
author: "Michael Foley"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Handbook of basic R coding concepts and practices."
---
--- 
title: "Coding in R"
author: "Michael Foley"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Handbook of basic R coding concepts and practices."
---

# Overview {-}

This is a handbook of basic R concepts and best practices.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<!--chapter:end:index.Rmd-->

# Data Manipulation {#data_manipulation}

## Tidyverse



## Data Manipulation

### Stringr

## Dates and Times


<!--chapter:end:01-data_manipulation.Rmd-->

# Web Scraping

Here is a review of existing methods.

<!--chapter:end:02-web_scraping.Rmd-->


# Coding Practices

Placeholder


## Efficient Coding
### Benchmarking
### Profiling
### Parallel Programming
### Other Efficiency Tips
## Defensive Coding

<!--chapter:end:03-coding_practices.Rmd-->

# Functions and Packages

```{r include=FALSE}
library(tidyverse)
```


## Functions

Functions are verbs, so name them that way (e.g., `run_linear_regression()` instead of `lm()`).

### Arguments

Functions have two types of arguments: data and detail. Start with the data argument so your function is compatible with the `%>%` operator. Name your detail arguments thoughtfully, but name the data argument `x`.

Specify a discrete list of parameter values with a character vector argument, then check the setting with `match.arg()`.

```{r collapse=TRUE, error=FALSE}
my_fun <- function(alternative = c("a", "b", "c")) {
  alternative = match.arg(alternative)
  print(alternative)
}

my_fun("b")

# if unspecified, `match.arg()` returns the first option.
my_fun()

# if misspecified, `match.arg()` returns an error.
 my_fun("d")
# Error in match.arg(alternative) : 'arg' should be one of “a”, “b”, “c”
```

If there is a default behavior that is too complicated to show conveniently, default it to `NULL` and explain the behavior the documentation.

```{r eval=FALSE}
# complicated default
my_fun <- function(x, labs = c(c("lab 1", "middle lab", "lab 3"),
                                  c("lbl 1", "middle lbl", "lbl 3"))) {
  labs = match.arg(labs)
  print(labs)
}
my_fun()

# better
my_fun <- function(x, labs = NULL) {
  if(is.null(labs)) {
    labs = c("lab 1", "middle lab", "lab 3")
  }
  print(labs)
}

my_fun()
```

If your code is a wrapper around other code, you do not need to specify all of the parameters that the sub-functions might use. Instead, allow for parameters with `...` and let them flow downward into the sub-functions. The downside here is that the user needs to consult the documentation of the interior functions to know what to pass.

```{r collapse=TRUE}
my_fun <- function(x, ...) {
  mean(x, ...)
}

c(1, 2, NA_real_, .5, 4) %>% my_fun(na.rm = TRUE)
```

Another problem with `...` is that it gets passed down to all functions, and that can cause an error.
```{r eval=FALSE}
my_fun <- function(x, ...) {
  mean(x, ...)
  str_detect(x, "chester", ...)
}

c(1, 2, NA_real_, .5, 4) %>% my_fun(na.rm = TRUE)
# returns
# Error in str_detect(x, "chester", ...) : unused argument (na.rm = TRUE)
```

Check arguments (assert) at the beginning of the function.

```{r}
my_fun <- function(x, na.rm = FALSE) {
  if(!is.numeric(x)) {
    stop("x is not class numeric; it's class is '", class(x), "`.")
  }
}
# my_fun(c("chester", "allen", "arthur")) 
# returns
# Error in my_fun(c("chester", "allen", "arthur")) : x is not class numeric; it's class is 'character`.
```

A better way to handle assertions with the **assertive** package.

```{r}
my_fun <- function(x, na.rm = FALSE) {
  # x must be a positive number
  assertive::assert_is_numeric(x)
  assertive::assert_all_are_positive(x)
  log(x)
}
# c(2, 3, -1, 4) %>% my_fun()
```

Exit a function early with `result()`.
```{r}
is_leap_year <- function(year) {
  if(year %% 400 == 0) {
    return(TRUE)
  }
  if(year %% 100 == 0) {
    return(FALSE)
  }  
  if(year %% 4 == 0) {
    return(TRUE)
  }
  FALSE
}
```

If the main purpose of a function is to print output, you may want to return something else silently. Do that with `invisible()`. 

R functions can only return a single value. Two ways around this are to return a list and storing objects as attributes `attr()`.

### Return Values and Scope

Objects exist within nested *environments* in R. Environments are like lists with inheritance. Your default environment is "R_GlobalEnv". 

```{r}
environmentName(env <- environment())
while(environmentName(env) != "R_EmptyEnv") {
  print(environmentName(env <- parent.env(env)))
}
```

When you define a function, it creates a new environment with the existing environment.

```{r}
x <- 1
y <- 2

# my_fun creates an environment within .GlobalEnv
my_fun <- function() {
  # If R cannot find an object in the current environment, it checks the parent.
  x <- x + y
}
my_fun()
# but not the other way around - x only changed its value in the sub-environ!
print(x)

# if you want to change x from .GlobalEnv, use the super-assignment operator
my_fun <- function() {
  x <<- x + y
}
my_fun()
print(x)
```

You can override it by declaring an environment from a list with `list2Env()`. 

```{r}
# You can create a sub-environment
my_env <- list2env(list(a_to_z = letters, numbers = 0:10))

exists("x")
exists("x", envir = my_env)
exists("x", envir = my_env, inherits = FALSE)
```

## Packages

<!--chapter:end:04-functions_and_packages.Rmd-->


# Object Oriented Programming

Placeholder


## S3
## R6
### Inheritance

<!--chapter:end:05-object_oriented_programming.Rmd-->

`r if (knitr::is_html_output()) '
# References {-}
'`

<!--chapter:end:06-references.Rmd-->

