# Inference

```{r include=FALSE}
library(tidyverse)
library(glue)
library(pwr)

set.seed(123456)
```

Statistical inference is the use of a sample's distribution to describe the population distribution. Hypothesis tests, confidence intervals, and effect size estimates are all examples of statistical inference. 

We wary of published study results. Identical studies might produce significant and non-significant results, yet only the significant result is likely to reach publication (publication bias). The researcher may have tortured the data until they found a statistically significant result. The study might suffer from low statistical power. Applying principles of inference can mitigate these problems.

There are at least three approaches to establishing statistical inference: frequentist, likelihood, and Bayesian. Think of them philosophically. The frequentist approach is the path of *action*. It rejects a null hypothesis if the *p*-value is low because repeated sample analyses are likely to agree. The likelihood apprach is the path of *knowledge*. It compares the observed summary measure to the likelihoods of the other possible realities. The Bayesian approach is the path of *belief*. It uses a summary measure to update the prior belief.

## *P*-Values

*P*-values express how surprising the summary measure is given the null hypothesis (H0).

Suppose you hypothesize that IQs have increased from the established mean of $\mu_0$ = 100. H0 is $\mu$ = 100 and the alternative hypothesis, H1, is $\mu$ > 100. Also suppose you are *right* and the population mean IQ is actually 106. Suppose the "population" is some very big number - 1,000,000 for convenience. 

```{r}
mu_0 <- 100
mu <- 106
sigma <- 15
N <- 1000000

pop_100 <- data.frame(person = seq(1, N), iq = rnorm(N, mu_0, sigma))
pop_106 <- data.frame(person = seq(1, N), iq = rnorm(N, mu, sigma))
```

Here is what the distribution of IQs might look like.

```{r echo=FALSE}
pop_100 %>%
  ggplot(aes(x = iq)) +
  geom_density() +
  theme_light() +
  labs(title = "Presumed population IQs are centered at 100 with SD = 15.")
```

You take a random sample of *n* = 30 IQs from the population. 

```{r collapse=TRUE}
n <- 30

x <- sample(pop_106$iq, n)

(x_bar <- mean(x))
```

You measure $\bar{x}$ = `r scales::number(x_bar, accuracy = 0.1)`, SD = `r scales::number(sd(x), accuracy = 0.1)`. How surprising is this result given H0 that $\mu$ is 100? I.e., what is the *probability* of observing $\bar{x}$ this extreme? According to the Central Limit Theorem (CLM) repeated samples of size *n* from a large population will yield $\bar{x}$ values that approach a normal distribution centered at $\mu$ with a standard deviation equal to the population SD ($\sigma$) divided by $sqrt{n}$. The standard deviation of the sampling distribution of the mean is commonly referred to as the *standard error* (SE). For a sample size of *n* = 30, and a population SD of 15, you'd expect repeated samples to converge on a mean of 100 with SE = 2.7. You can verify this empirically. 

```{r}
sim <- replicate(1000, mean(sample(pop_100$iq, 30)))
```

Here is the distribution of $\bar{x}$ measurements from 1,000 random samples from the hypothesized population.

```{r echo=FALSE}
x_quantiles <- quantile(sim, probs = c(.025, .975))

data.frame(iq = sim) %>%
  ggplot(aes(x = iq)) +
  geom_density() +
  geom_vline(xintercept = x_quantiles["2.5%"], linetype = 2, color = "firebrick") +
  geom_vline(xintercept = x_quantiles["97.5%"], linetype = 2, color = "firebrick") +
  theme_light() +
  labs(
    title = glue(
      "Presumed population IQs are centered at {scales::number(x_bar, accuracy = 0.1)} ",
      "with SD = {scales::number(sd(x), accuracy = 0.1)}"),
    caption = "95% of values captured within dashed lines."
    )
```

So what is the probability of measuring $\bar{x}$ = `r scales::number(x_bar, accuracy = 1)`? It is the probability of measuring a value $\frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}}$ = `r scales::number((x_bar - mu_0) / (sigma / sqrt(n)), accuracy = .1)` standard deviations from 100. You can look up the probability from the normal distribution. 

```{r collapse=TRUE}
q <- (x_bar - mu_0) / (sigma / sqrt(n))

pnorm(q, lower.tail = FALSE)
```

You can get a sense of what this probability means by overlaying the sampling distribution of the mean from the presumed (H0) distribution with the sampling distribution of the mean from the actual population. 

```{r echo=FALSE, warning=FALSE}
iq_range <- seq(90, 120, .01)

tmp <- data.frame(
  iq = iq_range,
  H0 = dnorm(iq_range, mean = mu_0,  sd = sigma / sqrt(n)),
  H1 = dnorm(iq_range, mean = x_bar, sd = sigma / sqrt(n))
)

alpha_x <- qnorm(.95, mu_0, sigma / sqrt(n))
alpha_y <- dnorm(alpha_x, mean = mu_0, sd = sigma / sqrt(n))

plot_alpha_beta <- function(x) {
  x %>% 
    pivot_longer(cols = c(H0, H1), values_to = "density") %>%
    mutate(alpha_beta = if_else((name == "H0" & iq >= alpha_x) | 
                                  (name == "H1" & iq <= alpha_x), 
                                density, NA_real_)) %>%
    ggplot(aes(x = iq, color = name, fill = name)) +
    geom_line(aes(y = density), size = 1) +
    geom_area(aes(y = alpha_beta), alpha = .4, show.legend = FALSE) +
    geom_vline(xintercept = x_bar, linetype = 2, size = .5, color = "steelblue") +
    scale_color_manual(values = c("H0" = "lightgoldenrod", "H1" = "lightsteelblue")) +
    scale_fill_manual(values = c("H0" = "lightgoldenrod", "H1" = "lightsteelblue")) +
    annotate("text", x = mu_0, y = .075, label = "1 - alpha", parse = TRUE) +
    annotate("text", x = x_bar, y = .075, label = "1 - beta", parse = TRUE) +
    annotate("text", x = x_bar - 4, y = .0075, label = "beta", parse = TRUE) +
    annotate("text", x = x_bar - .5, y = .0075, label = "alpha", parse = TRUE) +
    theme_light() +
    theme(panel.grid = element_blank(), legend.position = "top") +
    labs(
      title = glue("Measured x-bar of {scales::number(x_bar, accuracy = .1)} is ",
                   "outside than the .05 confidence levl of the H0 \n",
                   "value of {mu_0}."),
      fill = NULL, color = NULL, x = "IQ"
    )
}

tmp %>% plot_alpha_beta()
```

## Type I and II Errors

Either H0 or H1 is correct, and you must choose to either reject or non reject H0. That means there are four possible states at the end of your study. If your summary measure is extreme enough for you to declare a "positive" result and reject H0, you are either correct (true positive) or incorrect (false positive). False positives are called *Type I errors*. Alternatively, if your summary measure is *not* extreme enough for you to reject H0, you are either correct (true negative) or incorrect (false negative). False negatives are called *Type II errors*.

The probabilities of landing in these four states depend on your chosen significance level, $\alpha$, and on the statistical power of the study, 1 - $\beta$.

|                        |H0 True                       |H1 True                      |
|------------------------|------------------------------|-----------------------------|
|Significance test is positive, so you reject H0.| False Positive <br>Type I Error<br> Probability = $\alpha$    | True Positive <br>Good Call! <br> Probability = 1 - $\beta$ |
|Significance test is negative, so you do *not* reject H0.   | True Negative <br> Good Call!<br> Probability = ($1 - \alpha$)    | False Negative <br>Type II Error <br>Probability = $\beta$ |

$\alpha$ is the expected rate of Type I errors due to summary measures that are extreme by chance alone. $\beta$ is the expected rate of Type II errors due to summary measures that are not extreme by chance alone. In the IQ example, if the reality is $\mu$ = `r mu_0`, any sample mean over `r scales::number(qnorm(.95, mu_0, sigma / sqrt(n)), accuracy = .1)` would be a false positive under the $\alpha$ = .05 level of significance. if the reality is $\mu$ = `r mu`, any sample mean under `r scales::number(qnorm(.95, mu_0, sigma / sqrt(n)), accuracy = .1)` would be a false negative under the $1 - \beta$ = `r scales::number(pwr.t.test(n = n, d = (x_bar - mu_0) / sigma, sig.level = 0.05, type = "one.sample", alternative="greater")$power, accuracy = .01)` statistical power of the study.

## Statistical Power

Type II error rates ($\beta$) vary inversely with the power of the study (1 - $\beta$). Statistical power is an increasing function of a) sample size, b) effect size, and c) significance level. The positive association with significance level means there is a trade-off between Type I and Type II error rates. A small $\alpha$ sets a high bar for rejecting H0, but you run the risk of failing to appreciate a *real* difference. On the other hand, a small $\alpha$ sets a low bar for rejecting H0, and you run the risk of judging a random difference as a real difference.

The 1 - $\beta$ statistical power threshold is usually set at .80, similar to the $\alpha$ = .05 level of significance threshold convention. Given a real effect, a study with a statistical power of .80 will only find a positive test result 80% of the time. There may be such a thing as *too* much power, however. With a large enough sample size, even trivial effect sizes may yield a positive test result. You need to consider both sides of this coin.

A *power analysis* is frequently used to determine the sample size required to detect a threshold effect size given an $\alpha$ level of significance criteria. A power analysis expresses the relationship among four components. If you know any three, it can return the fourth. The components are a) 1 - $\beta$ (power), b) sample size, c) $\alpha$, and d) expected effect size.

Suppose you set 1 - $\beta$ at .8 and $\alpha$ at .05. You can use the power test to see the relationship between sample size and effect size. Let's do that with the IQ example.

```{r}
data.frame(n = 10:300) %>%
  mutate(d = map_dbl(n, ~ pwr.t.test(n = ., 
                                     sig.level = .05, 
                                     power = .80, 
                                     type = "one.sample", 
                                     alternative = "greater") %>% pluck("d")),
         delta = d * sigma
  ) %>%
  ggplot(aes(x = n, y = delta)) +
  geom_line() +
  geom_vline(xintercept = 30, linetype = 2, size = 1, color = "slateblue") +
  theme_light() +
  labs(title = "")

```


## An Exercise in p-values

What distribution of p-values would you expect if there is a true effect and you repeated the study many times? What if there is not true effect? The answer is completely determined by the statistical power of the study (the probability of observing true positives).

Let's run 100,000 simulations of an experiment where you sample 26 individuals and measure their IQ. We simulate the sample by generating 26 random values from the normal distribution centered at 106 with a standard deviation of 15. We will test whether the sample averages differ from the hypothesized population average of 100.

```{r}
library(tidyverse)

n_sims <- 1000 #number of simulated experiments

run_sim <- function(M = 106, n = 26, SD = 15) {
  sim <- data.frame(i = 1:n_sims) %>%
    mutate(
      x = map(i, ~ rnorm(n = n, mean = M, sd = SD)),
      z = map(x, ~ t.test(., mu = 100)),
      p = map_dbl(z, ~ .x$p.value)
    )
  sim
}

plot_sim <- function(x, M = 106, n = 26, SD = 15, n_breaks = 20, x_lim = c(0, 1), 
                     x_labels = waiver(), y_lim = c(0, n_sim)) {
  test_power = sum(x$p < .05) / n_sims
  x %>% 
    ggplot(aes(x = p)) +
    geom_bar(fill = "lightgoldenrod") + 
    scale_x_binned(n.breaks = n_breaks, limits = x_lim, labels = x_labels) +
    geom_hline(yintercept = n_sims * .05, color = "firebrick", linetype = 2) +
    annotate("text", x = .8, y = n_sims * .05, label = "5%") +
    theme_light() +
    scale_y_continuous(labels = scales::comma_format(), limits = c(0, n_sims)) +
    labs(
      title = glue(
        "Measured p-values in {n_sims} simulations of samples of size {n} distributed \n",
        "N({M}, {SD}^2) with H0 = 100. Power of the test is {scales::percent(test_power)}."
      )
    )
}

#With a mean difference of 6, and SD of 15, and a sample size of 26, the test has 50% power)
# The power of the study the percent of the times it correctly identifies a
# positive result. 
# study_power <- sum(sim$p < .05) / n_sims  

# This matches the formal calculation.
# study_power_thry <- pwr.t.test(
#   d = (M - 100) / SD,
#   n = n,
#   sig.level = 0.05,
#   type = "one.sample",
#   alternative="two.sided"
# )
```
Since the statistical power is the probability of observing a statistically significant 
result, if there is a true effect, we can also see the power in the figure itself. We can calculate the difference between p-values above 0.5 minus the p-values below 0.05, and divide this number by the number of simulations.

```{r}
# sim <- run_sim(M = 106, n = 26, SD = 15)
# 
# sim %>% plot_sim()
# 
# sim %>% filter(p<.05) %>% plot_sim(M = 106, n = 26, SD = 15, n_breaks = 100, x_lim = c(0, .05), waiver())
```

Change the sample size in line 10 from n<-26 to n<-51. Run the simulation by selecting 
all lines and pressing CTRL+Enter. What is the power in the simulation now that we have 
increased the sample size from 26 people to 51 people?

```{r}
sim_51 <- run_sim(M = 106, n = 51, SD = 15)

plot_sim(sim_51)
```

The power of the test increased to `r scales::percent(sum(sim_51$p < .05) / n_sims)`. The *p*-values are much steep er than with 50% power.

What would happen if there were no real difference in IQ scores? The *p*-value distribution is essentialy flat.

```{r}
# sim_null <- run_sim(M = 100, n = 26, SD = 15)
# 
# plot_sim(sim_null)
```

The leftmost bar on the plot (p<.05) are the Type I errors (false positives).

When there is no true effect, *p*-values are uniformly distributed under 
the null. Every *p*-value is equally likely when the null hypothesis is true, and every bar in 
the graph will contain 5% of all the p-values (as indicated by the dotted red line). When there is no true effect, p-values are uniformly distributed. When there is a true effect, the p-value distribution depends on the power, and the higher the power, the more p-values fall below 0.05, and the steeper the p-value distribution 
becomes.


Let’s take a look at just the p-values below 0.05. It's not the case that all p-values > 0.05 are support for the null-hypothesis, and all p-values below 0.05 are support for the alternative hypothesis. Re-run the simulation (still with M<-100), but focused on just the p values between 0 and .05.

```{r}
# sim_null %>% plot_sim(.001, x_lim = c(-.05, .051))
```

We see the same 
uniform distribution, but now every bar contains 1% of the p-values, so the p-value 
distribution is very flat and almost impossible to see (we will zoom in on the y-axis later
this assignment). The red line now clearly gives the frequency for each bar, assuming the 
null hypothesis is true.

## Further Reading

Pritha Bhandari has two nice posts on [Type I and Type II errors](https://www.scribbr.com/statistics/type-i-and-type-ii-errors/) and [Statistical Power](https://www.scribbr.com/statistics/statistical-power/). Daniel Lakens's Coursera class [Improving your statistical inferences](https://www.coursera.org/learn/statistical-inferences/home/welcome) has a great *p*-value simulation exercise in Week 1 (assignment)