# Group Differences

```{r include=FALSE}
library(tidyverse)
library(gtsummary)
library(foreign)
library(scales)
library(janitor)
library(flextable)
library(broom)
```

This section presents a statistical tests of _comparison_. Which test to use depends on the structure of the data. Below is a guide to which test to use.

**Continuous (interval or ratio) and Ordinal Outcomes**

- The **Independent samples _t_-test** (\@ref(ttest)) is the main way to compare a continuous dependent variable between the two levels of a binomial independent categorical variable. Revert to the nonparametric **Wilcoxon Rank Sum Test** (\@ref(wilcoxonranksum)) if the _t_-test assumptions fail. A special case arises when samples are paired. Paired samples are more like one-sample tests where the dependent variable is the _difference_ between the pairs. Use the **Paired Samples _t_-test** (\@ref(pairedttest)) or the nonparametric **Wilcoxon signed-rank test** (\@ref(wilcoxonsignedrank)).
- If the independent categorical variable is multinomial, conduct an **ANOVA** (\@ref(onewayanova)) test or the nonparametric **Kruskal-Wallis test** (\@ref(kw)).

**Discrete (count) Outcomes**

- The **Chi-square test of homogeneity** (\@ref(chisqhomogeneity)) is the main way to compare a discrete dependent variable among the levels of a binomial or multinomial independent categorical variable. Revert to the nonparametric **Fisher's Exact Test** (\@ref(fisherexact)) if the sample size is small. Handle the special case of paired samples with the **Pairwise Prop Test** (\@ref(pairwiseproptest)) or the nonparametric **McNemar's test** (\@ref(mcnemar)).

## Independent Samples t-Test {#ttest}

If a population measure *X* is normally distributed with mean $\mu_X$ and variance $\sigma_X^2$, and a population measure *Y* is normally distributed with mean $\mu_Y$ and variance $\sigma_Y^2$, then their difference is normally distributed with mean $d = \mu_X - \mu_Y$ and variance $\sigma_{XY}^2 = \sigma_X^2 + \sigma_Y^2$. By the CLT, as the sample sizes grow, a non-normally distributed *X* and *Y* will approach normality, and so will their difference.

The **independent samples t-test** evaluates an hypothesized difference, $d_0$ (H0: $d = d_0$), from the difference in sample means $\hat{d} = \bar{x} - \bar{y}$, or constructs a (1 - $\alpha$)% confidence interval around $\hat{d}$ to estimate $d$ within a margin of error, $\epsilon$.

In principal, you can evaluate $\hat{d}$ with either a *z*-test or a *t*-test. Both require independent samples and approximately normal sampling distributions. Sampling distributions are normal if the underlying populations are normally distributed, or if the sample sizes are large ($n_X$ and $n_Y$ $\ge$ 30). However, the *z*-test additionally requires known sampling distribution variances, $\sigma^2_X$ and $\sigma^2_Y$. These variances are never known, so always use the *t*-test.

The *z*-test assumes $d$ is normally distributed around $\hat{d} = d$ with standard error $SE = \sqrt{\frac{\sigma_X^2}{n_X} + \frac{\sigma_Y^2}{n_Y}}.$ The test statistic for H0: $d = d_0$ is $Z = \frac{\hat{d} - d_0}{SE}$. The (1 - $\alpha$)% CI is $d = \hat{d} \pm z_{(1 - \alpha {/} 2)} SE$.  

The *t*-test assumes $d$ has a *t*-distribution around $\hat{d} = d$ with standard error $SE = \sqrt{\frac{s_X^2}{n_X} + \frac{s_Y^2}{n_Y}}.$ The test statistic for H0: $d = d_0$ is $T = \frac{\hat{d} - d_0}{SE}$. The (1 - $\alpha$)% CI iss $d = \hat{d} \pm t_{(1 - \alpha / 2), (n_X + n_Y - 2)} SE$.

There is a complication with the *t*-test *SE* and degrees of freedom. If the sample sizes are small and the standard deviations from each population are similar (the ratios of $s_X$ and $s_Y$ are <2), pool the variances, $s_p^2 = \frac{(n_X - 1) s_X^2 + (n_Y-1) s_Y^2}{n_X + n_Y-2}$, so that $SE = s_p \sqrt{\frac{1}{n_X} + \frac{1}{n_Y}}$ and the degrees of freedom (df) = $n_X + n_Y - 2$ (the **pooled variances t-test**). Otherwise, $SE = \sqrt{\frac{s_X^2}{n_X} + \frac{s_Y^2}{n_Y}}$, but you reduce df using the Welch-Satterthwaite correction, $df = \frac{\left(\frac{s_X^2}{n_X} + \frac{s_Y^2}{n_Y}\right)^2}{\frac{s_X^4}{n_X^2\left(N_X-1\right)} + \frac{s_Y^4}{n_Y^2\left(N_Y-1\right)}}$ (the **separate variance t-test**, or **Welch's t-test**).

## Wilcoxon Rank Sum Test {#wilcoxonranksum}

The **Wilcoxon rank sum test**^[The Mann-Whitney U test is also called the Mann-Whitney U test, Wilcoxon-Mann-Whitney test, and the two-sample Wilcoxon test] is a nonparametric alternative to the independent-samples *t*-test. Use the the test when the samples are not normally distributed or when the response variables are ordinal rather continuous. In the first case where the normality assumption fails, the test evaluates H0 that the two samples are from the same population distribution. In the second case where the response variables are ordinal, the test evaluates the difference in medians. 

The Wilcoxon Rank Sum test ranks the response values, then sums the ranks for the reference group, $W = \sum R_1$. The test statistic is $U = W - \frac{n_2(n_2 + 1)}{2}$ where $n_2$ is the number of observations in the test group. $U$ will equal 0 if there is complete separation between the groups, and $n_1 n_2$ if there is complete overlap. Reject H0 if $U$ is sufficiently small.

## Case Study 1 {-}

```{r include=FALSE}
ind_num <- list()

ind_num$t_dat <- read.spss("./input/independent-samples-t-test.sav", to.data.frame = TRUE)
ind_num$t_n <- with(ind_num$t_dat, by(engagement, gender, length))
ind_num$t_mean <- with(ind_num$t_dat, by(engagement, gender, mean))
ind_num$t_sd <- with(ind_num$t_dat, by(engagement, gender, sd))

ind_num$mw_dat <- read.spss("./input/mann-whitney-test.sav", to.data.frame = TRUE)
ind_num$mw_n <- with(ind_num$mw_dat, by(engagement, gender, length))
ind_num$mw_median <- with(ind_num$mw_dat, by(engagement, gender, median))
ind_num$mw_sd <- with(ind_num$mw_dat, by(engagement, gender, sd))
```

A company shows an advertisement to $n_M$ = `r ind_num$t_n["Male"]` males and $n_F$ = `r ind_num$t_n["Female"]` females, then measures their engagement with a survey. Do the groups' mean engagement scores differ?

[Laerd](https://statistics.laerd.com/) has two data sets for this example. One meets the conditions for a *t*-test, and the other fails the normality test, forcing you to use the Mann-Whitney U test.

```{r fig.height=3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `t-Test` = ind_num$t_dat, 
  `Mann-Whitney` = ind_num$mw_dat, 
  .id = "set"
) %>%
  ggplot(aes(x = engagement, fill = gender)) +
  # geom_density(alpha = 0.6) +
  geom_histogram(bins = 20, alpha = 0.6, position = "dodge") +
  facet_wrap(~fct_rev(set)) +
  labs(title = "Mean Engagement Scores Counts", fill = NULL) +
  theme_light() +
  scale_fill_manual(values = c("Male" = "slategray", "Female" = "snow3"))
```

```{r fig.height=3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `t-test` = ind_num$t_dat,
  `Mann-Whitney` = ind_num$mw_dat,
  .id = "set"
) %>%
  group_by(set, gender) %>%
  summarize(.groups = "drop",
            mean_engage = mean(engagement),
            n = n(),
            se_engage = sd(engagement) / sqrt(n),
            ci_low = mean_engage - qt(.975, n-1) * se_engage,
            ci_high = mean_engage + qt(.975, n-1) * se_engage) %>%
  ggplot(aes(x = gender)) +
  geom_col(aes(y = mean_engage), fill = "snow3", color = "snow4", width = 0.25) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), color = "snow4", width = .125) +
  theme_minimal() +
  facet_wrap(~fct_rev(set)) +
  labs(x = NULL, y = NULL,
       title = "Mean Engagement Score", 
       subtitle = "Two data sets with similar means.",
       caption = "Error bars show 95% CI.")
```

The *t*-test data set has the following summary statistics.

```{r}
(ind_num$t_gt <- ind_num$t_dat %>% 
  gtsummary::tbl_summary(
    by = c(gender), 
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ))
```

> There were `r ind_num$t_n["Male"]` male and `r ind_num$t_n["Female"]` female participants. Data are mean $\pm$ standard deviation, unless otherwise stated. The advertisement was more engaging to male viewers, `r gtsummary::inline_text(ind_num$t_gt, engagement, column = "Male")`, than female viewers, `r gtsummary::inline_text(ind_num$t_gt, engagement, column = "Female")`.

The Mann-Whitney data set has the following summary statistics.

```{r}
(ind_num$mw_gt <- ind_num$mw_dat %>% 
  gtsummary::tbl_summary(
    by = c(gender), 
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ))
```

> There were `r ind_num$mw_n["Male"]` male and `r ind_num$mw_n["Female"]` female participants. Data are mean $\pm$ standard deviation, unless otherwise stated. The advertisement was more engaging to male viewers, `r gtsummary::inline_text(ind_num$mw_gt, engagement, column = "Male")`, than female viewers, `r gtsummary::inline_text(ind_num$mw_gt, engagement, column = "Female")`.

#### Conditions {-}

The independent samples *t*-test and Mann-Whitney U test apply when 1) the response variable is continuous, 2) the independent variable is binomial, and 3) the observations are independent. The decision between the *t*-test and Mann-Whitney stems from two additional conditions related to the data distribution - if both conditions hold, use the *t*-test; otherwise use Mann-Whitney.

1. **Outliers**. There should be no outliers in either group. Outliers exert a large influence on the mean and standard deviation. Test with a box plot. If there are outliers, you might be able to drop them or transform the data.
2. **Normality**.  Values should be *nearly* normally distributed. The *t*-test is robust to normality, but this condition is important with small sample sizes. Test with Q-Q plots or the Shapiro-Wilk test for normality. If the data is very non-normal, you might be able to transform the data.

If the data passes the two conditions, use the *t*-test, but now you need to check a third condition related to the variances to determine which flavor of the *t*-test to use.

3. **Homogeneous Variances**. Use *pooled-variances* if the variances are homogeneous; otherwise use the *separate variances* method. Test with Levene's test of equality of variances.

If the data does not pass the first two conditions, use Mann-Whitney, but now you need to check a third condition here as well. The condition does not affect how to perform the test, but rather how to interpret the results.

3. **Distribution shape**. If the distributions have the same shape, interpret the Mann-Whitney result as a comparison of the *medians*; otherwise interpret the result as a comparison of the *mean ranks*.

##### Checking for Outliers {-}

Assess outliers with a box plot. Box plot whiskers extend up to 1.5\*IQR from the upper and lower hinges and outliers (beyond the whiskers) are are plotted individually.

```{r echo=FALSE, fig.height=2.5, fig.width=7.5}
bind_rows(
  `t-test` = ind_num$t_dat,
  `Mann-Whitney` = ind_num$mw_dat,
  .id = "set"
) %>%
  ggplot(aes(x = gender, y = engagement)) +
  geom_boxplot(fill = "snow3", color = "snow4", alpha = 0.6, width = 0.25, 
               outlier.color = "goldenrod", outlier.size = 2) +
  theme_minimal() +
  facet_wrap(~fct_rev(set)) +
  labs(title = "Boxplot of Engagement Score", y = "Score", x = NULL)
```

For the *t* test data set,

>There were no outliers in the data, as assessed by inspection of a boxplot.

and for the Mann-Whitney data set,

>There was one outlier in the data, as assessed by inspection of a boxplot.

If the outliers are data entry errors or measurement errors, fix or discard them. If the outliers are genuine, you have a couple options before reverting to the Mann-Whitney U test.

* Leave it in if it doesn't affect the conclusion (compared to taking it out).
* Transform the variable. Don't do this unless the variable is also non-normal. Transformation also has the downside of making interpretation more difficult.

##### Checking for Normality {-}

Assume the population is normally distributed if *n* $\ge$ 30. Otherwise, assess a Q-Q plot, skewness and kurtosis values, or a histogram. If you still don't feel confident about normality, run a Shapiro-Wilk test.

There are only $n_M$ = `r ind_num$t_n["Male"]` male and $n_F$ = `r ind_num$t_n["Female"]` female observations, so you need to test normality. The QQ plot indicates normality in the t-test data set, but not in the Mann-Whitney data set.

```{r fig.height=3.5, fig.width=7.5}
bind_rows(
  `t-test` = ind_num$t_dat,
  `Mann-Whitney` = ind_num$mw_dat,
  .id = "set"
) %>%
  ggplot(aes(sample = engagement, group = gender, color = fct_rev(gender))) +
  stat_qq() +
  stat_qq_line(col = "goldenrod") +
  theme_minimal() + theme(legend.position = "top") +
  facet_wrap(~fct_rev(set)) +
  labs(title = "Normal Q-Q Plot", color = NULL)
```

Run Shapiro-Wilk separately for the males and for the females. Since we are looking at two data sets in tandem, there are four tests below. For the t-test data set, 

```{r}
(ind_num$t_shapiro <- split(ind_num$t_dat, ind_num$t_dat$gender) %>% 
  map(~shapiro.test(.$engagement))
)
```

>Engagement scores for each level of gender were normally distributed, as assessed by Shapiro-Wilk's test (*p* > .05).

For the Mann-Whitney data set,

```{r}
(ind_num$mw_shapiro <- split(ind_num$mw_dat, ind_num$mw_dat$gender) %>% 
  map(~shapiro.test(.$engagement))
)
```

>Engagement scores for each level of gender were not normally distributed for the Female sample, as assessed by Shapiro-Wilk's test (*p* = `r ind_num$mw_shapiro$Female$p.value %>% scales::comma(accuracy = 0.001)`).

If the data is not normally distributed, you still have a couple options before reverting to the Mann-Whitney U test.

* Transform the dependent variable.
* Carry on regardless - the independent samples *t*-test is fairly robust to deviations from normality.

##### Checking for Homogenous Variances {-}

If the data passed the outliers and normality tests, you will use the *t*-test, so now you need to test the variances to see which version (*pooled-variances* method if variances are homogeneous; *separate variances* if variances are heterogeneous). A rule of thumb is that homogeneous variances have a ratio of standard deviations between 0.5 and 2.0:

```{r}
sd(ind_num$t_dat %>% filter(gender == "Male") %>% pull(engagement)) /
  sd(ind_num$t_dat %>% filter(gender == "Female") %>% pull(engagement))
```

You can also use the *F* test to compare the ratio of the sample variances $\hat{r} = s_X^2 / s_Y^2$ to an hypothesized ratio of population variances $r_0 = \sigma_X^2 / \sigma_Y^2 = 1.$

```{r}
var.test(ind_num$t_dat %>% filter(gender == "Female") %>% pull(engagement), 
         ind_num$t_dat %>% filter(gender == "Male") %>% pull(engagement))
```

Bartlett's test is another option.

```{r}
bartlett.test(ind_num$t_dat$engagement, ind_num$t_dat$gender)
```

Levene's test is a third option. Levene's is less sensitive to departures from normality than Bartlett.

```{r}
(ind_num$levene <- with(ind_num$t_dat, 
                        car::leveneTest(engagement, gender, center = "mean"))
)
```

> There was homogeneity of variances for engagement scores for males and females, as assessed by Levene's test for equality of variances (*p* = `r ind_num$levene %>% pluck("Pr(>F)", 1) %>% scales::number(accuracy = 0.001)`).

##### Checking for Similar Distributions {-}

If the data fail either the outliers or the normality test, use the Mann-Whitney test. The Mann-Whitney data set failed both, so the Mann-Whitney test applies. Now you need to test the distributions to determine how to interpret its results. If the distributions are similarly shaped, interpret the Mann-Whitney U test as inferences about differences in medians between the two groups. If the distributions are dissimilar, interpret the test as inferences about the distributions, lower/higher scores and/or mean ranks.

```{r fig.height = 3.5, fig.width=7.5, echo=FALSE}
ind_num$mw_dat %>%
  ggplot(aes(x = engagement, color = fct_rev(gender))) +
  geom_density() +
  theme_minimal() +
  theme(legend.position = "right") +
  labs(title = "Engagement Distribution", color = NULL)
```

> Distributions of the engagement scores for males and females were similar, as assessed by visual inspection.

#### Test {-}

Conduct the *t*-test or the Mann-Whitney U test. 

##### t-Test {-}

The the *t*-test data the variances were equal, so the pooled-variances version applies (`t.test(var.equal = TRUE)`).

```{r}
(ind_num$t_test <- t.test(engagement ~ gender, data = ind_num$t_dat, var.equal = TRUE))
```

> There was a statistically significant difference in mean engagement score between males and females, with males scoring higher than females, `r scales::number(ind_num$t_test$estimate[1] - ind_num$t_test$estimate[2] %>% as.numeric(), accuracy = 0.01)` (95% CI, `r ind_num$t_test$conf.int[1] %>% scales::number(accuracy = 0.01)` to `r ind_num$t_test$conf.int[2] %>% scales::number(accuracy = 0.01)`), t(`r ind_num$t_test$parameter`) = `r ind_num$t_test$statistic %>% scales::number(accuracy = 0.001)`, *p* = `r ind_num$t_test$p.value %>% scales::number(accuracy = 0.001)`.

The effect size, Cohen's *d*, is defined as $d = |M_D| / s$, where $|M_D| = \bar{x} - \bar{y}$, and $s$ is the pooled sample standard deviation, $s_p = \sqrt{\frac{(n_X - 1) s_X^2 + (n_Y-1) s_Y^2}{n_X + n_Y-2}}$. $d <.2$ is considered trivial, $.2 \le d < .5$ small, and $.5 \le d < .8$ large.

```{r message=FALSE}
(d <- effectsize::cohens_d(engagement ~ gender, data = ind_num$t_dat, pooled_sd = TRUE))
```

> There was a large difference in mean engagement score between males and females, Cohen's d = `r scales::number(d$Cohens_d, accuracy = 0.01)` 95% CI [`r scales::number(d$CI_low, accuracy = 0.01)`, `r scales::number(d$CI_high, accuracy = 0.01)`]

Before rejecting the null hypothesis, construct a plot as a sanity check.

```{r echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=7.5}
d_bar <- ind_num$t_mean["Male"] - ind_num$t_mean["Female"]
d_0 <- 0
se <- effectsize::sd_pooled(engagement ~ gender, data = ind_num$t_dat) * 
  sqrt(1/ind_num$t_n["Male"] + 1/ind_num$t_n["Female"])
df <- sum(ind_num$t_n) - 2
lrr <- se * qt(.025, df)
urr <- se * qt(.975, df)
data.frame(d = seq(-1.0, 1.0, by = .01)) %>%
  mutate(t = (d - d_0) / se,
         prob = dt(x = t, df = df),
         lrr = if_else(d < lrr, prob, NA_real_),
         urr = if_else(d > urr, prob, NA_real_)) %>%
  ggplot() +
  geom_line(aes(x = d, y = prob)) +
  geom_area(aes(x = d, y = lrr), fill = "firebrick", alpha = 0.6) +
  geom_area(aes(x = d, y = urr), fill = "firebrick", alpha = 0.6) +
  geom_vline(aes(xintercept = d_0), size = 1) +
  geom_vline(aes(xintercept = d_bar), color = "goldenrod", size = 1, linetype = 2) +
  annotate("text", x = d_bar + .06, y = .05, 
           label = scales::number(d_bar, accuracy = .01), 
           size = 3, color = "goldenrod3") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.y = element_blank()) +
  labs(title = bquote("Two-Tailed t-test"),
       x = "Difference in Group Means (Male - Female)",
       y = "Probability",
       caption = "Shaded area is 95% CI.")
```

##### Wilcoxon Rank Sum test {-}

The reference level for the `gender` variable is males, so the Wilcoxon Rank Sum test statistic is the sum of male ranks minus $n_f(n_f + 1) / 2$ where $n_f$ is the number of females. You can calculate the test statistic by hand.

```{r}
(ind_num$mw_test_manual <- ind_num$mw_dat %>% 
  mutate(R = rank(engagement)) %>%
  group_by(gender) %>%
  summarize(.groups = "drop", n = n(), R = sum(R), meanR = sum(R)/n()) %>%
  pivot_wider(names_from = gender, values_from = c(n, R, meanR)) %>%
  mutate(U = R_Male - n_Female * (n_Female + 1) / 2))
```

Compare the test statistic to the Wilcoxon rank sum distribution with `pwilcox()`.

```{r}
pwilcox(
  q = ind_num$mw_test_manual[1, ]$U - 1, 
  m = ind_num$mw_test_manual[1, ]$n_Male, 
  n = ind_num$mw_test_manual[1, ]$n_Male, 
  lower.tail = FALSE
) * 2
```

There is a function for all this.

```{r}
(ind_num$mw_test <- wilcox.test(
  engagement ~ gender, 
  data = ind_num$mw_dat, 
  exact = TRUE, 
  correct = FALSE,
  conf.int = TRUE))
```

> Median engagement score was not statistically significantly different between males and females, *U* = `r ind_num$mw_test$statistic`, *p* = `r ind_num$mw_test$p.value %>% scales::number(accuracy = .001)`, using an exact sampling distribution for *U*.

Now you are ready to report the results. Here is how you would report the *t* test.

> Data are mean $\pm$ standard deviation, unless otherwise stated. There were `r ind_num$t_n[["Male"]]` male and `r ind_num$t_n[["Female"]]` female participants. An independent-samples t-test was run to determine if there were differences in engagement to an advertisement between males and females. There were no outliers in the data, as assessed by inspection of a boxplot. Engagement scores for each level of gender were normally distributed, as assessed by Shapiro-Wilk's test (*p* > .05), and there was homogeneity of variances, as assessed by Levene's test for equality of variances (*p* = `r ind_num$levene %>% pluck("Pr(>F)", 1) %>% scales::number(accuracy = 0.001)`). The advertisement was more engaging to male viewers (`r scales::number(ind_num$t_mean["Male"], accuracy = .01)` $\pm$ = `r scales::number(ind_num$t_sd["Male"], accuracy = .01)`) than female viewers (`r scales::number(ind_num$t_mean["Female"], accuracy = .01)` $\pm$ = `r scales::number(ind_num$t_sd["Female"], accuracy = .01)`), a statistically significant difference of `r scales::number(ind_num$t_test$estimate[1] - ind_num$t_test$estimate[2] %>% as.numeric(), accuracy = 0.01)` (95% CI, `r ind_num$t_test$conf.int[1] %>% scales::number(accuracy = 0.01)` to `r ind_num$t_test$conf.int[2] %>% scales::number(accuracy = 0.01)`), t(`r ind_num$t_test$parameter`) = `r ind_num$t_test$statistic %>% scales::number(accuracy = 0.001)`, *p* = `r ind_num$t_test$p.value %>% scales::number(accuracy = 0.001)`, *d* = `r d$Cohens_d %>% scales::number(accuracy = 0.01)`.

Here is how you would report the Mann-Whitney U-Test.

> A Mann-Whitney U test was run to determine if there were differences in engagement score between males and females. Distributions of the engagement scores for males and females were similar, as assessed by visual inspection. Median engagement score for males (`r ind_num$mw_median["Male"] %>% as.numeric() %>% scales::number(accuracy = .01)`) and females (`r ind_num$mw_median["Female"] %>% as.numeric() %>% scales::number(accuracy = .01)`) was not statistically significantly different, *U* = `r ind_num$mw_test$statistic`, *p* = `r ind_num$mw_test$p.value %>% scales::number(accuracy = .001)`, using an exact sampling distribution for *U*.

Had the distributions differed, you would report the Mann-Whitney like this:

> A Mann-Whitney U test was run to determine if there were differences in engagement score between males and females. Distributions of the engagement scores for males and females were not similar, as assessed by visual inspection. Engagement scores for males (mean rank = `r ind_num$mw_test_manual %>% pull(meanR_Male)`) and females (mean rank = `r ind_num$mw_test_manual %>% pull(meanR_Female)`) were not statistically significantly different, *U* = `r ind_num$mw_test$statistic`, *p* = `r ind_num$mw_test$p.value %>% scales::number(accuracy = .001)`, using an exact sampling distribution for *U*.

## Paired Samples t-Test {#pairedttest}

There are two common study designs that employ a paired samples t-test to compare two related groups. One relates the groups as two time points for the same subjects. The second relates the groups as two tests of the same subjects, e.g. comparing reaction time under two lighting conditions.

The *paired samples t-test* uses the mean of sampled paired differences $\bar{d}$ as an estimate of the mean of the population paired differences $\delta$ to evaluate an hypothesized mean $\delta_0$.  Test $H_0: \delta = \delta_0$ with test statistic $T = \frac{\bar{d} - \delta_0}{se}$, or define a $(1 - \alpha)\%$ confidence interval as $\delta = \bar{d} \pm t_{1 - \alpha / 2, n - 1} se$.  *The paired t-test is really just a one-sample mean t-test operating on variable that is defined as the difference between two variables*.

The paired samples *t* test applies when the sampling distribution of the mean of the population paired differences is normally distributed and there are no significant outliers.

## Wilcoxon Signed-Rank Test {#wilcoxonsignedrank}

The *Wilcoxon signed-rank test* is a nonparametric alternative to the paired-samples t-test for cases in which the paired differences fails the normality condition, but is at least symmetrically distributed.

The test statistic is the sum product of the difference signs (-1, +1) and the rank of the difference absolute values, $W = \sum_{i=1}^n sign (d_i) \cdot R_i$. The more differences that are of one sign, or of extreme magnitude, the larger $W$ is likely to be, and the more likely to reject $H_0$ of equality of medians.

### Sign Test {-}

The *sign test* is an alternative to the Wilcoxon signed-rank test for cases in which the paired differences fails the symmetrical distribution condition.

The test statistic is the count of pairs whose difference is positive, $W = cnt(d_i > 0)$. $W \sim b(n, 0.5)$, so the sign test is really just an exact binomial test (*exact sign test*), or for large *n*-size, the normal approximation to the binomial (*sign test*).

## Case Study 2 {-}

```{r include=FALSE}
drink <- list()

drink$t_dat <- read.spss("./input/paired-samples-t-test.sav", to.data.frame = TRUE) %>%
  mutate(diff = carb_protein - carb)
drink$t_n <- nrow(drink$t_dat)
drink$t_mean <- mean(drink$t_dat$carb_protein - drink$t_dat$carb)
drink$t_sd <- sd(drink$t_dat$carb_protein - drink$t_dat$carb)

drink$wilcoxon_dat <- read.spss("./input/wilcoxon-signed-rank-test.sav", 
                                to.data.frame = TRUE) %>%
  mutate(diff = carb_protein - carb)
drink$wilcoxon_n <- nrow(drink$wilcoxon_dat)
drink$wilcoxon_median <- list()
drink$wilcoxon_med$carb_protein <- median(drink$wilcoxon_dat$carb_protein)
drink$wilcoxon_med$carb <- median(drink$wilcoxon_dat$carb) 

drink$sign_dat <- read.spss("./input/sign-test.sav", to.data.frame = TRUE) %>%
  mutate(diff = carb_protein - carb)
drink$sign_n <- nrow(drink$sign_dat)
drink$sign_median <- list()
drink$sign_med$carb_protein <- median(drink$sign_dat$carb_protein)
drink$sign_med$carb <- median(drink$sign_dat$carb) 
```

$n$ = `r drink$t_n` athletes consume a carb-only or carb+protein drink prior to running as far as possible in 2 hours and a researcher records their distances under each condition. Do the distances differ from 0?

[Laerd](https://statistics.laerd.com/) has three data sets for this example. One meets the conditions for a t-test. The second fails the normality condition, but is symmetric and meets the conditions for the Wilcoxon test. The third fails the symmetry condition and requires the sign test.

##### t-test data set {-}

```{r}
(drink$t_gt <- drink$t_dat %>% 
  gtsummary::tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})"))
)
```

> There were `r drink$t_n` participants. Data are mean $\pm$ standard deviation, unless otherwise stated. Participants ran further after consuming the carbohydrate-protein drink, `r gtsummary::inline_text(drink$t_gt, carb_protein)` km, than the carbohydrate-only drink, `r gtsummary::inline_text(drink$t_gt, carb)` km.

##### Wilcoxon data set {-}

Once you learn you need Wilcoxon or the sign-test, show the median and IQR summary statistics instead.

```{r}
(drink$wilcoxon_gt <- drink$wilcoxon_dat %>% 
  gtsummary::tbl_summary()
)
```

> There were `r drink$wilcoxon_n` participants. Data are medians and IQR unless otherwise stated. Participants ran further after consuming the carbohydrate-protein drink, `r gtsummary::inline_text(drink$wilcoxon_gt, carb_protein)` km, than the carbohydrate-only drink, `r gtsummary::inline_text(drink$wilcoxon_gt, carb)` km.

##### Sign data set {-}

```{r}
(drink$sign_gt <- drink$sign_dat %>% 
  gtsummary::tbl_summary()
)
```

> There were `r drink$sign_n` participants. Data are median and IQR unless otherwise stated. Participants ran further after consuming the carbohydrate-protein drink, `r gtsummary::inline_text(drink$sign_gt, carb_protein)` km, than the carbohydrate-only drink, `r gtsummary::inline_text(drink$sign_gt, carb)` km.

#### Conditions {-}

The paired samples *t* test applies when the variable is continuous and partitioned into dependent pairs, Additionally, there are two conditions related to the data distribution. If either condition fails, consider the suggested work-around or move to the nonparametric alternatives.

1. **Outliers**. There should be no outliers in the differences because they exert a large influence on the mean and standard deviation. Test with a box plot. If there are outliers, you might be able to drop them if they do not affect the conclusion, or you can transform the data.
2. **Normality**. Differences should be *nearly* normally distributed ("nearly" because the *t*-test is robust to the normality assumption). This condition is especially important with small sample sizes. Test with Q-Q plots or the Shapiro-Wilk test for normality. If the data is very non-normal, you might be able to transform the data.

##### Outliers {-}

Assess outliers with a box plot. Box plot whiskers extend up to 1.5\*IQR from the upper and lower hinges and outliers (beyond the whiskers) are are plotted individually.

```{r echo=FALSE, fig.height=2.5, fig.width=7.5}
bind_rows(
  `t-test` = drink$t_dat,
  `Wilcoxon` = drink$wilcoxon_dat,
  `Sign` = drink$sign_dat,
  .id = "set"
) %>%
  ggplot(aes(x = "difference", y = diff)) +
  geom_boxplot(fill = "snow3", color = "snow4", alpha = 0.6, width = 0.25, 
               outlier.color = "goldenrod", outlier.size = 2) +
  theme_minimal() +
  facet_wrap(~fct_inorder(set)) +
  labs(title = "Boxplot of Difference", y = "Km", x = NULL)
```

>There were no outliers in the data, as assessed by inspection of a boxplot.

Had there been outliers, you might report

> *X* outliers were detected. Inspection of their values did not reveal them to be extreme and they were kept in the analysis.

If the outliers are data entry errors or measurement errors, fix them or discard them. If the outliers are genuine, you can try leaving them in or transforming the data.

##### Normality {-}

Assume the population is normally distributed if *n* $\ge$ 30. These data sets have *n* = `r drink$t_n` observations, so you cannot assume normality. Asses a Q-Q plot, skewness and kurtosis values, histogram, or Shapiro-Wilk test.

```{r fig.height=3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `t-test` = drink$t_dat,
  `Wilcoxon` = drink$wilcoxon_dat,
  `Sign` = drink$sign_dat,
  .id = "set"
) %>%
  ggplot(aes(sample = diff)) +
  stat_qq() +
  stat_qq_line(col = "goldenrod") +
  theme_minimal() + theme(legend.position = "top") +
  facet_wrap(~fct_inorder(set)) +
  labs(title = "Normal Q-Q Plot", color = NULL)
```

For the t-test data set, 

```{r}
(drink$t_shapiro <- shapiro.test(drink$t_dat$diff))
```

> The differences between the distance ran in the carbohydrate-only and carbohydrate-protein trial were normally distributed, as assessed by Shapiro-Wilk's test (*p* = `r drink$t_shapiro$p.value %>% scales::number(accuracy = 0.001)`).

For the Wilcoxon data set,

```{r}
(drink$wilcoxon_shapiro <- shapiro.test(drink$wilcoxon_dat$diff))
```

> The differences between the distance ran in the carbohydrate-only and carbohydrate-protein trial were *not* normally distributed, as assessed by Shapiro-Wilk's test (*p* = `r drink$wilcoxon_shapiro$p.value %>% scales::number(accuracy = 0.001)`).

For the sign-test data set,

```{r}
(drink$sign_shapiro <- shapiro.test(drink$sign_dat$diff))
```

> The differences between the distance ran in the carbohydrate-only and carbohydrate-protein trial were *not* normally distributed, as assessed by Shapiro-Wilk's test (*p* = `r drink$sign_shapiro$p.value %>% scales::number(accuracy = 0.001)`).

If the data is normally distributed, use the t-test. If not, you try transforming the dependent variable, or carrying on regardless since the *t*-test is fairly robust to deviations from normality.

##### Symmetric Distribution {-}

If the data passed the outliers test, but failed the normality test, as the Wilcoxon and sign test data sets above did, you will use the Wilcoxon signed-rank test or sign test. Now you need to test the distribution to determine which test. If the distribution is symmetric, use Wilcoxon; otherwise use the sign test.

```{r fig.height = 3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `Wilcoxon` = drink$wilcoxon_dat,
  `Sign` = drink$sign_dat,
  .id = "set"
) %>%
  mutate(diff_sign = case_when(diff < 0 ~ "-", diff > 0 ~ "+", TRUE ~ "0")) %>%
  ggplot(aes(x = diff, fill = fct_rev(diff_sign))) +
  geom_histogram(bins = 20, color = "snow4") +
  scale_fill_manual(values = list("-" = "firebrick", "+" = "steelblue", "0" = "grey")) +
  theme_bw() +
  theme(legend.position = "top") +
  labs(title = "Paired Samples Differences Distribution", fill = NULL, x = NULL, y = "n") +
  facet_wrap(~fct_rev(set))
```

For the Wilcoxon data set,

> The distribution of the differences between the carbohydrate-protein drink and the carbohydrate-only was symmetric, as assessed by visual inspection.

For the sign data set,

> The distribution of the differences between the carbohydrate-protein drink and the carbohydrate-only was *not* asymmetric, as assessed by visual inspection.

#### Test {-}

##### t-test {-}

```{r}
(drink$t_t <- t.test(x = drink$t_dat$carb_protein, y = drink$t_dat$carb, paired = TRUE)
)
```

> The carbohydrate-protein drink elicited an increase of `r scales::number(drink$t_t$estimate %>% as.numeric(), accuracy = 0.001)` (95% CI, `r drink$t_t$conf.int[1] %>% scales::number(accuracy = 0.001)` to `r drink$t_t$conf.int[2] %>% scales::number(accuracy = 0.001)`) km in the distance run in two hours compared to a carbohydrate-only drink.

The effect size, called Cohen's *d*, is the number of standard deviations the measured mean difference is from the hypothesized difference, $(\bar{d}-d_0) / s$, where $s$ is the sample standard deviation. .2 is small, .5 is medium, and .8 is large. This one is large.

```{r}
(drink$t_d <- effectsize::cohens_d(drink$t_dat$diff))
```

You are about to reject the null hypothesis. Construct a plot as a sanity check on your reasoning.

```{r echo=FALSE, warning=FALSE, fig.height=2.5, fig.width=6.5}
d_0 <- 0
d_bar <- drink$t_mean
se <- drink$t_sd / sqrt(drink$t_n)
ci_lwr <- d_0 - qt(.975, df = drink$t_n - 1) * se
ci_upr <- d_0 + qt(.975, df = drink$t_n - 1) * se
ci_range <- ci_upr - ci_lwr

data.frame(d = seq(ci_lwr - ci_range/2, ci_upr + ci_range/2, length = 100)) %>%
  mutate(t = (d - d_0) / se,
         prob = dt(x = t, df = drink$t_n - 1),
         lrr = if_else(d < ci_lwr, prob, NA_real_),
         urr = if_else(d > ci_upr, prob, NA_real_)) %>%
  ggplot(aes(x = d)) +
  geom_line(aes(y = prob)) +
  geom_area(aes(y = lrr), fill = "firebrick", alpha = 0.6) +
  geom_area(aes(y = urr), fill = "firebrick", alpha = 0.6) +
  geom_vline(aes(xintercept = d_0), size = 1) +
  geom_vline(aes(xintercept = d_bar), color = "goldenrod", size = 1, linetype = 2) +
  annotate("label", x = d_0, y = 0.25,  label = "d[0]", parse = TRUE,
           fill = "grey90", size = 3) +
  annotate("label", x = d_bar, y = 0.15, label = "~bar(d)", parse = TRUE,
           fill = "goldenrod", size = 3) +
  theme_minimal() +
  theme(legend.position="none",         axis.text.y = element_blank()) +
  labs(title = "Two-Tailed t-Test", x = expression(paste(delta)), y = "Probability")
```

Report the results.

> A paired-samples t-test was used to determine the effect of a new formula of sports drink on running performance. Instead of the regular, carbohydrate-only drink, the new sports drink contains a new carbohydrate-protein mixture. Twenty participants were recruited to the study who each performed two trials in which they had to run as far as possible in two hours on a treadmill. In one of the trials they drank the carbohydrate-only drink and in the other trial they drank the carbohydrate-protein drink. The order of the trials was counterbalanced and the distance they ran in both trials was recorded.
>
>Two outliers were detected that were more than 1.5 box-lengths from the edge of the box in a boxplot. Inspection of their values did not reveal them to be extreme and they were kept in the analysis. The assumption of normality was not violated, as assessed by Shapiro-Wilk's test (*p* = `r drink$t_shapiro$p.value %>% scales::number(accuracy = 0.001)`). 
>
>Data are mean $\pm$ standard deviation, unless otherwise stated. Participants ran further after consuming the carbohydrate-protein drink, `r gtsummary::inline_text(drink$t_gt, carb_protein)` km, than the carbohydrate-only drink, `r gtsummary::inline_text(drink$t_gt, carb)` km, a statistically significant increase of `r scales::number(drink$t_t$estimate %>% as.numeric(), accuracy = 0.001)` (95% CI, `r drink$t_t$conf.int[1] %>% scales::number(accuracy = 0.001)` to `r drink$t_t$conf.int[2] %>% scales::number(accuracy = 0.001)`) km, *t*(`r drink$t_t$parameter`) = `r drink$t_t$statistic %>% scales::number(accuracy = 0.001)`, *p* = `r drink$t_t$p.value %>% scales::number(accuracy = .0001)`, *d* = `r drink$t_d$Cohens_d %>% scales::number(accuracy = .01)`.

##### Wilcoxon Signed-Rank Test {-}

From the distribution plot, you can see that most of the signs were positive, and the largest absolute difference values were among the positives, so expect a pretty large test statistic.

```{r}
(drink$wilcoxon_test <- wilcox.test(drink$wilcoxon_dat$carb_protein,
                                    drink$wilcoxon_dat$carb, 
                                    paired = TRUE))
```
> The carbohydrate-protein drink elicited a statistically significant median increase in distance run in two hours compared to the carbohydrate-only drink, *W* = `r drink$wilcoxon_test$statistic %>% number(accuracy = 1)`, *p* = `r drink$wilcoxon_test$p.value %>% scales::number(accuracy = .001)`.

Report the results.

> A Wilcoxon signed-rank test was conducted to determine the effect of a new formula of sports drink on running performance. Instead of the regular, carbohydrate-only drink, the new sports drink contains a new carbohydrate-protein mixture. Twenty participants were recruited to the study who each performed two trials in which they had to run as far as possible in two hours on a treadmill. In one of the trials they drank the carbohydrate-only drink and in the other trial they drank the carbohydrate-protein drink. The order of the trials was counterbalanced and the distance they ran in both trials was recorded.
>
>The difference scores were approximately symmetrically distributed, as assessed by a histogram with superimposed normal curve.
>
>Data are medians unless otherwise stated. Of the 20 participants recruited to the study, the carbohydrate-protein drink elicited an increase in the distance run in 17 participants compared to the carbohydrate-only drink, whereas two participants saw no improvement and one participant did not run as far with the carbohydrate-protein drink. There was a statistically significant median increase in distance run (`r median(drink$sign_dat$diff) %>% scales::number(accuracy = .0001)` km) when subjects imbibed the carbohydrate-protein drink (`r median(drink$sign_dat$carb_protein) %>% scales::number(accuracy = .001)` km) compared to the carbohydrate-only drink (`r median(drink$sign_dat$carb) %>% scales::number(accuracy = .001)` km), *W* = `r drink$wilcoxon_test$statistic %>% number(accuracy = 1)`, *p* = `r drink$wilcoxon_test$p.value %>% scales::number(accuracy = .0001)`.

##### Sign Signed-Rank Test {-}

Conduct the exact sign test since the n-size is not so large that we need the normal approximation to the binomial. Notice `n` is the count of *non-zero* differences.

```{r}
(drink$sign_test <- binom.test(sum(drink$sign_dat$diff > 0), 
                               n = sum(drink$sign_dat$diff != 0)))
```

> The carbohydrate-protein drink elicited a statistically significant median increase in distance run (`r median(drink$sign_dat$diff) %>% scales::number(accuracy = .001)` km) compared to the carbohydrate-only drink, *p* = `r drink$sign_test$p.value %>% scales::number(accuracy = .001)`.

Report the results.

> An exact sign test was conducted to determine the effect of a new formula of sports drink on running performance. Instead of the regular, carbohydrate-only drink, the new sports drink contains a new carbohydrate-protein mixture. Twenty participants were recruited to the study who each performed two trials in which they had to run as far as possible in two hours on a treadmill. In one of the trials they drank the carbohydrate-only drink and in the other trial they drank the carbohydrate-protein drink. The order of the trials was counterbalanced and the distance they ran in both trials was recorded.
>
> An exact sign test was used to determine whether there was a statistically significant median difference between the distance ran when participants drank a carbohydrate-protein drink compared to a carbohydrate-only drink. Data are medians unless otherwise stated. Of the `r drink$sign_n` participants recruited to the study, the carbohydrate-protein drink elicited an increase in the distance run in `r sum(drink$sign_dat$diff > 0)` participants compared to the carbohydrate-only drink, whereas `r sum(drink$sign_dat$diff < 0)` participants did not run as far and `r sum(drink$sign_dat$diff == 0)` participant saw no improvement with the carbohydrate-protein drink. There was a statistically significant median increase in distance run (`r median(drink$sign_dat$diff) %>% scales::number(accuracy = .0001)` km) when subjects imbibed the carbohydrate-protein drink (`r median(drink$sign_dat$carb_protein) %>% scales::number(accuracy = .001)` km) compared to the carbohydrate-only drink (`r median(drink$sign_dat$carb) %>% scales::number(accuracy = .001)` km), *p* = `r drink$sign_test$p.value %>% scales::number(accuracy = .0001)`.

## One-way ANOVA {#onewayanova}

Classic analysis of variance (ANOVA) compares the mean responses from *experimental* studies. However, ANOVA also compares the mean responses from *observational* studies, but conclusions are just less rigorous. The one-way ANOVA is part of a larger family of statistical tests where the statistical model is called the General Linear Model (GLM). GLMs with a single dependent variable, such as the one-way ANOVA, can also be run through the GLM procedure. The GLM procedure can also handle multiple independent variables, fixed and random factors, weighting variables, and covariates. 

Most of these notes are gleaned from [PSU STAT-502](https://online.stat.psu.edu/stat502) "Analysis of Variance and Design of Experiments" covers ANOVA. [Laerd Statistics](https://statistics.laerd.com/premium/spss/owa/one-way-anova-in-spss.php) is useful for writing up your results for reports.

Use the one-way ANOVA test to compare the mean response of a continuous dependent variable among the (typically >=3) levels of a factor variable.

Here is a case study. Researchers compare the plant growth among three fertilizers and a control group. Data set `greenhouse` contains 6 observations per each of the *k* = 4 treatment levels (*N* = 24) - a balanced design.

```{r include=FALSE}
greenhouse <- tribble(
  ~group, ~growth,
"Control",      21,
"Control",      19.5,
"Control",      22.5,
"Control",      21.5,
"Control",      20.5,
"Control",      21,
"F1",      32,
"F1",      30.5,
"F1",      25,
"F1",      27.5,
"F1",      28,
"F1",      28.6,
"F2",      22.5,
"F2",      26,
"F2",      28,
"F2",      27,
"F2",      26.5,
'F2',      25.2,
"F3",      28,
"F3",      27.5,
"F3",      31,
"F3",      29.5,
"F3",      30,
"F3",      29.2
) %>%
  mutate(group = factor(group), id = row_number())

skimr::skim(greenhouse)
```

All three fertilizers produced more growth than the control group. Fertilizers *F1* and *F3* appear to be about tied for most growth, but it is unclear if the fertilizers are significantly different from each other.  

```{r echo=FALSE, fig.height=2.5}
greenhouse %>%
  group_by(group) %>%
  summarize(.groups = "drop",
            mean_growth = mean(growth),
            cl_025 = mean_growth + qnorm(.025) * sd(growth) / sqrt(n()),
            cl_975 = mean_growth + qnorm(.975) * sd(growth) / sqrt(n())) %>%
  ggplot(aes(x = group, y = mean_growth)) +
  geom_col(fill = "snow3", color = "snow3", alpha = 0.6, width = 0.5) +
  geom_errorbar(aes(ymin = cl_025, ymax = cl_975, width = 0.3)) +
  theme_minimal() +
  labs(title = "Mean Growth by Group",
       x = NULL, y = "Growth (cm)",
       caption = "Error Bars: 95% CI")
```

```{r echo=FALSE}
library(table1)
# table1 requires this metadata
label(greenhouse$growth) <- "Growth"
units(greenhouse$growth) <- "cm"

# Quick and good:
# table1(~ growth | group, data = greenhouse)

# Better: prints "Treated" at the top.
strata <- c(split(greenhouse, greenhouse$group),
            list("All treated" = subset(greenhouse, group %in% c("F1", "F2", "F3"))),
            list(Overall = greenhouse))
labels <- list(
  variables = list(growth = render.varlabel(greenhouse$growth)),
  groups = list("", "Treated", "")
)
table1(strata, labels, groupspan = c(1, 4, 1))
```

```{r}
greenhouse_desc <- greenhouse %>% 
  group_by(group) %>% 
  summarize(.groups = "drop", n = n(), mean = mean(growth), sd = sd(growth))
```

Data is presented as mean $\pm$ standard deviation. Plant growth (growth) increased from the control (*n* = `r pull(greenhouse_desc[1, "n"])`, `r pull(greenhouse_desc[1, "mean"])` $\pm$ `r format(pull(greenhouse_desc[1,"sd"]), nsmall = 1)`), to fertilizer 1 (*n* = `r pull(greenhouse_desc[2, "n"])`, `r pull(greenhouse_desc[2, "mean"])` $\pm$ `r format(pull(greenhouse_desc[2,"sd"]), digits = 2, nsmall = 1)`), fertilizer 2 (*n* = `r pull(greenhouse_desc[3, "n"])`, `r pull(greenhouse_desc[3, "mean"])` $\pm$ `r format(pull(greenhouse_desc[3,"sd"]), digits = 2, nsmall = 1)`), and fertilizer 3 (*n* = `r pull(greenhouse_desc[4, "n"])`, `r pull(greenhouse_desc[4, "mean"])` $\pm$ `r format(pull(greenhouse_desc[4,"sd"]), digits = 2, nsmall = 1)`) fertilizer groups.

ANOVA decomposes the deviation of observation $Y_{ij}$ around the overall mean $\bar{Y}_{..}$ into two parts: the deviation of the observations around their treatment means, $SSE$, and the deviation of the treatment means around the overall mean, $SSR$. Their ratio, $F = SSR/SSE$ follows an *F*-distribution with $k-1$ numerator dof and $N-k$ denominator dof. The more observation variance captured by the treatments, the large is $F$, and the less likely that the null hypothesis, $H_0 = \mu_1 = \mu_2 = \cdots = \mu_k$ is true.

```{r echo=FALSE}
tmp <- tribble(
  ~Source, ~SS, ~df, ~MS, ~F,
  "SSR", "$\\sum{n_i(\\bar{Y}_{i.} - \\bar{Y}_{..})^2}$", "$k - 1$", "${SSR}/{(k - 1)}$", "${MSR}/{MSE}$", 
  "SSE", "$\\sum(Y_{ij} - \\bar{Y}_{i.})^2$", "$N - k$", "${SSE}/{(N - k)}$", "",
  "SST", "$\\sum(Y_{ij} - \\bar{Y}_{..})^2$", "$N - 1$", "", ""
)

tmp %>% 
  knitr::kable(format = "html", caption = "ANOVA Table") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%
  kableExtra::row_spec(row = 0, align = "c") #%>%
  # kableExtra::footnote(
  #   general_title = "Note.",
  #   general = "Compare the *F*-statistic to the *F*-distribution with $k-1$ numerator degrees of freedom and $N-k$ denominator degrees of freedom",
  #  footnote_as_chunk = TRUE
  #   )
```

Run an ANOVA test in R like this:

```{r}
greenhouse_aov <- aov(growth ~ group, data = greenhouse)
greenhouse_anova <- anova(greenhouse_aov)

greenhouse_anova %>% 
  tidy() %>%
  flextable() %>%
  set_table_properties(width = 0.8, layout = "autofit") %>%
  colformat_num(j = c(3, 4, 5), digits = 1) %>%
  colformat_num(j = 6, digits = 4) %>%
  set_caption("Results of ANOVA for Growth vs Fertilizer Group")
```

The one-way ANOVA indicates amount of growth was statistically significantly different for different levels of fertilizer group, *F*(3, 20) = 27.5, *p* < .0001.

BTW, it is worth noting the relationship with linear regression. The regression model intercept is the overall mean and the coefficient estimators indirectly indicate the group means. The analysis of variance table in a regression model shows how much of the overall variance is explained by those coefficient estimators. It's the same thing.

You may also want to report the $\omega^2$ effect size, 

$$\omega^2 = \frac{SSR - df_R \cdot MSE}{MSE + SST}$$
```{r}
greenhouse_omega <- sjstats::anova_stats(greenhouse_anova) %>% 
  filter(term == "group") %>%
  pull(omegasq)
```
$\omega^2$ ranges from -1 to +1. In this example, $\omega^2$ is `r greenhouse_omega`.

#### ANOVA Conditions

The ANOVA test applies when the dependent variable is continuous, the independent variable is categorical, and the observations are independent *within* groups. Independence means the observations are from a random sample, or from an experiment using random assignment. Each group's size should be less than 10% of its population size. The groups must also be independent of each other (non-paired, and non-repeated measures). Additionally, there are three conditions related to the data distribution. If any condition does not hold, and the suggested work-arounds do not work, switch to the nonparametric Kruskal-Wallis Test (Chapter \@ref(kw)).

1. **No outliers**. There should be no significant outliers in the groups. Outliers exert a large influence on the mean and standard deviation. Test with a box plot. If there are outliers, you might be able to drop them or transform the data.
2. **Normality**.  Each group's values should be *nearly* normally distributed ("nearly" because ANOVA is considered robust to the normality assumption). This condition is especially important with small sample sizes. Test with the Q-Q plots or the Shapiro-Wilk test for normality. If the data is very non-normal, you might be able to transform your response variable.
3. **Equal Variances**.  The group variances should be roughly equal. This condition is especially important when sample sizes differ. Test with a box plot, rule of thumb, or one of the formal [homogeneity of variance](http://www.cookbook-r.com/Statistical_analysis/Homogeneity_of_variance/) (external) tests such as Bartlett, and Levene. If the variances are very different, use a Games-Howell post hoc test instead of the Tukey post hoc test.

##### Outliers {-}

Assess outliers with a box plot. Box plot whiskers extend up to 1.5\*IQR from the upper and lower hinges and outliers (beyond the whiskers) are are plotted individually. Our example includes an outlier in fertilizer group *F2*.

```{r echo=FALSE, fig.height=2.5}
greenhouse %>%
  ggplot(aes(x = group, y = growth)) +
  geom_boxplot(fill = "snow3", color = "snow4", alpha = 0.6, width = 0.5, 
               outlier.color = "goldenrod", outlier.size = 2) +
  theme_minimal() +
  labs(title = "Boxplot of Growth vs Fertilizer Group",
       y = "Growth (cm)", x = "Fertilizer Group")
```

Outliers might occur from data entry errors or measurement errors, so investigate and fix or throw them out. However, if the outlier is a genuine extreme value, you still have a couple options before reverting to Kruskal-Wallis.

* Transform the dependent variable. Don't do this unless the data is also non-normal. It also has the downside of making interpretation more difficult.
* Leave it in if it doesn't affect the conclusion (compared to taking it out).

Lets try removing the outlier (*id#* 13). 

```{r}
greenhouse_aov2 <- aov(growth ~ group, data = greenhouse %>% filter(!id == 13))
greenhouse_anova2 <- anova(greenhouse_aov2)
```

```{r echo=FALSE}
greenhouse_anova2 %>% 
  tidy() %>%
  flextable() %>%
  set_table_properties(width = 0.8, layout = "autofit") %>%
  colformat_num(j = c(3, 4, 5), digits = 1) %>%
  colformat_num(j = 6, digits = 4) %>%
  set_caption("Results of ANOVA for Growth vs Fertilizer Group") %>%
  footnote(i = 1, j = 1,
           value = as_paragraph("Note: One outlier in group F2 removed."),
           ref_symbols = c(""),
           part = "header", inline = TRUE)

```

The conclusion is the same, so leaving it in is fine!

##### Normality {-}

You can assume the populations are normally distributed if $n_j >= 30$. Otherwise, try the Q-Q plot, or skewness and kurtosis values, or histograms. If you still don't feel confident about normality, run a [Shapiro-Wilk Test] or Kolmogorov-Smirnov Test. If $n_j >= 50$, stick with graphical methods because at larger sample sizes Shapiro-Wilk flags even minor deviations from normality.

The QQ plots below appear to be approximately normal.

```{r}
greenhouse %>%
  ggplot(aes(sample = growth)) +
  stat_qq() +
  stat_qq_line(col = "goldenrod") +
  facet_wrap(~group) +
  theme_minimal() +
  labs(title = "Normal Q-Q Plot")
```

The Shapiro-Wilk test corroborates this conclusion - it fails to reject the null hypothesis of normally distributed populations.

```{r}
x <- by(greenhouse, greenhouse$group, function(x) shapiro.test(x$growth) %>% tidy())

x[1:4] %>%
  bind_rows() %>%
  mutate(group = names(x)) %>%
  dplyr::select(group, everything(), - method) %>%
  flextable() %>% 
  set_table_properties(width = 0.6, layout = "autofit") %>%
  set_caption("Shapiro-Wilk Normality Test")
```

If the data is not normally distributed, you still have a couple options before reverting to Kruskal-Wallis.

* Transform the dependent variable. Transformations will generally only work when the distribution of scores in all groups are the same shape. They also have the drawback of making the data less interpretable.
* carry on regardless. One-way ANOVA is fairly robust to deviations from normality, particularly if the sample sizes are nearly equal.

##### Equal Variances {-}

The equality of sample variances condition is less critical when sample sizes are similar among the groups. One rule of thumb is that no group's standard deviation should be more than double that of any other.  In this case `F1` is more than double `Control`.

```{r echo=FALSE}
greenhouse %>% 
  group_by(group) %>% 
  summarize(.groups = "drop", sd = sd(growth)) %>%
  flextable() %>%
  set_table_properties(width = 0.5, layout = "autofit")
```

There are two other common tests, Bartlett and Levene. NIST has a good write-up for [Levene](https://www.itl.nist.gov/div898/handbook/eda/section3/eda35a.htm) and for [Bartlett](https://www.itl.nist.gov/div898/handbook/eda/section3/eda357.htm). Levene is less sensitive than Bartlett to departures from normality, so if you know your data is normally distributed, then use Bartlett.

Levene's test fails to reject the null hypothesis of equality of variance.

```{r}
greenhouse_levene <- car::leveneTest(growth ~ group, data = greenhouse) 
greenhouse_levene %>% 
  tidy() %>%
  flextable() %>%
  set_table_properties(width = 0.6, layout = "autofit") %>% 
  set_caption("Levene's Test for Homogeneity of Variance")
```

So does Bartlett.  

```{r}
bartlett.test(growth ~ group, data = greenhouse) %>% 
  tidy() %>%
  dplyr::select(-method) %>%
  flextable() %>%
  set_table_properties(width = 0.6, layout = "autofit") %>% 
  set_caption("Bartlett's Test for Homogeneity of Variance")
```

Heterogeneity is a common problem in ANOVA. Transforming the response variable can often remove the heterogeneity. The Box-Cox procedure can help find a good transformation. The MASS::boxcox() function calculates a profile of log-likelihoods for a power transformation of the response variable $Y^\lambda$.

|$\lambda$ | $Y^\lambda$ | Transformation |
|---|---|---|
|2 | $Y^2$ | Square |
|1 | $Y^1$ | (no transformation) |
|.5 | $Y^{.5}$ | Square Root |
|0 | $\ln(Y)$ | Log |
|-.5 | $Y^{-.5}$ | Inverse Square Root |
|-1 | $Y^{-1}$ | Inverse|

The Box-Cox procedure does not recommend any particular transformation of the data in this case.

```{r message=FALSE}
MASS::boxcox(greenhouse_aov, plotit = TRUE)
```

#### Custom Contrasts

Taking this route is appropriate if you have specific hypotheses about the differences between the groups of your independent variable. For example, we might want to test whether the mean of the treatments differ from the control group, $H_0: \sum_i^K{c_i u_i} = 0$ where $c_i = (1, -1/3, -1/3, -1/3)$. You can test a constrast using the **multcomp** package.

```{r}
greenhouse_glht <- multcomp::glht(greenhouse_aov, linfct = multcomp::mcp(group = c(-1, 1/3, 1/3, 1/3)))
greenhouse_glht_smry <- summary(greenhouse_glht)
greenhouse_confint <- confint(greenhouse_glht)
greenhouse_glht_smry
```

Growth was statistically significantly higher in the fertilizer groups (mean of `r round(greenhouse_glht$coef[1] + mean(greenhouse_glht$coef[2:4]), 1)`) compared to the sedentary group (`r greenhouse_desc[1, ]$mean %>% round(1)` $\pm$ `r greenhouse_desc[1, ]$sd %>% round(1)`), a mean difference of `r greenhouse_confint$confint[1]` (95% CI, `r greenhouse_confint$confint[2]` to `r greenhouse_confint$confint[3]`), *p* = `r greenhouse_glht_smry$test$pvalues[1]`.

#### Tukey Post Hoc Test

The *F* test does not indicate which populations cause the rejection of $H_0$. For this, use one of the post-hoc tests: Tukey, Fisher's Least Significant Difference (LSD), Bonferroni, Scheffe, or Dunnett. Post hoc tests are appropriate if you are investigating all possible pairwise comparisons with no specific hypotheses about specific groups differing from others.

Here is the Tukey test. As expected, all three fertilizer factor levels differ from the control.  `F3` differed from `F2`, but `F1` was not significantly different from either `F2` or `F3`.

```{r}
greenhouse_tukey <- TukeyHSD(greenhouse_aov)
greenhouse_tukey %>% 
  tidy() %>%
  flextable() %>%
  set_table_properties(width = 0.8, layout = "autofit") %>% 
  colformat_num(j = c(4:6), digits = 1) %>%
  colformat_num(j = 7, digits = 3) %>%
  set_caption("Tukey multiple comparisons of means") %>%
  footnote(i = 1, j = c(1),
            value = as_paragraph(
              paste0("95% family-wise confidence level\n",
                "Fit: aov(formula = growth ~ group, data = greenhouse)")),
            ref_symbols = c(""),
            part = "header")
```

Data are mean $\pm$ standard deviation. There was an increase in growth from `r pull(greenhouse_desc[1, "mean"])` $\pm$ `r pull(greenhouse_desc[1, "sd"])` in the control group to `r pull(greenhouse_desc[2, "mean"])` $\pm$ `r format(pull(greenhouse_desc[2, "sd"]), digits = 2, nsmall = 1)` in the group with fertilizer F1, an increase of `r greenhouse_tukey$group[1, "diff"]` (95% CI, `r format(greenhouse_tukey$group[1, "lwr"], digits = 2, nsmall = 1)` to `r format(greenhouse_tukey$group[1, "upr"], digits = 2, nsmall = 1)`), which was statistically significant (*p* < .0001)... etc.

#### Reporting a One-Way ANOVA

Report like this.

> A one-way ANOVA was conducted to determine if plant growth was different for groups with different fertilizer types. Plants were classified into four groups: control (*n* = 6), fertilizer 1 (*n* = 6), fertilizer 2 (*n* = 6), and fertilizer 3 (n = 6). There was a single outlier, as assessed by boxplot, and was retained because it did not change the conclusions; data was normally distributed for each group, as assessed by Shapiro-Wilk test (*p* > .05); and there was homogeneity of variances, as assessed by Levene's test of homogeneity of variances (*p* = `r round(greenhouse_levene[1, 3], 3)`). Data is presented as mean $\pm$ standard deviation. Plant growth was statistically significantly different between different fertilizer groups, *F*(`r greenhouse_anova$Df[1]`, `r greenhouse_anova$Df[2]`) = `r round(greenhouse_anova[1, 4], 3)`, p < .0005, $\omega^2$ = `r greenhouse_omega`. Plant growth increased from the control (`r pull(greenhouse_desc[1, "mean"])` $\pm$ `r format(pull(greenhouse_desc[1,"sd"]), nsmall = 1)`), to fertilizer F1 (`r pull(greenhouse_desc[2, "mean"])` $\pm$ `r format(pull(greenhouse_desc[2,"sd"]), digits = 2, nsmall = 1)`), fertilizer F2 (`r format(pull(greenhouse_desc[3, "mean"]), digits = 3, nsmall = 1)` $\pm$ `r format(pull(greenhouse_desc[3,"sd"]), digits = 2, nsmall = 1)`), and fertilizer F3 (`r pull(greenhouse_desc[4, "mean"])` $\pm$ `r format(pull(greenhouse_desc[4,"sd"]), digits = 2, nsmall = 1)`) fertilizer groups. Tukey post hoc analysis revealed statistically significant increases from control to F1 (`r format(greenhouse_tukey$group[1, 1], digits=2, nsmall = 1)`, 95% CI (`r format(greenhouse_tukey$group[1, 2], digits=2, nsmall = 1)` to `r format(greenhouse_tukey$group[1, 3], digits=2, nsmall = 1)`), *p* = `r format(greenhouse_tukey$group[1, 4], digits=2, nsmall = 3)`), control to F2 (`r format(greenhouse_tukey$group[2, 1], digits=2, nsmall = 1)`, 95% CI (`r format(greenhouse_tukey$group[2, 2], digits=2, nsmall = 1)` to `r format(greenhouse_tukey$group[2, 3], digits=2, nsmall = 1)`), *p* = `r format(greenhouse_tukey$group[2, 4], digits=2, nsmall = 3)`), and control to F3 (`r format(greenhouse_tukey$group[3, 1], digits=2, nsmall = 1)`, 95% CI (`r format(greenhouse_tukey$group[3, 2], digits=2, nsmall = 1)` to `r format(greenhouse_tukey$group[3, 3], digits=2, nsmall = 1)`), *p* = `r format(greenhouse_tukey$group[3, 4], digits=2, nsmall = 3)`), as well as the increase from F2 to F3 (`r format(greenhouse_tukey$group[6, 1], digits=2, nsmall = 1)`, 95% CI (`r format(greenhouse_tukey$group[6, 2], digits=2, nsmall = 1)` to `r format(greenhouse_tukey$group[6, 3], digits=2, nsmall = 1)`), *p* = `r format(greenhouse_tukey$group[6, 4], digits=2, nsmall = 3)`), but there were no statistically significant group differences between F1 and F2 or F1 and F3.


### Welch's ANOVA w/Games-Howell

Welch's ANOVA test is an alternative to the one-way ANOVA test in cases where the equality of variances assumption is violated.

Here is a case study. Researchers compare the force (in newtons) generated in three steps. Data set `newton` contains 30 observations per each of the *k* = 3 step levels (*N* = 90) - a balanced design.

```{r echo=FALSE, fig.height=2.5}
newton <- data.frame(
  newtons = c(400.95, 445.58, 499.7, 344.62, 397.19, 424.1, 692.2, 436.68, 541.27, 
              518.01, 618.62, 493.19, 368.08, 405.96, 391.81, 305.59, 345.88, 458.15, 
              320.92, 371.11, 373.94, 434.02, 484.49, 360.24, 495.93, 464.1, 455.07, 
              343.19, 358.93, 325.86, 524.63, 584.21, 592.76, 567.12, 716.15, 481.64, 
              518.3, 759, 419.33, 641.5, 713.04, 657.78, 469.29, 420.89, 440.02, 417.39, 
              489.02, 586.63, 477.39, 461.69, 564.12, 565.96, 427.77, 503.48, 494.44, 
              577.08, 425.99, 428.64, 461.9, 417.68, 820.65, 788.84, 644.03, 779.9, 
              892.91, 571.47, 625.09, 751.91, 582.32, 870.45, 886.16, 938.5, 605.69, 
              572.6, 643.57, 770.42, 660.53, 576.62, 437.25,593.72, 441.11, 591.43, 
              749.08, 670.02, 475.11, 455.26, 526.59, 560.3, 474.65, 511.5),
  step = factor(c(rep("A", 30), rep("B", 30), rep("C", 30)))
)

newton %>%
  group_by(step) %>%
  summarize(.groups = "drop",
            mean = mean(newtons),
            cl_025 = mean + qnorm(.025) * sd(newtons) / sqrt(n()),
            cl_975 = mean + qnorm(.975) * sd(newtons) / sqrt(n())) %>%
  ggplot(aes(x = step, y = mean)) +
  geom_col(fill = "snow3", color = "snow3", alpha = 0.6, width = 0.5) +
  geom_errorbar(aes(ymin = cl_025, ymax = cl_975, width = 0.3)) +
  theme_minimal() +
  labs(title = "Mean Newtons by Step",
       x = "Step", y = "Newtons",
       caption = "Error Bars: 95% CI")
```

```{r echo=FALSE}
# table1 requires this metadata
label(newton$newtons) <- "Force"
units(newton$newtons) <- "newtons"

table1(~ newtons | step, data = newton)
```

```{r}
newton_desc <- newton %>% 
  group_by(step) %>% 
  summarize(.groups = "drop", n = n(), mean = mean(newtons), sd = sd(newtons))
```

Data is presented as mean $\pm$ standard deviation. Force (newtons) increased from step 1 (*n* = `r pull(newton_desc[1, "n"])`, `r pull(newton_desc[1, "mean"]) %>% format(digits = 3, nsmall = 0)` $\pm$ `r pull(newton_desc[1,"sd"]) %>% format(digits = 3, nsmall = 0)`), to step 2 (*n* = `r pull(newton_desc[2, "n"])`, `r pull(newton_desc[2, "mean"]) %>% format(digits = 3, nsmall = 0)` $\pm$ `r pull(newton_desc[2,"sd"]) %>% format(digits = 3, nsmall = 0)`), to step 3 (*n* = `r pull(newton_desc[3, "n"])`, `r pull(newton_desc[3, "mean"]) %>% format(digits = 3, nsmall = 0)` $\pm$ `r pull(newton_desc[3,"sd"]) %>% format(digits = 3, nsmall = 0)`).

Start by running the standard ANOVA test:

```{r}
newton_aov <- aov(newtons ~ step, data = newton)
newton_anova <- anova(newton_aov)

newton_anova %>% 
  tidy() %>%
  flextable() %>%
  set_table_properties(width = 0.8, layout = "autofit") %>%
  colformat_num(j = c(3, 4, 5), digits = 1) %>%
  colformat_num(j = 6, digits = 4) %>%
  set_caption("Results of ANOVA for Force vs Step")
```

The one-way ANOVA indicates amount of force was statistically significantly different for different levels of step, *F*(2, 87) = 28.4, *p* < .0001.

#### ANOVA Conditions

Check the three ANOVA conditions: no outliers, normality, and equal variances.

##### Outliers {-}

Assess outliers with a box plot. Our example includes an outlier in step *A*.

```{r echo=FALSE, fig.height=2.5}
newton %>%
  ggplot(aes(x = step, y = newtons)) +
  geom_boxplot(fill = "snow3", color = "snow4", alpha = 0.6, width = 0.5, 
               outlier.color = "goldenrod", outlier.size = 2) +
  theme_minimal() +
  labs(title = "Boxplot of Force vs Step",
       y = "Force (newtons)", x = "Step")
```

You can either transform the dependent variable, see if taking it out changes your conclusion, or use a nonparametric test. Let's try removing the outlier (*id#* 13). 

```{r}
newton2 <- newton %>% mutate(id = row_number())
newton_aov2 <- aov(newtons ~ step, data = newton2 %>% filter(!id == 7))
newton_anova2 <- anova(newton_aov2)
```

```{r echo=FALSE}
newton_anova2 %>% 
  tidy() %>%
  flextable() %>%
  set_table_properties(width = 0.8, layout = "autofit") %>%
  colformat_num(j = c(3, 4, 5), digits = 1) %>%
  colformat_num(j = 6, digits = 4) %>%
  set_caption("Results of ANOVA for Force vs Step") %>%
  footnote(i = 1, j = 1,
           value = as_paragraph("Note: One outlier in step A removed."),
           ref_symbols = c(""),
           part = "header", inline = TRUE)

```

The conclusion is the same, so leaving it in is fine!

##### Normality {-}

You can assume the populations are normally distributed if $n_j >= 30$, but I'll examine the Q-Q plot and run a [Shapiro-Wilk Test] anyway.

The QQ plots below appear to be approximately normal...

```{r}
newton %>%
  ggplot(aes(sample = newtons)) +
  stat_qq() +
  stat_qq_line(col = "goldenrod") +
  facet_wrap(~step) +
  theme_minimal() +
  labs(title = "Normal Q-Q Plot")
```

...but the Shapiro-Wilk test fails for step *A* and *B* -- evidence of its sensitivity for large *n*. I will ignore this violation.

```{r}
x <- by(newton, newton$step, function(x) shapiro.test(x$newtons) %>% tidy())

x[1:3] %>%
  bind_rows() %>%
  mutate(group = names(x)) %>%
  dplyr::select(group, everything(), - method) %>%
  flextable() %>% 
  set_table_properties(width = 0.6, layout = "autofit") %>%
  set_caption("Shapiro-Wilk Normality Test")
```

##### Equal Variances {-}

The equality of sample variances condition is less critical when sample sizes are similar among the groups. Following the rule of thumb that no group's standard deviation be more than double that of any other, we look okay.

```{r echo=FALSE}
newton %>% 
  group_by(step) %>% 
  summarize(.groups = "drop", sd = sd(newtons)) %>%
  flextable() %>%
  set_table_properties(width = 0.5, layout = "autofit")
```

However, Levene's test rejects the null hypothesis of equality of variance.

```{r}
newton_levene <- car::leveneTest(newtons ~ step, data = newton) 
newton_levene %>% 
  tidy() %>%
  flextable() %>%
  set_table_properties(width = 0.6, layout = "autofit") %>% 
  set_caption("Levene's Test for Homogeneity of Variance")
```

So does Bartlett.  

```{r}
bartlett.test(newtons ~ step, data = newton) %>% 
  tidy() %>%
  dplyr::select(-method) %>%
  flextable() %>%
  set_table_properties(width = 0.6, layout = "autofit") %>% 
  set_caption("Bartlett's Test for Homogeneity of Variance")
```

We could transform the response variable to remove the heterogeneity. The Box-Cox procedure suggests an inverse square root transformation.

|$\lambda$ | $Y^\lambda$ | Transformation |
|---|---|---|
|2 | $Y^2$ | Square |
|1 | $Y^1$ | (no transformation) |
|.5 | $Y^{.5}$ | Square Root |
|0 | $\ln(Y)$ | Log |
|-.5 | $Y^{-.5}$ | Inverse Square Root |
|-1 | $Y^{-1}$ | Inverse|

The Box-Cox procedure does not recommend any particular transformation of the data in this case.

```{r message=FALSE}
# MASS::boxcox(newton_aov, plotit = TRUE)
```

```{r}
newton3 <- newton %>%
  mutate(newtons_isr = newtons^(-0.5))
newton_levene3 <- car::leveneTest(newtons_isr ~ step, data = newton3) 
newton_levene3 %>% 
  tidy() %>%
  flextable() %>%
  set_table_properties(width = 0.6, layout = "autofit") %>% 
  set_caption("Levene's Test for Homogeneity of Variance")
```

Huzzah - it worked! Before we continue on, we should backtrack and re-test the outliers and normality conditions. However, because the point of this section is to try Welch's ANOVA, I'm going use it instead of transforming the response variable. Use `oneway.test(..., var.equal = FALSE)` to run a Welch's ANOVA.

```{r}
newton_anova <- oneway.test(newtons ~ step, data = newton, var.equal = FALSE)
newton_anova
```

Welch's ANOVA indicates amount of force was statistically significantly different for different steps, *F*(2, 56.2) = 26.2, *p* < .0001.

I don't think you can calculate $\omega^2$ for a Welch's ANOVA object.

#### Games-Howell Post Hoc Test

Use the `PMCMRplus::gamesHowellTest()` to run the Games-Howell post hoc test. As expected, the three steps differ from each other.

```{r message=FALSE}
newton_games_howell <- rstatix::games_howell_test(newton, newtons ~ step)
newton_games_howell %>%
  flextable() %>% autofit() %>% set_caption("Games-Howell Post Hoc Test")
```

#### Reporting a Welch's ANOVA

> A Welch's ANOVA was conducted to determine if force was different for different steps. Measurements were classified into three groups: A (*n* = 30), B (*n* = 30), and C (n = 30). There was a single outlier, as assessed by boxplot, and was retained because it did not change the conclusions; data was normally distributed for each group, as assessed by Q-Q plot. ; Homogeneity of variances was violated, as assessed by Levene's Test of Homogeneity of Variance (*p* = `r round(newton_levene[1, 3], 3)`). Data is presented as mean $\pm$ standard deviation. Force was statistically significantly different between different steps, *F*(`r newton_anova$parameter[1]`, `r newton_anova$parameter[2]`) = `r round(newton_anova$statistic, 3)`, p < .0005. Force increased from A (`r pull(newton_desc[1, "mean"])` $\pm$ `r format(pull(newton_desc[1,"sd"]), nsmall = 1)`), to B (`r pull(newton_desc[2, "mean"])` $\pm$ `r format(pull(newton_desc[2,"sd"]), digits = 2, nsmall = 1)`), to C (`r pull(newton_desc[3, "mean"])` $\pm$ `r format(pull(newton_desc[3,"sd"]), digits = 2, nsmall = 1)`). Games-Howell post hoc analysis revealed statistically significant increases from A to B, (`r format(newton_games_howell[1, ]$estimate, digits=2, nsmall = 1)`, 95% CI (`r format(newton_games_howell[1, ]$conf.low, digits=2, nsmall = 1)` to `r format(newton_games_howell[1, ]$conf.low, digits=2, nsmall = 1)`), *p* = `r format(newton_games_howell[1, ]$p.adj, digits=1, nsmall = 3)`), A to C (`r format(newton_games_howell[2, ]$estimate, digits=2, nsmall = 1)`, 95% CI (`r format(newton_games_howell[2, ]$conf.low, digits=2, nsmall = 1)` to `r format(newton_games_howell[2, ]$conf.low, digits=2, nsmall = 1)`), *p* = `r format(newton_games_howell[2, ]$p.adj, digits=1, nsmall = 3)`), and B to C (`r format(newton_games_howell[3, ]$estimate, digits=2, nsmall = 1)`, 95% CI (`r format(newton_games_howell[3, ]$conf.low, digits=2, nsmall = 1)` to `r format(newton_games_howell[3, ]$conf.low, digits=2, nsmall = 1)`), *p* = `r format(newton_games_howell[3, ]$p.adj, digits=1, nsmall = 3)`).


### MANOVA

Multi-factor ANOVA (MANOVA) is a method to compare mean responses by treatment factor level of two or more treatments applied in combination. The null hypotheses are $H_0: \mu_{1.} = \mu_{2.} = \dots = \mu_{a.}$ for the $a$ levels of factor 1, $H_0: \mu_{.1} = \mu_{.2} = \dots = \mu_{.b}$ for the $b$ levels of factor 2, etc. for all the factors in the experiment, and $H_0: $ no interaction for all the factor interactions.

There are two equivalent ways to state the MANOVA model:

$$Y_{ijk} = \mu_{ij} + \epsilon_{ijk}$$

In this notation $Y_{ijk}$ refers to the $k^{th}$ observation in the $j^{th}$ level of factor two and the $i^{th}$ level of factor 1.  Potentially there could be additional factors.  This model formulation decomposes the response into a cell mean and an error term.  The second makes the factor effect more explicit and is thus more common:

$$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} +  \epsilon_{ijk}$$

### Multiple Variance Comparison F Test


### Example
*A study investigates the relationship between oxygen update and two explanatory variables: smoking, and type of stress test.  A sample of* $n = 27$ *persons, 9 non-smoking, 9 moderately-smoking, and 9 heavy-smoking are divided into three stress tests, bicycle, treadmill, and steps and their oxygen uptake was measured.  Is oxygen uptake related to smoking status and type of stress test?  Is there an interaction effect between smoking status and type of stress test?*
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(nortest)  # for Anderson-Darling test
library(stats)  # for anova

smoker <- c(1, 1, 1, 1, 1, 1, 1, 1, 1, 
            2, 2, 2, 2, 2, 2, 2, 2, 2, 
            3, 3, 3, 3, 3, 3, 3, 3, 3)
stress <- c(1, 1, 1, 2, 2, 2, 3, 3, 3,
            1, 1, 1, 2, 2, 2, 3, 3, 3,
            1, 1, 1, 2, 2, 2, 3, 3, 3)
oxytime <- c(12.8, 13.5, 11.2, 16.2, 18.1, 17.8, 22.6, 19.3, 18.9,
             10.9, 11.1, 9.8, 15.5, 13.8, 16.2, 20.1, 21.0, 15.9,
             8.7, 9.2, 7.5, 14.7, 13.2, 8.1, 16.2, 16.1, 17.8)
oxy <- data.frame(oxytime, smoker, stress)
oxy$smoker <- ordered(oxy$smoker,
                      levels = c(1, 2, 3),
                      labels = c("non-smoker", "moderate", "heavy"))
oxy$stress <- factor(oxy$stress,
                     labels = c("bicycle", "treadmill", "steps"))

lm_oxy <- lm(oxytime~smoker+stress+smoker*stress, data = oxy)
anova(lm_oxy)
```



[SFU BIO710](http://online.sfsu.edu/efc/classes/biol710/manova/MANOVAnewest.pdf)

## KruskalWallis Test {#kw}

The **Kruskal-Wallis H test**^[The Kruskal-Wallis H test is also called the one-way ANOVA on ranks] measures the difference of a continuous or ordinal dependent variable between two or more categorical, independent groups. It is a rank-based nonparametric alternative to the one-way ANOVA test. Use Kruskal-Wallis if the dependent variable fails the ANOVA normality condition, or if the dependent variable is ordinal.

**Assumptions**

Kruskal-Wallis has no assumptions per se, but the interpretation of its results depend on the distribution of the dependent variable. If the dependent variable distribution has a similar shape across the groups of the independent categorical variable, then Kruskal-Wallis is a test of differences in their medians. Otherwise, Kruskal-Wallis is a test of differences in their distributions.

**How it Works**

The Kruskal-Wallis H test ranks the dependent variable irrespective of its group. It averages the ranks by the independent categorical variable groups. The test statistic approximately follows a $\chi^2$ distribution with _k_  1 degrees of freedom, where _k_ is the number of groups of the independent variable. The null hypothesis is that the rank means are equal. If you reject the null hypothesis, run a post hoc test to determine which groups differ.

## Case Study 3 {-}

Laerd Statistics's data sets for [ANOVA](https://statistics.laerd.com/premium/spss/owa/one-way-anova-in-spss-6.php) and [Kruskal-Wallis](https://statistics.laerd.com/premium/spss/kwht/kruskal-wallis-test-in-spss-6.php) are identical, so I'll use a single data set for the case study.

```{r}
cs3 <- list()

# cs3$kw_dat <- read.spss("./input/kruskal-wallis-h-test.sav", to.data.frame = TRUE)
# cs3$anova_dat <- read.spss("./input/one-way-anova.sav", to.data.frame = TRUE)
cs3$dat <- read.spss("./input/kruskal-wallis-h-test.sav", to.data.frame = TRUE)
```

A study tests whether physically active individuals are more are better able to cope with workplace stress. The researcher categorizes $n$ = `r nrow(cs3$dat)` participants by physical activity level ("Sedentary", "Low", "Moderate", and "High") and measures their ability to cope with workplace-related stress (CWWS) as the average score of a series of Likert items on a questionnaire (higher scores indicating a greater CWWS ability). Is CWWS dependent on physical activity level?

```{r}
(cs3$gt <- cs3$dat %>% tbl_summary(by = group, label = list(coping_stress = "CWWR")))
```

### Assumptions

Recall that the one-way ANOVA test is valid under three assumptions. One, there are no significant outliers that influence the group mean. Two, the dependent variable is at least _approximately_ (ANOVA is robust to this assumption) normally distributed for each group if the sample size is small (for large sample sizes the Central Limit Theorem shows normality is unnecessary). Three, the dependent variable should have equal variances across groups. ANOVA is only sensitive to this condition if the group sample sizes are not similar.

Kruskal-Wallis has no assumptions per se, but the interpretation of its results depend on the distribution of the dependent variable. If the distributions are similar, then the test results tell you whether the medians differ. Otherwise, the test results tell you whether the distributions differ.

Use a boxplot to assess outliers and the data distribution. Values greater than 1.5 IQR from the box edges constitute an outlier. There are no outliers here. If there had been, and you cannot justify removing the data points on grounds of bad data, you can still use the ANOVA test if it leads to the same conclusion with and without the outlier(s). You can also try transforming the dependent variable. However, the best practice is to revert to the Kruskal-Wallis H test. There is no accepted practice for determining whether distributions are similar. The boxplot reveals a wider range of values for "Low" group, but this is close enough to conclude the distributions are similar. 

```{r fig.height=3.5, fig.width=7.5}
cs3$dat %>%
  ggplot(aes(x = group, y = coping_stress)) +
  geom_boxplot()
```

Use the Shapiro-Wilk test of normality if your sample size is small (<50) and the Normal Q-Q plot is ambiguous. The Normal Q-Q plot below looks good for all groups except perhaps the "Low" group. The Shapiro-Wilk test confirms this, with all *p*-values over .05. Had the data failed the normality test, you could probably carry on anyway since the test is robust to normality, or you could try transforming the dependent variable, or revert to the Kruskal-Wallis H test. 

```{r}
cs3$dat %>% 
  ggplot(aes(sample = coping_stress)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(facets = vars(group)) +
  labs(title = "Q-Q Plot", x = "Theoretical", y = "Sample")
```

```{r}
with(cs3$dat, by(coping_stress, group, shapiro.test))
```

ANOVA's equality of sample variances condition is less critical when sample sizes are similar among the groups (as they are here). A rule of thumb is that no group's standard deviation should be more than double that of any other. In this case, "Moderate" and "Low" are more than double "Sedentary".

```{r echo=FALSE}
cs3$dat %>% 
  group_by(group) %>% 
  summarize(.groups = "drop", n = n(), sd = sd(coping_stress)) %>%
  arrange(sd) %>%
  mutate(multiple = sd / first(sd))
```

There are two common tests, Bartlett and Levene^[NIST has a good write-up for [Bartlett](https://www.itl.nist.gov/div898/handbook/eda/section3/eda357.htm) and [Levene](https://www.itl.nist.gov/div898/handbook/eda/section3/eda35a.htm)]. Levene is less sensitive to departures from normality. Neither test rejects the null hypothesis of equality of variance here.

```{r collapse=TRUE}
(cs3$levene <- car::leveneTest(coping_stress ~ group, data = cs3$dat))
(cs3$bartlet <- bartlett.test(coping_stress ~ group, data = cs3$dat))
```

Heterogeneity of variances is a common problem in ANOVA. Box-Cox procedure can help find a good transformation to remove heterogeneity. `MASS::boxcox()` calculates a profile of log-likelihoods for a power transformation of the dependent variable $Y^\lambda$.

|$\lambda$ | $Y^\lambda$ | Transformation |
|---|---|---|
|2 | $Y^2$ | Square |
|1 | $Y^1$ | (no transformation) |
|.5 | $Y^{.5}$ | Square Root |
|0 | $\ln(Y)$ | Log |
|-.5 | $Y^{-.5}$ | Inverse Square Root |
|-1 | $Y^{-1}$ | Inverse|

The Box-Cox procedure does not recommend any particular transformation of the data in this case.

```{r message=FALSE}

MASS::boxcox(aov(coping_stress ~ group, data = cs3$dat), plotit = TRUE)
```

### ANOVA

<ANOVA with post hoc testing vs custom contrasts>

<Conduct a Tukey post hoc test if you have no specific hypothesis about two groups differing or want to see all group differences. If you want to compare two groups or set of groups, then use a custom contrast.>

<To generate the effect size called partial eta squared ($\eta^2$)>



### Kruskal-Wallis Test

Run a Kruskal-Wallis H test with `kruskal.test()`. 

```{r}
(cs3$kruskal <- kruskal.test(coping_stress ~ group, data = cs3$dat))
```

The dependent variable has similarly shaped distributions for all groups of the independent variable, so you can conclude the median CWWS scores were statistically significantly different between groups, $\chi^2$(`r cs3$kruskal$parameter`) = `r comma(cs3$kruskal$statistic, .1)`, *p* = `r comma(cs3$kruskal$p.value, .0001)`. Otherwise you would conclude the _distributions_ differ. You rejected the null hypothesis, so continue on with a post hoc test to determine which medians (similar distributions) or mean ranks (dissimilar distributions) differ with the Dunn procedure using a Bonferroni correction for multiple comparisons.

```{r}
(cs3$dunn <- FSA::dunnTest(coping_stress ~ group, data = cs3$dat, method = "bonferroni"))
```

*P.adj* equals *P.unadj* multiplied by the number of comparisons (`r nrow(cs3$dunn$res)`). You could report the adjusted _p_ or the unadjusted _p_ with a note that you accepted statistical significance at the _p_ < .05 / `r nrow(cs3$dunn$res)` = `r comma(.05 / nrow(cs3$dunn$res), .0001)` level.

Now you can report your results.

> A Kruskal-Wallis test was conducted to determine if there were differences in CWWS scores between groups that differed in their level of physical activity: the "sedentary" (_n_ = `r inline_text(cs3$gt, variable = coping_stress, column = "Sedentary", pattern = "{N_obs}")`), "low" (n = `r inline_text(cs3$gt, variable = coping_stress, column = "Low", pattern = "{N_obs}")`), "moderate" (n = `r inline_text(cs3$gt, variable = coping_stress, column = "Moderate", pattern = "{N_obs}")`) and "high" (n = `r inline_text(cs3$gt, variable = coping_stress, column = "High", pattern = "{N_obs}")`) physical activity level groups. Distributions of CWWS scores were similar for all groups, as assessed by visual inspection of a boxplot. Median CWWS scores were statistically significantly different between the different levels of physical activity group, $\chi^2$(`r cs3$kruskal$parameter`) = `r comma(cs3$kruskal$statistic, .1)`, *p* = `r comma(cs3$kruskal$p.value, .0001)`. Subsequently, pairwise comparisons were performed using Dunn's (1964) procedure with a Bonferroni correction for multiple comparisons. Adjusted p-values are presented. This post hoc analysis revealed statistically significant differences in CWWS scores between the sedentary (Mdn = `r inline_text(cs3$gt, variable = coping_stress, column = "Sedentary", pattern = "{median}")`) and moderate (Mdn = `r inline_text(cs3$gt, variable = coping_stress, column = "Moderate", pattern = "{median}")`) (*p* = `r cs3$dunn$res %>% filter(Comparison == "Moderate - Sedentary") %>% pull(P.adj) %>% comma(.0001)`) and sedentary and high (Mdn = `r inline_text(cs3$gt, variable = coping_stress, column = "High", pattern = "{median}")`) (*p* = `r cs3$dunn$res %>% filter(Comparison == "High - Sedentary") %>% pull(P.adj) %>% comma(.0001)`) physical activity groups, but not between the low physical activity group (Mdn = `r inline_text(cs3$gt, variable = coping_stress, column = "Low", pattern = "{median}")`) or any other group combination.

Had the distributions been different, you would report "CWWS scores" instead of "Median CWWS scores" and report the mean ranks instead of Mdn. Unfortunately, you cannot retrieve those ranks from the test object, so you would have to calculate them yourself.

## Chi-Square Test of Homogeneity {#chisqhomogeneity}

The chi-square test of homogeneity tests whether frequency counts of the *R* levels of a categorical variable are distributed identically across the *C* populations. It tests whether observed joint frequency counts $O_{ij}$ differ from expected frequency counts $E_{ij}$ under the *independence model* (the model of independent explanatory variables, $\pi_{ij} = \pi_{i+} \pi_{+j}$. $H_0$ is $O_{ij} = E_{ij}$. The chi-square homogeneity test can be extended to cases where $I$ and/or $J$ is greater than 2.

There are two possible test statistics for this test, Pearson $X^2 = \sum \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$, and deviance $G^2 = 2 \sum_{ij} O_{ij} \log \left( \frac{O_{ij}}{E_{ij}} \right)$.

**Side note: z-Test of Two Proportions**

The z-test and chi-square test produce the same statistical significance result because they are algebraically identical.

The *z*-test uses the difference in sample proportions $\hat{d} = p_1 - p_2$ as an estimate of the difference in population proportions $\delta = \pi_1 - \pi_2$ to evaluate an hypothesized difference in population proportions $d_0 = \pi_0 - \pi_1$ and/or construct a $(1\alpha)\%$ confidence interval around $\hat{d}$ to estimate $\delta$ within a margin of error $\epsilon$.

The *z*-test applies when the central limit theorem conditions hold so that the normal distribution approximates the binomial distribution.

* the sample is independently drawn, meaning random assignment (experiments) or random sampling without replacement from $n < 10\%$ of the population (observational studies),
* there are at least $n_i p_i >= 5$ successes and $n_i (1 - p_i) >= 5$ failures for each group $i$,
* the sample sizes are both $n_i >= 30$, and 
* the probability of success for each group is not extreme, $0.2 < \pi_i < 0.8$.

If these conditions hold, the sampling distribution of $\delta$ is normally distributed around $\hat{d}$ with standard error $se_\hat{d} = \sqrt{\frac{p_1(1 - p_1)}{n_1} + \frac{p_2(1  p_2)}{n_2}}$. The measured values $\hat{d}$ and $se_\hat{d}$ approximate the population values $\delta$ and $se_\delta$. Define a $(1  \alpha)\%$ confidence interval as $\hat{d} \pm z_{\alpha / 2}se_\hat{d}$ or test the hypothesis of $d = d_0$ with test statistic $z = \frac{\hat{d}  d_0}{se_{d_0}}$ where $se_{d_0} = \sqrt{p^*(1 - p^*) \left(\frac{1}{n_1} + \frac{1}{n_2}\right)}$ and $p^*$ is the overall success probability.


## Fisher's Exact Test {#fisherexact}

Fisher's exact test is an "exact test" in that the p-value is calculated exactly from the hypergeometric distribution rather than relying on the approximation that the test statistic distribution approaches $\chi^2$ as $n \rightarrow \infty$.

The test is applicable in situations where

* the row totals $n_{i+}$ and the column totals $n_+j$ are fixed by study design (rarely applies), and
* the expected values of >20% of cells (at least 1 cell in a 2x2 table) have expected cell counts >5, and no expected cell count is <1.

The p-value from the test is computed as if the margins of the table are fixed. This leads under a null hypothesis of independence to a hypergeometric distribution of the numbers in the cells of the table ([Wikipedia](https://en.wikipedia.org/wiki/Fisher%27s_exact_test)). Fisher's exact test is useful for small *n*-size samples where the chi-squared distribution assumption of the chi-squared and G-test tests fails. Fisher's exact test is overly conservative (p values too high) for large *n*-sizes.

The Hypergeometric density function is
$$f_X(k|N, K, n) = \frac{{{K}\choose{k}}{{N-K}\choose{n-k}}}{{N}\choose{n}}.$$

The density is the exact hypergeometric probability of observing this particular arrangement of the data, assuming the given marginal totals, on the null hypothesis that the conditional probabilities are equal.

## Case Study 4 {-}

```{r include=FALSE}
ind_discrete <- list()

ind_discrete$chisq_dat <- read.spss(
  "./input/test-of-two-proportions-individual-scores.sav",
  to.data.frame = TRUE
)

ind_discrete$fisher_dat <- ind_discrete$chisq_dat[seq(2, 100, 2),]
```

The case study below uses a data set from [Laerd](https://statistics.laerd.com/) and a second modified version. The first data set passes the chi-square test of homogeneity requirements. The second (in parentheses), fails the n-sizes test. 

A researcher recruits `r nrow(ind_discrete$chisq_dat)` (`r nrow(ind_discrete$fisher_dat)`) patients who have a "high" classification of cholesterol and who currently have a poor lifestyle. The researcher randomly assigns `r ind_discrete$chisq_dat %>% filter(intervention == "Drug") %>% nrow()` (`r ind_discrete$fisher_dat %>% filter(intervention == "Drug") %>% nrow()`) of them to a drug intervention and `r ind_discrete$chisq_dat %>% filter(intervention == "Lifestyle") %>% nrow()` (`r ind_discrete$fisher_dat %>% filter(intervention == "Lifestyle") %>% nrow()`) to a lifestyle intervention. After six months, a doctor reclassifies the patients as either still having a "high" classification of cholesterol or now having a "normal" classification of cholesterol.

```{r fig.height=3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `Chi-sq` = ind_discrete$chisq_dat,
  `Fisher` = ind_discrete$fisher_dat,
  .id = "set"
) %>%
  count(set, intervention, risk_level) %>%
  group_by(set, intervention) %>%
  summarize(.groups = "drop",
            p = sum(if_else(risk_level == "Normal", n, as.integer(0))) / sum(n),
            n = sum(n),
            se = sqrt(p * (1 - p) / n),
            ci_lwr = p - qt(.975, n - 1) * se,
            ci_upr = p + qt(.975, n - 1) * se) %>%
  ggplot(aes(x = intervention)) +
  geom_col(aes(y = p), fill = "snow3", color = "snow4", width = 0.25) +
  geom_errorbar(aes(ymin = ci_lwr, ymax = ci_upr), color = "snow4", width = .125) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~set) +
  labs(x = NULL, y = NULL,
       title = "Cholesterol Classification Improvement from High to Normal",
       subtitle = "Two data sets with different n-sizes.",
       caption = "Error bars show 95% CI.")
```

The chi-sq data set has the following summary statistics.

```{r echo=FALSE}
x <- ind_discrete$chisq_dat %>%
  tabyl(intervention, risk_level) %>%
  adorn_totals(where = c("row", "col")) %>%
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")
  
ind_discrete$chisq_test <- ind_discrete$chisq_dat %>%
  tabyl(intervention, risk_level) %>%
  chisq.test()
y <- ind_discrete$chisq_test$expected %>%
  adorn_totals(where = c("row", "col")) %>%
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")

bind_rows(list(Observed = x, Expected = y), .id = "Data") %>%
  flextable() %>% 
  set_caption("Chisq Data Set: Observed vs Expected") %>% 
  autofit() %>%
  flextable::border(i = 3, border.bottom = officer::fp_border()) %>%
  flextable::merge_v(j = 1)
```

The Fisher data set has the following summary statistics.

```{r echo=FALSE}
x <- ind_discrete$fisher_dat %>%
  tabyl(intervention, risk_level) %>%
  adorn_totals(where = c("row", "col")) %>%
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")
  
ind_discrete$fisher_chisq_test <- ind_discrete$fisher_dat %>%
  tabyl(intervention, risk_level) %>%
  chisq.test()
y <- ind_discrete$fisher_chisq_test$expected %>%
  adorn_totals(where = c("row", "col")) %>%
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")

bind_rows(list(Observed = x, Expected = y), .id = "Data") %>%
  flextable() %>% 
  set_caption("Fisher Data Set: Observed vs Expected") %>% 
  autofit() %>%
  flextable::border(i = 3, border.bottom = officer::fp_border()) %>%
  flextable::merge_v(j = 1)
```

#### Conditions {-}

##### n-Size {-}

The chi-square test of homogeneity applies with the CLT conditions hold.

* the sample is independently drawn,
* there are at least 5 successes (Normal) and failures (High) for each group $i$,
* the sample sizes for both groups are >=30, and 
* the probability of success for each group is not extreme, $0.2 < \pi_i < 0.8$.

The conditions hold for the chi-sq data set, but not for the Fisher data set.

#### Test {-}

##### Chi-Square {-}

```{r}
(ind_discrete$chisq_test <- ind_discrete$chisq_dat %>%
  tabyl(intervention, risk_level) %>%
  chisq.test(correct = FALSE))
```

> `r nrow(ind_discrete$chisq_dat)` patients with a high cholesterol classification were randomly assigned to either a drug or lifestyle intervention, `r ind_discrete$chisq_dat %>% filter(intervention == "Drug") %>% nrow()` in each intervention. The test of two proportions used was the chi-square test of homogeneity. At the conclusion of the drug intervention, `r ind_discrete$chisq_dat %>% filter(intervention == "Drug" & risk_level == "Normal") %>% nrow()` patients (`r tabyl(ind_discrete$chisq_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Drug") %>% pull(Normal) %>% scales::percent()`) had improved their cholesterol classification from high to normal compared to `r ind_discrete$chisq_dat %>% filter(intervention == "Lifestyle" & risk_level == "Normal") %>% nrow()` patients (`r tabyl(ind_discrete$chisq_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Lifestyle") %>% pull(Normal) %>% scales::percent()`) in the lifestyle intervention, a difference in proportions of `r (tabyl(ind_discrete$chisq_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Drug") %>% pull(Normal) - tabyl(ind_discrete$chisq_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Lifestyle") %>% pull(Normal)) %>% scales::number(accuracy = .01)`, *p* = `r ind_discrete$chisq_test$p.value %>% scales::number(accuracy = .0001)`.

##### Fisher {-}

```{r}
(ind_discrete$fisher_test <- ind_discrete$fisher_dat %>%
  tabyl(intervention, risk_level) %>%
  fisher.test())
```

> `r nrow(ind_discrete$fisher_dat)` patients with a high cholesterol classification were randomly assigned to either a drug or lifestyle intervention, `r ind_discrete$fisher_dat %>% filter(intervention == "Drug") %>% nrow()` in each intervention. At the conclusion of the drug intervention, `r ind_discrete$fisher_dat %>% filter(intervention == "Drug" & risk_level == "Normal") %>% nrow()` patients (`r tabyl(ind_discrete$fisher_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Drug") %>% pull(Normal) %>% scales::percent()`) had improved their cholesterol classification from high to normal compared to `r ind_discrete$fisher_dat %>% filter(intervention == "Lifestyle" & risk_level == "Normal") %>% nrow()` patients (`r tabyl(ind_discrete$fisher_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Lifestyle") %>% pull(Normal) %>% scales::percent()`) in the lifestyle intervention. Due to small sample sizes, Fisher's exact test was run. There was a non-statistically significant difference in proportions of `r (tabyl(ind_discrete$fisher_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Drug") %>% pull(Normal) - tabyl(ind_discrete$fisher_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Lifestyle") %>% pull(Normal)) %>% scales::number(accuracy = .01)`, *p* = `r ind_discrete$fisher_test$p.value %>% scales::number(accuracy = .0001)`.

## Pairwise Prop Test {#pairwiseproptest}
```{r}
library(tidyverse)
M <- 3573
F <- 4177
dat <- tribble(
  ~gender, ~src, ~Y, ~N,
  "Male", "Indeed", 1699, M-1699,
  "Male", "LinkedIn", 1755, M-1755,
  "Male", "Google", 1578, M-1578,
  "Female", "Indeed", 2554, F-2554,
  "Female", "LinkedIn", 1914, F-1914,
  "Female", "Google", 1694, F-1694
)
prop.test(x = dat$Y, n = dat$Y + dat$N)
pairwise.prop.test(x = dat$Y, n = dat$Y + dat$N)
```

## McNemar's Test {#mcnemar}

This test applies when you have paired samples.

Wilcoxon Paired-Sample applies when the variable distributions are non-normally distributed and samples are paired.

