# Two-Group Differences

```{r include=FALSE}
library(tidyverse)
library(gtsummary)
library(foreign)
library(scales)
library(janitor)
library(flextable)
```

Use *independent samples* tests to either *describe* a variable's frequency or central tendency difference between two independent groups, or to *compare* the difference to a hypothesized value.

If the data generating process produces continuous outcomes (interval or ratio) and the outcomes are symmetrically distributed, the difference in the sample means, $\hat{d} = \bar{x} - \bar{y}$, is a random variable centered at the population difference, $d = \mu_X - \mu_Y$. You can use a theoretical distribution (normal or student t) to estimate a 95% confidence interval (CI) around $d$, or compare $\hat{d}$ to an hypothesized population difference, $d_0$. If you (somehow) know the sampling distribution variances $\sigma^2_X$ and $\sigma^2_Y$, or the Central Limit Theorem (CLT) conditions hold, you can assume the random variable is normally distributed and use the *z*-test, otherwise assume the random variable has a student t distribution and use the *t*-test.^[The *t*-test returns nearly the same result as the *z*-test when the CLT holds, so in practice no one bothers with the *z*-test except as an aid to teach the *t*-test.] If the data generating process produces continuous outcomes that are *not* symmetrically distributed, use a non-parametric test like the Mann-Whitney U test.

If the data generating process produces discrete outcomes (counts), the sample count, $x$, is a random variable from a Poisson, binomial, normal, or multinomial distribution, or a random variable from a theoretical outcome. For two independent samples, the data can be organized into a two-way table - a frequency table for two categorical variables. If you have a single categorical predictor variable, you can test whether the joint frequency counts differ from the expected frequency counts in the saturated model. You analyze a two-way table one of two ways.

* If you only care about comparing two levels (like when the response variable is binary), conduct a proportion difference *z*-test or a Fisher exact-test.

* If you want to compare the joint frequency counts to *expected* frequency counts under the *independence model* (the model of independent explanatory variables), conduct a Pearson's chi-squared independence test, or a *G*-test.

## Independent Samples (Continuous)

### Independent Samples t-Test {-}

If a population measure *X* is normally distributed with mean $\mu_X$ and variance $\sigma_X^2$, and a population measure *Y* is normally distributed with mean $\mu_Y$ and variance $\sigma_Y^2$, then their difference is normally distributed with mean $d = \mu_X - \mu_Y$ and variance $\sigma_{XY}^2 = \sigma_X^2 + \sigma_Y^2$. By the CLT, as the sample sizes grow, a non-normally distributed *X* and *Y* will approach normality, and so will their difference.

The **independent samples t-test** evaluates an hypothesized difference, $d_0$ (H0: $d = d_0$), from the difference in sample means $\hat{d} = \bar{x} - \bar{y}$, or constructs a (1 - $\alpha$)% confidence interval around $\hat{d}$ to estimate $d$ within a margin of error, $\epsilon$.

In principal, you can evaluate $\hat{d}$ with either a *z*-test or a *t*-test. Both require independent samples and approximately normal sampling distributions. Sampling distributions are normal if the underlying populations are normally distributed, or if the sample sizes are large ($n_X$ and $n_Y$ $\ge$ 30). However, the *z*-test additionally requires known sampling distribution variances, $\sigma^2_X$ and $\sigma^2_Y$. These variances are never known, so always use the *t*-test.

The *z*-test assumes $d$ is normally distributed around $\hat{d} = d$ with standard error $SE = \sqrt{\frac{\sigma_X^2}{n_X} + \frac{\sigma_Y^2}{n_Y}}.$ The test statistic for H0: $d = d_0$ is $Z = \frac{\hat{d} - d_0}{SE}$. The (1 - $\alpha$)% CI is $d = \hat{d} \pm z_{(1 - \alpha {/} 2)} SE$.  

The *t*-test assumes $d$ has a *t*-distribution around $\hat{d} = d$ with standard error $SE = \sqrt{\frac{s_X^2}{n_X} + \frac{s_Y^2}{n_Y}}.$ The test statistic for H0: $d = d_0$ is $T = \frac{\hat{d} - d_0}{SE}$. The (1 - $\alpha$)% CI iss $d = \hat{d} \pm t_{(1 - \alpha / 2), (n_X + n_Y - 2)} SE$.

There is a complication with the *t*-test *SE* and degrees of freedom. If the sample sizes are small and the standard deviations from each population are similar (the ratios of $s_X$ and $s_Y$ are <2), pool the variances, $s_p^2 = \frac{(n_X - 1) s_X^2 + (n_Y-1) s_Y^2}{n_X + n_Y-2}$, so that $SE = s_p \sqrt{\frac{1}{n_X} + \frac{1}{n_Y}}$ and the degrees of freedom (df) = $n_X + n_Y - 2$ (the **pooled variances t-test**). Otherwise, $SE = \sqrt{\frac{s_X^2}{n_X} + \frac{s_Y^2}{n_Y}}$, but you reduce df using the Welch-Satterthwaite correction, $df = \frac{\left(\frac{s_X^2}{n_X} + \frac{s_Y^2}{n_Y}\right)^2}{\frac{s_X^4}{n_X^2\left(N_X-1\right)} + \frac{s_Y^4}{n_Y^2\left(N_Y-1\right)}}$ (the **separate variance t-test**, or **Welch's t-test**).

### Wilcoxon Rank Sum Test {-}

The **Mann-Whitney U test**^[This test is sometimes called the Mann-Whitney test, Mann-Whitney U test, Wilcoxon-Mann-Whitney test, and the two-sample Wilcoxon test], is a nonparametric alternative to the independent-samples *t*-test. Use the the test when the samples are not normally distributed or when the response variables are ordinal rather continuous. In the first case where the normality assumption fails, the test evaluates H0 that the two samples are from the same population distribution. In the second case where the response variables are ordinal, the test evaluates the difference in medians. 

The Wilcoxon Rank Sum test ranks the response values, then sums the ranks for the reference group, $W = \sum R_1$. The test statistic is $U = W - \frac{n_2(n_2 + 1)}{2}$ where $n_2$ is the number of observations in the test group. $U$ will equal 0 if there is complete separation between the groups, and $n_1 n_2$ if there is complete overlap. Reject H0 if $U$ is sufficiently small.

### Case Study {-}

```{r include=FALSE}
ind_num <- list()

ind_num$t_dat <- read.spss("./input/independent-samples-t-test.sav", to.data.frame = TRUE)
ind_num$t_n <- with(ind_num$t_dat, by(engagement, gender, length))
ind_num$t_mean <- with(ind_num$t_dat, by(engagement, gender, mean))
ind_num$t_sd <- with(ind_num$t_dat, by(engagement, gender, sd))

ind_num$mw_dat <- read.spss("./input/mann-whitney-test.sav", to.data.frame = TRUE)
ind_num$mw_n <- with(ind_num$mw_dat, by(engagement, gender, length))
ind_num$mw_median <- with(ind_num$mw_dat, by(engagement, gender, median))
ind_num$mw_sd <- with(ind_num$mw_dat, by(engagement, gender, sd))
```

A company shows an advertisement to $n_M$ = `r ind_num$t_n["Male"]` males and $n_F$ = `r ind_num$t_n["Female"]` females, then measures their engagement with a survey. Do the groups' mean engagement scores differ?

[Laerd](https://statistics.laerd.com/) has two data sets for this example. One meets the conditions for a *t*-test, and the other fails the normality test, forcing you to use the Mann-Whitney U test.

```{r fig.height=3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `t-Test` = ind_num$t_dat, 
  `Mann-Whitney` = ind_num$mw_dat, 
  .id = "set"
) %>%
  ggplot(aes(x = engagement, fill = gender)) +
  # geom_density(alpha = 0.6) +
  geom_histogram(bins = 20, alpha = 0.6, position = "dodge") +
  facet_wrap(~fct_rev(set)) +
  labs(title = "Mean Engagement Scores Counts", fill = NULL) +
  theme_light() +
  scale_fill_manual(values = c("Male" = "slategray", "Female" = "snow3"))
```

```{r fig.height=3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `t-test` = ind_num$t_dat,
  `Mann-Whitney` = ind_num$mw_dat,
  .id = "set"
) %>%
  group_by(set, gender) %>%
  summarize(.groups = "drop",
            mean_engage = mean(engagement),
            n = n(),
            se_engage = sd(engagement) / sqrt(n),
            ci_low = mean_engage - qt(.975, n-1) * se_engage,
            ci_high = mean_engage + qt(.975, n-1) * se_engage) %>%
  ggplot(aes(x = gender)) +
  geom_col(aes(y = mean_engage), fill = "snow3", color = "snow4", width = 0.25) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), color = "snow4", width = .125) +
  theme_minimal() +
  facet_wrap(~fct_rev(set)) +
  labs(x = NULL, y = NULL,
       title = "Mean Engagement Score", 
       subtitle = "Two data sets with similar means.",
       caption = "Error bars show 95% CI.")
```

The *t*-test data set has the following summary statistics.

```{r}
(ind_num$t_gt <- ind_num$t_dat %>% 
  gtsummary::tbl_summary(
    by = c(gender), 
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ))
```

> There were `r ind_num$t_n["Male"]` male and `r ind_num$t_n["Female"]` female participants. Data are mean $\pm$ standard deviation, unless otherwise stated. The advertisement was more engaging to male viewers, `r gtsummary::inline_text(ind_num$t_gt, engagement, column = "Male")`, than female viewers, `r gtsummary::inline_text(ind_num$t_gt, engagement, column = "Female")`.

The Mann-Whitney data set has the following summary statistics.

```{r}
(ind_num$mw_gt <- ind_num$mw_dat %>% 
  gtsummary::tbl_summary(
    by = c(gender), 
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ))
```

> There were `r ind_num$mw_n["Male"]` male and `r ind_num$mw_n["Female"]` female participants. Data are mean $\pm$ standard deviation, unless otherwise stated. The advertisement was more engaging to male viewers, `r gtsummary::inline_text(ind_num$mw_gt, engagement, column = "Male")`, than female viewers, `r gtsummary::inline_text(ind_num$mw_gt, engagement, column = "Female")`.

#### Conditions {-}

The independent samples *t*-test and Mann-Whitney U test apply when 1) the response variable is continuous, 2) the independent variable is binomial, and 3) the observations are independent. The decision between the *t*-test and Mann-Whitney stems from two additional conditions related to the data distribution - if both conditions hold, use the *t*-test; otherwise use Mann-Whitney.

1. **Outliers**. There should be no outliers in either group. Outliers exert a large influence on the mean and standard deviation. Test with a box plot. If there are outliers, you might be able to drop them or transform the data.
2. **Normality**.  Values should be *nearly* normally distributed. The *t*-test is robust to normality, but this condition is important with small sample sizes. Test with Q-Q plots or the Shapiro-Wilk test for normality. If the data is very non-normal, you might be able to transform the data.

If the data passes the two conditions, use the *t*-test, but now you need to check a third condition related to the variances to determine which flavor of the *t*-test to use.

3. **Homogeneous Variances**. Use *pooled-variances* if the variances are homogeneous; otherwise use the *separate variances* method. Test with Levene's test of equality of variances.

If the data does not pass the first two conditions, use Mann-Whitney, but now you need to check a third condition here as well. The condition does not affect how to perform the test, but rather how to interpret the results.

3. **Distribution shape**. If the distributions have the same shape, interpret the Mann-Whitney result as a comparison of the *medians*; otherwise interpret the result as a comparison of the *mean ranks*.

##### Checking for Outliers {-}

Assess outliers with a box plot. Box plot whiskers extend up to 1.5\*IQR from the upper and lower hinges and outliers (beyond the whiskers) are are plotted individually.

```{r echo=FALSE, fig.height=2.5, fig.width=7.5}
bind_rows(
  `t-test` = ind_num$t_dat,
  `Mann-Whitney` = ind_num$mw_dat,
  .id = "set"
) %>%
  ggplot(aes(x = gender, y = engagement)) +
  geom_boxplot(fill = "snow3", color = "snow4", alpha = 0.6, width = 0.25, 
               outlier.color = "goldenrod", outlier.size = 2) +
  theme_minimal() +
  facet_wrap(~fct_rev(set)) +
  labs(title = "Boxplot of Engagement Score", y = "Score", x = NULL)
```

For the *t* test data set,

>There were no outliers in the data, as assessed by inspection of a boxplot.

and for the Mann-Whitney data set,

>There was one outlier in the data, as assessed by inspection of a boxplot.

If the outliers are data entry errors or measurement errors, fix or discard them. If the outliers are genuine, you have a couple options before reverting to the Mann-Whitney U test.

* Leave it in if it doesn't affect the conclusion (compared to taking it out).
* Transform the variable. Don't do this unless the variable is also non-normal. Transformation also has the downside of making interpretation more difficult.

##### Checking for Normality {-}

Assume the population is normally distributed if *n* $\ge$ 30. Otherwise, assess a Q-Q plot, skewness and kurtosis values, or a histogram. If you still don't feel confident about normality, run a Shapiro-Wilk test.

There are only $n_M$ = `r ind_num$t_n["Male"]` male and $n_F$ = `r ind_num$t_n["Female"]` female observations, so you need to test normality. The QQ plot indicates normality in the t-test data set, but not in the Mann-Whitney data set.

```{r fig.height=3.5, fig.width=7.5}
bind_rows(
  `t-test` = ind_num$t_dat,
  `Mann-Whitney` = ind_num$mw_dat,
  .id = "set"
) %>%
  ggplot(aes(sample = engagement, group = gender, color = fct_rev(gender))) +
  stat_qq() +
  stat_qq_line(col = "goldenrod") +
  theme_minimal() + theme(legend.position = "top") +
  facet_wrap(~fct_rev(set)) +
  labs(title = "Normal Q-Q Plot", color = NULL)
```

Run Shapiro-Wilk separately for the males and for the females. Since we are looking at two data sets in tandem, there are four tests below. For the t-test data set, 

```{r}
(ind_num$t_shapiro <- split(ind_num$t_dat, ind_num$t_dat$gender) %>% 
  map(~shapiro.test(.$engagement))
)
```

>Engagement scores for each level of gender were normally distributed, as assessed by Shapiro-Wilk's test (*p* > .05).

For the Mann-Whitney data set,

```{r}
(ind_num$mw_shapiro <- split(ind_num$mw_dat, ind_num$mw_dat$gender) %>% 
  map(~shapiro.test(.$engagement))
)
```

>Engagement scores for each level of gender were not normally distributed for the Female sample, as assessed by Shapiro-Wilk's test (*p* = `r ind_num$mw_shapiro$Female$p.value %>% scales::comma(accuracy = 0.001)`).

If the data is not normally distributed, you still have a couple options before reverting to the Mann-Whitney U test.

* Transform the dependent variable.
* Carry on regardless - the independent samples *t*-test is fairly robust to deviations from normality.

##### Checking for Homogenous Variances {-}

If the data passed the outliers and normality tests, you will use the *t*-test, so now you need to test the variances to see which version (*pooled-variances* method if variances are homogeneous; *separate variances* if variances are heterogeneous). A rule of thumb is that homogeneous variances have a ratio of standard deviations between 0.5 and 2.0:

```{r}
sd(ind_num$t_dat %>% filter(gender == "Male") %>% pull(engagement)) /
  sd(ind_num$t_dat %>% filter(gender == "Female") %>% pull(engagement))
```

You can also use the *F* test to compare the ratio of the sample variances $\hat{r} = s_X^2 / s_Y^2$ to an hypothesized ratio of population variances $r_0 = \sigma_X^2 / \sigma_Y^2 = 1.$

```{r}
var.test(ind_num$t_dat %>% filter(gender == "Female") %>% pull(engagement), 
         ind_num$t_dat %>% filter(gender == "Male") %>% pull(engagement))
```

Bartlett's test is another option.

```{r}
bartlett.test(ind_num$t_dat$engagement, ind_num$t_dat$gender)
```

Levene's test is a third option. Levene's is less sensitive to departures from normality than Bartlett.

```{r}
(ind_num$levene <- with(ind_num$t_dat, 
                        car::leveneTest(engagement, gender, center = "mean"))
)
```

> There was homogeneity of variances for engagement scores for males and females, as assessed by Levene's test for equality of variances (*p* = `r ind_num$levene %>% pluck("Pr(>F)", 1) %>% scales::number(accuracy = 0.001)`).

##### Checking for Similar Distributions {-}

If the data fail either the outliers or the normality test, use the Mann-Whitney test. The Mann-Whitney data set failed both, so the Mann-Whitney test applies. Now you need to test the distributions to determine how to interpret its results. If the distributions are similarly shaped, interpret the Mann-Whitney U test as inferences about differences in medians between the two groups. If the distributions are dissimilar, interpret the test as inferences about the distributions, lower/higher scores and/or mean ranks.

```{r fig.height = 3.5, fig.width=7.5, echo=FALSE}
ind_num$mw_dat %>%
  ggplot(aes(x = engagement, color = fct_rev(gender))) +
  geom_density() +
  theme_minimal() +
  theme(legend.position = "right") +
  labs(title = "Engagement Distribution", color = NULL)
```

> Distributions of the engagement scores for males and females were similar, as assessed by visual inspection.

#### Test {-}

Conduct the *t*-test or the Mann-Whitney U test. 

##### t-Test {-}

The the *t*-test data the variances were equal, so the pooled-variances version applies (`t.test(var.equal = TRUE)`).

```{r}
(ind_num$t_test <- t.test(engagement ~ gender, data = ind_num$t_dat, var.equal = TRUE))
```

> There was a statistically significant difference in mean engagement score between males and females, with males scoring higher than females, `r scales::number(ind_num$t_test$estimate[1] - ind_num$t_test$estimate[2] %>% as.numeric(), accuracy = 0.01)` (95% CI, `r ind_num$t_test$conf.int[1] %>% scales::number(accuracy = 0.01)` to `r ind_num$t_test$conf.int[2] %>% scales::number(accuracy = 0.01)`), t(`r ind_num$t_test$parameter`) = `r ind_num$t_test$statistic %>% scales::number(accuracy = 0.001)`, *p* = `r ind_num$t_test$p.value %>% scales::number(accuracy = 0.001)`.

The effect size, Cohen's *d*, is defined as $d = |M_D| / s$, where $|M_D| = \bar{x} - \bar{y}$, and $s$ is the pooled sample standard deviation, $s_p = \sqrt{\frac{(n_X - 1) s_X^2 + (n_Y-1) s_Y^2}{n_X + n_Y-2}}$. $d <.2$ is considered trivial, $.2 \le d < .5$ small, and $.5 \le d < .8$ large.

```{r message=FALSE}
(d <- effectsize::cohens_d(engagement ~ gender, data = ind_num$t_dat, pooled_sd = TRUE))
```

> There was a large difference in mean engagement score between males and females, Cohen's d = `r scales::number(d$Cohens_d, accuracy = 0.01)` 95% CI [`r scales::number(d$CI_low, accuracy = 0.01)`, `r scales::number(d$CI_high, accuracy = 0.01)`]

Before rejecting the null hypothesis, construct a plot as a sanity check.

```{r echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=7.5}
d_bar <- ind_num$t_mean["Male"] - ind_num$t_mean["Female"]
d_0 <- 0
se <- effectsize::sd_pooled(engagement ~ gender, data = ind_num$t_dat) * 
  sqrt(1/ind_num$t_n["Male"] + 1/ind_num$t_n["Female"])
df <- sum(ind_num$t_n) - 2
lrr <- se * qt(.025, df)
urr <- se * qt(.975, df)
data.frame(d = seq(-1.0, 1.0, by = .01)) %>%
  mutate(t = (d - d_0) / se,
         prob = dt(x = t, df = df),
         lrr = if_else(d < lrr, prob, NA_real_),
         urr = if_else(d > urr, prob, NA_real_)) %>%
  ggplot() +
  geom_line(aes(x = d, y = prob)) +
  geom_area(aes(x = d, y = lrr), fill = "firebrick", alpha = 0.6) +
  geom_area(aes(x = d, y = urr), fill = "firebrick", alpha = 0.6) +
  geom_vline(aes(xintercept = d_0), size = 1) +
  geom_vline(aes(xintercept = d_bar), color = "goldenrod", size = 1, linetype = 2) +
  annotate("text", x = d_bar + .06, y = .05, 
           label = scales::number(d_bar, accuracy = .01), 
           size = 3, color = "goldenrod3") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.y = element_blank()) +
  labs(title = bquote("Two-Tailed t-test"),
       x = "Difference in Group Means (Male - Female)",
       y = "Probability",
       caption = "Shaded area is 95% CI.")
```

##### Wilcoxon Rank Sum test {-}

The reference level for the `gender` variable is males, so the Wilcoxon Rank Sum test statistic is the sum of male ranks minus $n_f(n_f + 1) / 2$ where $n_f$ is the number of females. You can calculate the test statistic by hand.

```{r}
(ind_num$mw_test_manual <- ind_num$mw_dat %>% 
  mutate(R = rank(engagement)) %>%
  group_by(gender) %>%
  summarize(.groups = "drop", n = n(), R = sum(R), meanR = sum(R)/n()) %>%
  pivot_wider(names_from = gender, values_from = c(n, R, meanR)) %>%
  mutate(U = R_Male - n_Female * (n_Female + 1) / 2))
```

Compare the test statistic to the Wilcoxon rank sum distribution with `pwilcox()`.

```{r}
pwilcox(
  q = ind_num$mw_test_manual[1, ]$U - 1, 
  m = ind_num$mw_test_manual[1, ]$n_Male, 
  n = ind_num$mw_test_manual[1, ]$n_Male, 
  lower.tail = FALSE
) * 2
```

There is a function for all this.

```{r}
(ind_num$mw_test <- wilcox.test(
  engagement ~ gender, 
  data = ind_num$mw_dat, 
  exact = TRUE, 
  correct = FALSE,
  conf.int = TRUE))
```

> Median engagement score was not statistically significantly different between males and females, *U* = `r ind_num$mw_test$statistic`, *p* = `r ind_num$mw_test$p.value %>% scales::number(accuracy = .001)`, using an exact sampling distribution for *U*.

Now you are ready to report the results. Here is how you would report the *t* test.

> Data are mean $\pm$ standard deviation, unless otherwise stated. There were `r ind_num$t_n[["Male"]]` male and `r ind_num$t_n[["Female"]]` female participants. An independent-samples t-test was run to determine if there were differences in engagement to an advertisement between males and females. There were no outliers in the data, as assessed by inspection of a boxplot. Engagement scores for each level of gender were normally distributed, as assessed by Shapiro-Wilk's test (*p* > .05), and there was homogeneity of variances, as assessed by Levene's test for equality of variances (*p* = `r ind_num$levene %>% pluck("Pr(>F)", 1) %>% scales::number(accuracy = 0.001)`). The advertisement was more engaging to male viewers (`r scales::number(ind_num$t_mean["Male"], accuracy = .01)` $\pm$ = `r scales::number(ind_num$t_sd["Male"], accuracy = .01)`) than female viewers (`r scales::number(ind_num$t_mean["Female"], accuracy = .01)` $\pm$ = `r scales::number(ind_num$t_sd["Female"], accuracy = .01)`), a statistically significant difference of `r scales::number(ind_num$t_test$estimate[1] - ind_num$t_test$estimate[2] %>% as.numeric(), accuracy = 0.01)` (95% CI, `r ind_num$t_test$conf.int[1] %>% scales::number(accuracy = 0.01)` to `r ind_num$t_test$conf.int[2] %>% scales::number(accuracy = 0.01)`), t(`r ind_num$t_test$parameter`) = `r ind_num$t_test$statistic %>% scales::number(accuracy = 0.001)`, *p* = `r ind_num$t_test$p.value %>% scales::number(accuracy = 0.001)`, *d* = `r d$Cohens_d %>% scales::number(accuracy = 0.01)`.

Here is how you would report the Mann-Whitney U-Test.

> A Mann-Whitney U test was run to determine if there were differences in engagement score between males and females. Distributions of the engagement scores for males and females were similar, as assessed by visual inspection. Median engagement score for males (`r ind_num$mw_median["Male"] %>% as.numeric() %>% scales::number(accuracy = .01)`) and females (`r ind_num$mw_median["Female"] %>% as.numeric() %>% scales::number(accuracy = .01)`) was not statistically significantly different, *U* = `r ind_num$mw_test$statistic`, *p* = `r ind_num$mw_test$p.value %>% scales::number(accuracy = .001)`, using an exact sampling distribution for *U*.

Had the distributions differed, you would report the Mann-Whitney like this:

> A Mann-Whitney U test was run to determine if there were differences in engagement score between males and females. Distributions of the engagement scores for males and females were not similar, as assessed by visual inspection. Engagement scores for males (mean rank = `r ind_num$mw_test_manual %>% pull(meanR_Male)`) and females (mean rank = `r ind_num$mw_test_manual %>% pull(meanR_Female)`) were not statistically significantly different, *U* = `r ind_num$mw_test$statistic`, *p* = `r ind_num$mw_test$p.value %>% scales::number(accuracy = .001)`, using an exact sampling distribution for *U*.

## Paired Samples (Continuous)

There are two common study designs that employ a paired samples t-test to compare two related groups. One relates the groups as two time points for the same subjects. The second relates the groups as two tests of the same subjects, e.g. comparing reaction time under two lighting conditions.

### Paired Samples t-Test {-}

The *paired samples t-test* uses the mean of sampled paired differences $\bar{d}$ as an estimate of the mean of the population paired differences $\delta$ to evaluate an hypothesized mean $\delta_0$.  Test $H_0: \delta = \delta_0$ with test statistic $T = \frac{\bar{d} - \delta_0}{se}$, or define a $(1 - \alpha)\%$ confidence interval as $\delta = \bar{d} \pm t_{1 - \alpha / 2, n - 1} se$.  *The paired t-test is really just a one-sample mean t-test operating on variable that is defined as the difference between two variables*.

The paired samples *t* test applies when the sampling distribution of the mean of the population paired differences is normally distributed and there are no significant outliers.

### Wilcoxon Signed-Rank Test {-}

The *Wilcoxon signed-rank test* is a nonparametric alternative to the paired-samples t-test for cases in which the paired differences fails the normality condition, but is at least symmetrically distributed.

The test statistic is the sum product of the difference signs (-1, +1) and the rank of the difference absolute values, $W = \sum_{i=1}^n sign (d_i) \cdot R_i$. The more differences that are of one sign, or of extreme magnitude, the larger $W$ is likely to be, and the more likely to reject $H_0$ of equality of medians.

### Sign Test {-}

The *sign test* is an alternative to the Wilcoxon signed-rank test for cases in which the paired differences fails the symmetrical distribution condition.

The test statistic is the count of pairs whose difference is positive, $W = cnt(d_i > 0)$. $W \sim b(n, 0.5)$, so the sign test is really just an exact binomial test (*exact sign test*), or for large *n*-size, the normal approximation to the binomial (*sign test*).

### Case Study {-}


```{r include=FALSE}
drink <- list()

drink$t_dat <- read.spss("./input/paired-samples-t-test.sav", to.data.frame = TRUE) %>%
  mutate(diff = carb_protein - carb)
drink$t_n <- nrow(drink$t_dat)
drink$t_mean <- mean(drink$t_dat$carb_protein - drink$t_dat$carb)
drink$t_sd <- sd(drink$t_dat$carb_protein - drink$t_dat$carb)

drink$wilcoxon_dat <- read.spss("./input/wilcoxon-signed-rank-test.sav", 
                                to.data.frame = TRUE) %>%
  mutate(diff = carb_protein - carb)
drink$wilcoxon_n <- nrow(drink$wilcoxon_dat)
drink$wilcoxon_median <- list()
drink$wilcoxon_med$carb_protein <- median(drink$wilcoxon_dat$carb_protein)
drink$wilcoxon_med$carb <- median(drink$wilcoxon_dat$carb) 

drink$sign_dat <- read.spss("./input/sign-test.sav", to.data.frame = TRUE) %>%
  mutate(diff = carb_protein - carb)
drink$sign_n <- nrow(drink$sign_dat)
drink$sign_median <- list()
drink$sign_med$carb_protein <- median(drink$sign_dat$carb_protein)
drink$sign_med$carb <- median(drink$sign_dat$carb) 
```

$n$ = `r drink$t_n` athletes consume a carb-only or carb+protein drink prior to running as far as possible in 2 hours and a researcher records their distances under each condition. Do the distances differ from 0?

[Laerd](https://statistics.laerd.com/) has three data sets for this example. One meets the conditions for a t-test. The second fails the normality condition, but is symmetric and meets the conditions for the Wilcoxon test. The third fails the symmetry condition and requires the sign test.

##### t-test data set {-}

```{r}
(drink$t_gt <- drink$t_dat %>% 
  gtsummary::tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})"))
)
```

> There were `r drink$t_n` participants. Data are mean $\pm$ standard deviation, unless otherwise stated. Participants ran further after consuming the carbohydrate-protein drink, `r gtsummary::inline_text(drink$t_gt, carb_protein)` km, than the carbohydrate-only drink, `r gtsummary::inline_text(drink$t_gt, carb)` km.

##### Wilcoxon data set {-}

Once you learn you need Wilcoxon or the sign-test, show the median and IQR summary statistics instead.

```{r}
(drink$wilcoxon_gt <- drink$wilcoxon_dat %>% 
  gtsummary::tbl_summary()
)
```

> There were `r drink$wilcoxon_n` participants. Data are medians and IQR unless otherwise stated. Participants ran further after consuming the carbohydrate-protein drink, `r gtsummary::inline_text(drink$wilcoxon_gt, carb_protein)` km, than the carbohydrate-only drink, `r gtsummary::inline_text(drink$wilcoxon_gt, carb)` km.

##### Sign data set {-}

```{r}
(drink$sign_gt <- drink$sign_dat %>% 
  gtsummary::tbl_summary()
)
```

> There were `r drink$sign_n` participants. Data are median and IQR unless otherwise stated. Participants ran further after consuming the carbohydrate-protein drink, `r gtsummary::inline_text(drink$sign_gt, carb_protein)` km, than the carbohydrate-only drink, `r gtsummary::inline_text(drink$sign_gt, carb)` km.

#### Conditions {-}

The paired samples *t* test applies when the variable is continuous and partitioned into dependent pairs, Additionally, there are two conditions related to the data distribution. If either condition fails, consider the suggested work-around or move to the non-parametric alternatives.

1. **Outliers**. There should be no outliers in the differences because they exert a large influence on the mean and standard deviation. Test with a box plot. If there are outliers, you might be able to drop them if they do not affect the conclusion, or you can transform the data.
2. **Normality**. Differences should be *nearly* normally distributed ("nearly" because the *t*-test is robust to the normality assumption). This condition is especially important with small sample sizes. Test with Q-Q plots or the Shapiro-Wilk test for normality. If the data is very non-normal, you might be able to transform the data.

##### Outliers {-}

Assess outliers with a box plot. Box plot whiskers extend up to 1.5\*IQR from the upper and lower hinges and outliers (beyond the whiskers) are are plotted individually.

```{r echo=FALSE, fig.height=2.5, fig.width=7.5}
bind_rows(
  `t-test` = drink$t_dat,
  `Wilcoxon` = drink$wilcoxon_dat,
  `Sign` = drink$sign_dat,
  .id = "set"
) %>%
  ggplot(aes(x = "difference", y = diff)) +
  geom_boxplot(fill = "snow3", color = "snow4", alpha = 0.6, width = 0.25, 
               outlier.color = "goldenrod", outlier.size = 2) +
  theme_minimal() +
  facet_wrap(~fct_inorder(set)) +
  labs(title = "Boxplot of Difference", y = "Km", x = NULL)
```

>There were no outliers in the data, as assessed by inspection of a boxplot.

Had there been outliers, you might report

> *X* outliers were detected. Inspection of their values did not reveal them to be extreme and they were kept in the analysis.

If the outliers are data entry errors or measurement errors, fix them or discard them. If the outliers are genuine, you can try leaving them in or transforming the data.

##### Normality {-}

Assume the population is normally distributed if *n* $\ge$ 30. These data sets have *n* = `r drink$t_n` observations, so you cannot assume normality. Asses a Q-Q plot, skewness and kurtosis values, histogram, or Shapiro-Wilk test.

```{r fig.height=3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `t-test` = drink$t_dat,
  `Wilcoxon` = drink$wilcoxon_dat,
  `Sign` = drink$sign_dat,
  .id = "set"
) %>%
  ggplot(aes(sample = diff)) +
  stat_qq() +
  stat_qq_line(col = "goldenrod") +
  theme_minimal() + theme(legend.position = "top") +
  facet_wrap(~fct_inorder(set)) +
  labs(title = "Normal Q-Q Plot", color = NULL)
```

For the t-test data set, 

```{r}
(drink$t_shapiro <- shapiro.test(drink$t_dat$diff))
```

> The differences between the distance ran in the carbohydrate-only and carbohydrate-protein trial were normally distributed, as assessed by Shapiro-Wilk's test (*p* = `r drink$t_shapiro$p.value %>% scales::number(accuracy = 0.001)`).

For the Wilcoxon data set,

```{r}
(drink$wilcoxon_shapiro <- shapiro.test(drink$wilcoxon_dat$diff))
```

> The differences between the distance ran in the carbohydrate-only and carbohydrate-protein trial were *not* normally distributed, as assessed by Shapiro-Wilk's test (*p* = `r drink$wilcoxon_shapiro$p.value %>% scales::number(accuracy = 0.001)`).

For the sign-test data set,

```{r}
(drink$sign_shapiro <- shapiro.test(drink$sign_dat$diff))
```

> The differences between the distance ran in the carbohydrate-only and carbohydrate-protein trial were *not* normally distributed, as assessed by Shapiro-Wilk's test (*p* = `r drink$sign_shapiro$p.value %>% scales::number(accuracy = 0.001)`).

If the data is normally distributed, use the t-test. If not, you try transforming the dependent variable, or carrying on regardless since the *t*-test is fairly robust to deviations from normality.

##### Symmetric Distribution {-}

If the data passed the outliers test, but failed the normality test, as the Wilcoxon and sign test data sets above did, you will use the Wilcoxon signed-rank test or sign test. Now you need to test the distribution to determine which test. If the distribution is symmetric, use Wilcoxon; otherwise use the sign test.

```{r fig.height = 3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `Wilcoxon` = drink$wilcoxon_dat,
  `Sign` = drink$sign_dat,
  .id = "set"
) %>%
  mutate(diff_sign = case_when(diff < 0 ~ "-", diff > 0 ~ "+", TRUE ~ "0")) %>%
  ggplot(aes(x = diff, fill = fct_rev(diff_sign))) +
  geom_histogram(bins = 20, color = "snow4") +
  scale_fill_manual(values = list("-" = "firebrick", "+" = "steelblue", "0" = "grey")) +
  theme_bw() +
  theme(legend.position = "top") +
  labs(title = "Paired Samples Differences Distribution", fill = NULL, x = NULL, y = "n") +
  facet_wrap(~fct_rev(set))
```

For the Wilcoxon data set,

> The distribution of the differences between the carbohydrate-protein drink and the carbohydrate-only was symmetric, as assessed by visual inspection.

For the sign data set,

> The distribution of the differences between the carbohydrate-protein drink and the carbohydrate-only was *not* asymmetric, as assessed by visual inspection.

#### Test {-}

##### t-test {-}

```{r}
(drink$t_t <- t.test(x = drink$t_dat$carb_protein, y = drink$t_dat$carb, paired = TRUE)
)
```

> The carbohydrate-protein drink elicited an increase of `r scales::number(drink$t_t$estimate %>% as.numeric(), accuracy = 0.001)` (95% CI, `r drink$t_t$conf.int[1] %>% scales::number(accuracy = 0.001)` to `r drink$t_t$conf.int[2] %>% scales::number(accuracy = 0.001)`) km in the distance run in two hours compared to a carbohydrate-only drink.

The effect size, called Cohen's *d*, is the number of standard deviations the measured mean difference is from the hypothesized difference, $(\bar{d}-d_0) / s$, where $s$ is the sample standard deviation. .2 is small, .5 is medium, and .8 is large. This one is large.

```{r}
(drink$t_d <- effectsize::cohens_d(drink$t_dat$diff))
```

You are about to reject the null hypothesis. Construct a plot as a sanity check on your reasoning.

```{r echo=FALSE, warning=FALSE, fig.height=2.5, fig.width=6.5}
d_0 <- 0
d_bar <- drink$t_mean
se <- drink$t_sd / sqrt(drink$t_n)
ci_lwr <- d_0 - qt(.975, df = drink$t_n - 1) * se
ci_upr <- d_0 + qt(.975, df = drink$t_n - 1) * se
ci_range <- ci_upr - ci_lwr

data.frame(d = seq(ci_lwr - ci_range/2, ci_upr + ci_range/2, length = 100)) %>%
  mutate(t = (d - d_0) / se,
         prob = dt(x = t, df = drink$t_n - 1),
         lrr = if_else(d < ci_lwr, prob, NA_real_),
         urr = if_else(d > ci_upr, prob, NA_real_)) %>%
  ggplot(aes(x = d)) +
  geom_line(aes(y = prob)) +
  geom_area(aes(y = lrr), fill = "firebrick", alpha = 0.6) +
  geom_area(aes(y = urr), fill = "firebrick", alpha = 0.6) +
  geom_vline(aes(xintercept = d_0), size = 1) +
  geom_vline(aes(xintercept = d_bar), color = "goldenrod", size = 1, linetype = 2) +
  annotate("label", x = d_0, y = 0.25,  label = "d[0]", parse = TRUE,
           fill = "grey90", size = 3) +
  annotate("label", x = d_bar, y = 0.15, label = "~bar(d)", parse = TRUE,
           fill = "goldenrod", size = 3) +
  theme_minimal() +
  theme(legend.position="none",         axis.text.y = element_blank()) +
  labs(title = "Two-Tailed t-Test", x = expression(paste(delta)), y = "Probability")
```

Report the results.

> A paired-samples t-test was used to determine the effect of a new formula of sports drink on running performance. Instead of the regular, carbohydrate-only drink, the new sports drink contains a new carbohydrate-protein mixture. Twenty participants were recruited to the study who each performed two trials in which they had to run as far as possible in two hours on a treadmill. In one of the trials they drank the carbohydrate-only drink and in the other trial they drank the carbohydrate-protein drink. The order of the trials was counterbalanced and the distance they ran in both trials was recorded.
>
>Two outliers were detected that were more than 1.5 box-lengths from the edge of the box in a boxplot. Inspection of their values did not reveal them to be extreme and they were kept in the analysis. The assumption of normality was not violated, as assessed by Shapiro-Wilk's test (*p* = `r drink$t_shapiro$p.value %>% scales::number(accuracy = 0.001)`). 
>
>Data are mean $\pm$ standard deviation, unless otherwise stated. Participants ran further after consuming the carbohydrate-protein drink, `r gtsummary::inline_text(drink$t_gt, carb_protein)` km, than the carbohydrate-only drink, `r gtsummary::inline_text(drink$t_gt, carb)` km, a statistically significant increase of `r scales::number(drink$t_t$estimate %>% as.numeric(), accuracy = 0.001)` (95% CI, `r drink$t_t$conf.int[1] %>% scales::number(accuracy = 0.001)` to `r drink$t_t$conf.int[2] %>% scales::number(accuracy = 0.001)`) km, *t*(`r drink$t_t$parameter`) = `r drink$t_t$statistic %>% scales::number(accuracy = 0.001)`, *p* = `r drink$t_t$p.value %>% scales::number(accuracy = .0001)`, *d* = `r drink$t_d$Cohens_d %>% scales::number(accuracy = .01)`.

##### Wilcoxon Signed-Rank Test {-}

From the distribution plot, you can see that most of the signs were positive, and the largest absolute difference values were among the positives, so expect a pretty large test statistic.

```{r}
(drink$wilcoxon_test <- wilcox.test(drink$wilcoxon_dat$carb_protein,
                                    drink$wilcoxon_dat$carb, 
                                    paired = TRUE))
```
> The carbohydrate-protein drink elicited a statistically significant median increase in distance run in two hours compared to the carbohydrate-only drink, *W* = `r drink$wilcoxon_test$statistic %>% number(accuracy = 1)`, *p* = `r drink$wilcoxon_test$p.value %>% scales::number(accuracy = .001)`.

Report the results.

> A Wilcoxon signed-rank test was conducted to determine the effect of a new formula of sports drink on running performance. Instead of the regular, carbohydrate-only drink, the new sports drink contains a new carbohydrate-protein mixture. Twenty participants were recruited to the study who each performed two trials in which they had to run as far as possible in two hours on a treadmill. In one of the trials they drank the carbohydrate-only drink and in the other trial they drank the carbohydrate-protein drink. The order of the trials was counterbalanced and the distance they ran in both trials was recorded.
>
>The difference scores were approximately symmetrically distributed, as assessed by a histogram with superimposed normal curve.
>
>Data are medians unless otherwise stated. Of the 20 participants recruited to the study, the carbohydrate-protein drink elicited an increase in the distance run in 17 participants compared to the carbohydrate-only drink, whereas two participants saw no improvement and one participant did not run as far with the carbohydrate-protein drink. There was a statistically significant median increase in distance run (`r median(drink$sign_dat$diff) %>% scales::number(accuracy = .0001)` km) when subjects imbibed the carbohydrate-protein drink (`r median(drink$sign_dat$carb_protein) %>% scales::number(accuracy = .001)` km) compared to the carbohydrate-only drink (`r median(drink$sign_dat$carb) %>% scales::number(accuracy = .001)` km), *W* = `r drink$wilcoxon_test$statistic %>% number(accuracy = 1)`, *p* = `r drink$wilcoxon_test$p.value %>% scales::number(accuracy = .0001)`.

##### Sign Signed-Rank Test {-}

Conduct the exact sign test since the n-size is not so large that we need the normal approximation to the binomial. Notice `n` is the count of *non-zero* differences.

```{r}
(drink$sign_test <- binom.test(sum(drink$sign_dat$diff > 0), 
                               n = sum(drink$sign_dat$diff != 0)))
```

> The carbohydrate-protein drink elicited a statistically significant median increase in distance run (`r median(drink$sign_dat$diff) %>% scales::number(accuracy = .001)` km) compared to the carbohydrate-only drink, *p* = `r drink$sign_test$p.value %>% scales::number(accuracy = .001)`.

Report the results.

> An exact sign test was conducted to determine the effect of a new formula of sports drink on running performance. Instead of the regular, carbohydrate-only drink, the new sports drink contains a new carbohydrate-protein mixture. Twenty participants were recruited to the study who each performed two trials in which they had to run as far as possible in two hours on a treadmill. In one of the trials they drank the carbohydrate-only drink and in the other trial they drank the carbohydrate-protein drink. The order of the trials was counterbalanced and the distance they ran in both trials was recorded.
>
> An exact sign test was used to determine whether there was a statistically significant median difference between the distance ran when participants drank a carbohydrate-protein drink compared to a carbohydrate-only drink. Data are medians unless otherwise stated. Of the `r drink$sign_n` participants recruited to the study, the carbohydrate-protein drink elicited an increase in the distance run in `r sum(drink$sign_dat$diff > 0)` participants compared to the carbohydrate-only drink, whereas `r sum(drink$sign_dat$diff < 0)` participants did not run as far and `r sum(drink$sign_dat$diff == 0)` participant saw no improvement with the carbohydrate-protein drink. There was a statistically significant median increase in distance run (`r median(drink$sign_dat$diff) %>% scales::number(accuracy = .0001)` km) when subjects imbibed the carbohydrate-protein drink (`r median(drink$sign_dat$carb_protein) %>% scales::number(accuracy = .001)` km) compared to the carbohydrate-only drink (`r median(drink$sign_dat$carb) %>% scales::number(accuracy = .001)` km), *p* = `r drink$sign_test$p.value %>% scales::number(accuracy = .0001)`.

## Independent Samples (Discrete)

There are three common tests: the z-test of two proportions; chi-square test of homogeneity; and Fisher's exact test. The z-test and chi-square test produce the same statistical significance result because they are algebraically identical, however the chi-square test is more commonly used. Both tests require large *n*>30 sample sizes. When this assumption is violated, use Fisher's exact.

### z-Test of Two Proportions {-}

The *z*-test uses the difference in sample proportions $\hat{d} = p_1 - p_2$ as an estimate of the difference in population proportions $\delta = \pi_1 - \pi_2$ to evaluate an hypothesized difference in population proportions $d_0 = \pi_0 - \pi_1$ and/or construct a $(1−\alpha)\%$ confidence interval around $\hat{d}$ to estimate $\delta$ within a margin of error $\epsilon$.

The *z*-test applies when the central limit theorem conditions hold so that the normal distribution approximates the binomial distribution.

* the sample is independently drawn, meaning random assignment (experiments) or random sampling without replacement from $n < 10\%$ of the population (observational studies),
* there are at least $n_i p_i >= 5$ successes and $n_i (1 - p_i) >= 5$ failures for each group $i$,
* the sample sizes are both $n_i >= 30$, and 
* the probability of success for each group is not extreme, $0.2 < \pi_i < 0.8$.

If these conditions hold, the sampling distribution of $\delta$ is normally distributed around $\hat{d}$ with standard error $se_\hat{d} = \sqrt{\frac{p_1(1 - p_1)}{n_1} + \frac{p_2(1 − p_2)}{n_2}}$. The measured values $\hat{d}$ and $se_\hat{d}$ approximate the population values $\delta$ and $se_\delta$. Define a $(1 − \alpha)\%$ confidence interval as $\hat{d} \pm z_{\alpha / 2}se_\hat{d}$ or test the hypothesis of $d = d_0$ with test statistic $z = \frac{\hat{d} − d_0}{se_{d_0}}$ where $se_{d_0} = \sqrt{p^*(1 - p^*) \left(\frac{1}{n_1} + \frac{1}{n_2}\right)}$ and $p^*$ is the overall success probability.

### Chi-Square Test of Homogeneity {-}

The chi-square test of homogeneity tests whether frequency counts of the *R* levels of a categorical variable are distributed identically across the *C* populations. It tests whether observed joint frequency counts $O_{ij}$ differ from expected frequency counts $E_{ij}$ under the *independence model* (the model of independent explanatory variables, $\pi_{ij} = \pi_{i+} \pi_{+j}$. $H_0$ is $O_{ij} = E_{ij}$. The chi-square homogeneity test can be extended to cases where $I$ and/or $J$ is greater than 2.

There are two possible test statistics for this test, Pearson $X^2 = \sum \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$, and deviance $G^2 = 2 \sum_{ij} O_{ij} \log \left( \frac{O_{ij}}{E_{ij}} \right)$.

### Fisher's Exact Test {-}

Fisher's exact test is an "exact test" in that the p-value is calculated exactly from the hypergeometric distribution rather than relying on the approximation that the test statistic distribution approaches $\chi^2$ as $n \rightarrow \infty$.

The test is applicable in situations where

* the row totals $n_{i+}$ and the column totals $n_+j$ are fixed by study design (rarely applies), and
* the expected values of >20% of cells (at least 1 cell in a 2x2 table) have expected cell counts >5, and no expected cell count is <1.

The p-value from the test is computed as if the margins of the table are fixed. This leads under a null hypothesis of independence to a hypergeometric distribution of the numbers in the cells of the table ([Wikipedia](https://en.wikipedia.org/wiki/Fisher%27s_exact_test)). Fisher's exact test is useful for small *n*-size samples where the chi-squared distribution assumption of the chi-squared and G-test tests fails. Fisher's exact test is overly conservative (p values too high) for large *n*-sizes.

The Hypergeometric density function is
$$f_X(k|N, K, n) = \frac{{{K}\choose{k}}{{N-K}\choose{n-k}}}{{N}\choose{n}}.$$

The density is the exact hypergeometric probability of observing this particular arrangement of the data, assuming the given marginal totals, on the null hypothesis that the conditional probabilities are equal.

### Case Study {-}

```{r include=FALSE}
ind_discrete <- list()

ind_discrete$chisq_dat <- read.spss(
  "./input/test-of-two-proportions-individual-scores.sav",
  to.data.frame = TRUE
)

ind_discrete$fisher_dat <- ind_discrete$chisq_dat[seq(2, 100, 2),]
```

The case study below uses a data set from [Laerd](https://statistics.laerd.com/) and a second modified version. The first data set passes the chi-square test of homogeneity requirements. The second (in parentheses), fails the n-sizes test. 

A researcher recruits `r nrow(ind_discrete$chisq_dat)` (`r nrow(ind_discrete$fisher_dat)`) patients who have a "high" classification of cholesterol and who currently have a poor lifestyle. The researcher randomly assigns `r ind_discrete$chisq_dat %>% filter(intervention == "Drug") %>% nrow()` (`r ind_discrete$fisher_dat %>% filter(intervention == "Drug") %>% nrow()`) of them to a drug intervention and `r ind_discrete$chisq_dat %>% filter(intervention == "Lifestyle") %>% nrow()` (`r ind_discrete$fisher_dat %>% filter(intervention == "Lifestyle") %>% nrow()`) to a lifestyle intervention. After six months, a doctor reclassifies the patients as either still having a "high" classification of cholesterol or now having a "normal" classification of cholesterol.

```{r fig.height=3.5, fig.width=7.5, echo=FALSE}
bind_rows(
  `Chi-sq` = ind_discrete$chisq_dat,
  `Fisher` = ind_discrete$fisher_dat,
  .id = "set"
) %>%
  count(set, intervention, risk_level) %>%
  group_by(set, intervention) %>%
  summarize(.groups = "drop",
            p = sum(if_else(risk_level == "Normal", n, as.integer(0))) / sum(n),
            n = sum(n),
            se = sqrt(p * (1 - p) / n),
            ci_lwr = p - qt(.975, n - 1) * se,
            ci_upr = p + qt(.975, n - 1) * se) %>%
  ggplot(aes(x = intervention)) +
  geom_col(aes(y = p), fill = "snow3", color = "snow4", width = 0.25) +
  geom_errorbar(aes(ymin = ci_lwr, ymax = ci_upr), color = "snow4", width = .125) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~set) +
  labs(x = NULL, y = NULL,
       title = "Cholesterol Classification Improvement from High to Normal",
       subtitle = "Two data sets with different n-sizes.",
       caption = "Error bars show 95% CI.")
```

The chi-sq data set has the following summary statistics.

```{r echo=FALSE}
x <- ind_discrete$chisq_dat %>%
  tabyl(intervention, risk_level) %>%
  adorn_totals(where = c("row", "col")) %>%
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")
  
ind_discrete$chisq_test <- ind_discrete$chisq_dat %>%
  tabyl(intervention, risk_level) %>%
  chisq.test()
y <- ind_discrete$chisq_test$expected %>%
  adorn_totals(where = c("row", "col")) %>%
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")

bind_rows(list(Observed = x, Expected = y), .id = "Data") %>%
  flextable() %>% 
  set_caption("Chisq Data Set: Observed vs Expected") %>% 
  autofit() %>%
  flextable::border(i = 3, border.bottom = officer::fp_border()) %>%
  flextable::merge_v(j = 1)
```

The Fisher data set has the following summary statistics.

```{r echo=FALSE}
x <- ind_discrete$fisher_dat %>%
  tabyl(intervention, risk_level) %>%
  adorn_totals(where = c("row", "col")) %>%
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")
  
ind_discrete$fisher_chisq_test <- ind_discrete$fisher_dat %>%
  tabyl(intervention, risk_level) %>%
  chisq.test()
y <- ind_discrete$fisher_chisq_test$expected %>%
  adorn_totals(where = c("row", "col")) %>%
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")

bind_rows(list(Observed = x, Expected = y), .id = "Data") %>%
  flextable() %>% 
  set_caption("Fisher Data Set: Observed vs Expected") %>% 
  autofit() %>%
  flextable::border(i = 3, border.bottom = officer::fp_border()) %>%
  flextable::merge_v(j = 1)
```

#### Conditions {-}

##### n-Size {-}

The chi-square test of homogeneity applies with the CLT conditions hold.

* the sample is independently drawn,
* there are at least 5 successes (Normal) and failures (High) for each group $i$,
* the sample sizes for both groups are >=30, and 
* the probability of success for each group is not extreme, $0.2 < \pi_i < 0.8$.

The conditions hold for the chi-sq data set, but not for the Fisher data set.

#### Test {-}

##### Chi-Square {-}

```{r}
(ind_discrete$chisq_test <- ind_discrete$chisq_dat %>%
  tabyl(intervention, risk_level) %>%
  chisq.test(correct = FALSE))
```

> `r nrow(ind_discrete$chisq_dat)` patients with a high cholesterol classification were randomly assigned to either a drug or lifestyle intervention, `r ind_discrete$chisq_dat %>% filter(intervention == "Drug") %>% nrow()` in each intervention. The test of two proportions used was the chi-square test of homogeneity. At the conclusion of the drug intervention, `r ind_discrete$chisq_dat %>% filter(intervention == "Drug" & risk_level == "Normal") %>% nrow()` patients (`r tabyl(ind_discrete$chisq_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Drug") %>% pull(Normal) %>% scales::percent()`) had improved their cholesterol classification from high to normal compared to `r ind_discrete$chisq_dat %>% filter(intervention == "Lifestyle" & risk_level == "Normal") %>% nrow()` patients (`r tabyl(ind_discrete$chisq_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Lifestyle") %>% pull(Normal) %>% scales::percent()`) in the lifestyle intervention, a difference in proportions of `r (tabyl(ind_discrete$chisq_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Drug") %>% pull(Normal) - tabyl(ind_discrete$chisq_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Lifestyle") %>% pull(Normal)) %>% scales::number(accuracy = .01)`, *p* = `r ind_discrete$chisq_test$p.value %>% scales::number(accuracy = .0001)`.

##### Fisher {-}

```{r}
(ind_discrete$fisher_test <- ind_discrete$fisher_dat %>%
  tabyl(intervention, risk_level) %>%
  fisher.test())
```

> `r nrow(ind_discrete$fisher_dat)` patients with a high cholesterol classification were randomly assigned to either a drug or lifestyle intervention, `r ind_discrete$fisher_dat %>% filter(intervention == "Drug") %>% nrow()` in each intervention. At the conclusion of the drug intervention, `r ind_discrete$fisher_dat %>% filter(intervention == "Drug" & risk_level == "Normal") %>% nrow()` patients (`r tabyl(ind_discrete$fisher_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Drug") %>% pull(Normal) %>% scales::percent()`) had improved their cholesterol classification from high to normal compared to `r ind_discrete$fisher_dat %>% filter(intervention == "Lifestyle" & risk_level == "Normal") %>% nrow()` patients (`r tabyl(ind_discrete$fisher_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Lifestyle") %>% pull(Normal) %>% scales::percent()`) in the lifestyle intervention. Due to small sample sizes, Fisher's exact test was run. There was a non-statistically significant difference in proportions of `r (tabyl(ind_discrete$fisher_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Drug") %>% pull(Normal) - tabyl(ind_discrete$fisher_dat, intervention, risk_level) %>% adorn_percentages() %>% filter(intervention == "Lifestyle") %>% pull(Normal)) %>% scales::number(accuracy = .01)`, *p* = `r ind_discrete$fisher_test$p.value %>% scales::number(accuracy = .0001)`.

## Paired Samples (Discrete)
### McNemar's Test {-}

This test applies when you have paired samples.

Wilcoxon Paired-Sample applies when the variable distributions are non-normally distributed and samples are paired.